### 2017-03-30 paper.tex
Initial commit

\input{header.tex}

\begin{document}

\begin{titlepage}
\maketitle
\end{titlepage}

\begin{abstract}

\end{abstract}

\newpage
\tableofcontents

\newpage
\chapter{Introduction}
\label{sec:introduction}

\newpage
\chapter{Literature Review}
\label{sec:literature-review}

Most HAB missions led by universities, schools or amateur are built around
commercially available small computers or micro-controller boards: Redland
Green School students used a Raspberry Pi (\cite{rpi2014}) running
python scripts as an On Board Computer (\textit{OBC}) for their HAB mission
(\cite{Hinschelwood2015}), while other projects have been ran on AVR
micro-controller based boards like the Arduino Mega
(\cite{AtmelCorporation2015}).

In the United Kingdom, radio licenses do not cover airborne transmission, making
packet radio protocols like AX.25 unusable for HAB missions 
(\cite{ukhasradio2016}). However, the Office of Communications allows the use
of certain \textit{license-exempt} frequencies like the 70cm wavelength band,
provided that the transmission power does not exceed 10mW (\cite{Ofcom2014}).
For this reason, most missions launched from the United Kingdom use the
Radiometrix NTX2 transmitter (\cite{radiometrix2012}). Because of the
constraints, these missions usually transmit their telemetry as text using the
Radio-Teletype (RTTY) protocol.

While most HAB projects make use of commercial-off-the-shelf components, the
finished payloads are usually custom, single-purpose designs that are little,
if at all modular. In the design document for the Titan-1 HAB mission, Bombasaro
describes a hardware bus based on tightly coupled sensors which communicate
over two different data buses, SPI and I2C, and a custom-made, single-purpose
flight software (\cite{Bombasaro2015}).

Because of the risks and cost involved, nano-satellites rely more on modular
designs, allowing each team to work on their module independently. In his paper,
Volstad describes the design of the data bus of the NTNU CubeSat: while the OBC
is a based on a custom made circuit board, it is designed to provide a standard
power interface, as well as access to an I2C standard bus
(\cite{NXPSemiconductors2014}) to each payload (\cite{Volstad2011}). The I2C
protocol was chosen because of its low power consumption, and because if only
requires two lines (clock and data). Thomas Clausen describes in his 2001 paper
how a simple packet protocol built on top of I2C itself is used for data
transfer and error detection in Aalborg University's CubeSats
(\cite{Clausen2001}).

Satellite missions last a lot longer than High-Altitude Balloons' and the
requirements for radio transmissions differ: while satellites can only
communicate when their orbit passes above a ground station and require precise
speed and data volume planning, as described by Sandy Anthunes in his book
(\cite{Antunes2015}), non-floating HAB missions allow for line-of-sight
communications from liftoff to late into the descent of the payload.

There are however some technologies that can be adapted from satellites to be
used in HAB applications. The University of Arizona uses a custom packet format
over 434MHz radio, containing raw binary values for each instrument's
measurements (\cite{Eatchel2002}). The BRITE-Austria CubeSat mission uses the
AX.25 packet radio protocol (used by amateur radio users), which allows the
use of off-the-shelf transmission and reception hardware rather than custom-made
circuits (\cite{Traussnig2007}).

Some nano-satellites use custom telemetry data format, which allows the team
to minimise the volume of data to be sent over radio. The Planetary Society's
Lightsail mission has an uplink connection that allows the ground station
to request specific logs or data. By default, The spacecraft only communicates
a custom beacon containing a summary of the its housekeeping data
(\cite{planetary2016}). Uplink being impractical with the transmission power
limits imposed in the United Kingdom, such a selective telemetry system cannot
be relied upon for HAB missions.

To improve collaboration and allow the deployment of large networks of
satellites, probes and other space vehicles, the Consultative Committee for
Space Data Systems has defined multiple communication standards that for
the different types of networks encountered in spacecrafts.

SpaceWire defines how instruments' data can be accessed by having the OBC
remotely poll their memory (\cite{parkes2005}). A similar system is used to
fetch data from ROM chips over I2C and could be used to build a simple data bus.

The CCSDS Space Packet protocol defines a packet format used
to encapsulate application data (from instruments for example) that can be sent
over a ``Space Link'', an analogue of the OSI model's link layer
(\cite{Stallings1987}) using CCSDS frames to encapsulate packets originating
from multiple instruments, ground stations and spacecrafts (\cite{ccsds2003}).

While HAB applications do not require the level of complexity of CCSDS
protocols — addressing is not needed since the network only contains two
endpoints, communicating in a single direction — some details of the Space
Packet protocol are worth reusing (Forward Error Correction, different
Application IDs for each payload).

\section{Analysis of the Problem}

\newpage
\chapter{Proposed Solution and Methodology}
\label{sec:methodology}

The proposed solution for this project is inspired by the bus model that
supports the architecture of most commercial satellites. A standard software
and hardware platform provides common facilities required on all High-Altitude
flights – position tracking, internal data transfers and radio telemetry to the
ground station – allowing payloads (scientific instruments for example) to
only generate data and make it available to the platform.

\begin{figure}[H]
\includegraphics[width=0.35\textwidth]{payloadbus.pdf}
\centering
\caption{Payload bus architecture}
\end{figure}

The platform, or \textit{Payload Bus} is composed of the following components hardware and software components:

\begin{description}

\item[On Board Computer:] the computer that runs the flight software and
controls the data bus and all communications units (also called
\textit{Bus Controller}).

\item[Flight Software:] the firmware installed on the \textit{OBC}, that
controls the bus and the execution of the mission.

\item[Power Bus:] the hardware interface that provides electrical power to
payloads.

\item[Data Bus:] the hardware interface and software protocol through which
payloads communicate data to the flight computer.

\item[Navigation System:] the hardware (GNSS receiver) and software module used
by the flight software to keep track of the balloon's geographic coordinates and
altitude.

\item[Radio System:] the hardware (transmitter), protocols and software module
used by the \textit{OBC} to transmit collected data and metadata to the
ground station.

\item[Ground Segment:] the software required to record and decode the signal
received from the platform.

\end{description}

The following chapter describes the design processes used to develop and
integrate those components into a functioning prototype of the Abertay
High-Altitude Bus (\textit{AHABus}), and the methods used to test the viability
and performance of the design.

\section{Radio Communications Protocol}

This section describes the design process that led to the radio communications
protocol used by the AHABus platform to send payload data and flight metadata
to the ground station. Its role is to ensure that arbitrary binary data – each
payload can use their own data format – as stably and reliably as possible. As
discussed in the following subsections, the design and implementation choice
were also driven by legal constraints: in the United Kingdom, airborne radio
transmitters are limited in frequency band (70cm wavelength) and power
(transmission power inferior to 10mW).

\subsection{Requirements}

Designing and building a communication layer that can be relied up was an
important requirement for the rest of the project. Transmitting as much as
possible of a mission's scientific data is primordial in case the platform
is carried too far by winds and cannot be recovered after landing.

From the constraints discussed above, the requirements below were established
for the radio communications system:

\begin{enumerate}
    \item provide a system to transmit binary data.
    \item minimise the amount of overhead and bandwidth required.
    \item allow multiple payload's data to be sent on the same link.
    \item provide stable transmissions on a low-power link.
    \item correct as many transmission-induced data errors as possible.
    \item allow ground crew to track the balloon's location over time.
\end{enumerate}

%TODO: mention one-way

As stated in chapter \ref{sec:literature-review}, the United Kingdom's Office
of Communications restrict what frequencies can be used for airborne
radio transmissions, and how much power the transmissions can be sent at.
Those restrictions constrain what existing protocols can be used to transmit
arbitrary data over radio. This disqualifies existing packet radio protocols
that would have fulfilled most of the requirements, in use in other countries
like AX.25.

Most UK-launched High Altitude Balloon missions make use of Radio-Teletype
(\textit{RTTY}) to send ASCII-encoded text. This was considered insufficient
for the purpose of a common platform for two main reasons: First, ASCII does
not handle binary data by default -- only some byte values would be displayed
as ASCII, while the rest would translate to non-printing characters. This leads
to data loss, unless the binary data is encoded in an ASCII-compatible format
like base64, which would require more bandwidth to send a similar amount of
data. Furthermore, ASCII alone does not provide either a way to multiplex data
coming from different payloads, or any kind of error checking or error
correction. This means that such facilities must be built on top of the
plain-text protocol.

RTTY itself functions in the same way as wired UART serial links. It can provide
some level of error checking if the transmission of a parity bit is enabled, but
it does not carry enough information to correct errors. Additionally, RTTY is
purely a link layer, and does not provide any kind of encapsulation necessary
for multiplexing data from different payloads.

\subsection{Protocol Specification}

The AHABus radio protocol was designed to match the requirements for the radio
system as closely as possible. While the specification presented in this section
is the last version, which was implemented in the AHABus prototype, some parts
of the protocol evolved or had to be changed during the prototyping process.
% TODO: explain more? less?

The protocol was designed in layers, some equivalent to the ones present in the
OSI model (\cite{Stallings1987}). This approach was chosen because it allows
some level of modularity – if a layer proves problematic, it can be swapped
for another with minimal impact on the other layers. Each layer fulfils a
specific purpose:

\begin{description}
\item[Link Layer:] The lowest-level layer's purpose is to establish a one-way
binary stream between the two end points of the radio link.

\item[Frame Layer:] The frame layer relies on the Link Layer to transmit its
data units, and provides reliability and error correction to higher-level 
layers.

\item[Packet Layer:] The packet layer is the highest-level layer. It provides
a way for multiple payloads' data to share bandwidth as well as metadata
transmission.
\end{description}

\subsubsection{Link Layer}

\subsubsection{Forward Error Correction}

\subsubsection{Frame Layer}

\begin{center}
\begin{bytefield}{16}
    \bitheader{0,7,8,15} \\
    \begin{rightwordgroup}{Header}
        \bitbox{8}{01011010} & \bitbox{8}{version} \\
        \bitbox{16}{sequence number}
    \end{rightwordgroup} \\
    \begin{leftwordgroup}{Payload}
        \wordbox[lrt]{1}{data} \\
        \skippedwords \\
        \wordbox[lrb]{1}{}
    \end{leftwordgroup} \\
    \begin{rightwordgroup}{Error Correction}
        \wordbox[lrt]{1}{checksum} \\
        \skippedwords \\
        \wordbox[lrb]{1}{}
    \end{rightwordgroup}
\end{bytefield}
\end{center}

\subsubsection{Packet Layer}

\begin{center}
\begin{bytefield}{32}
    \bitheader{0,7,8,15,16,23,24,31} \\
    \begin{rightwordgroup}{Header}
        \bitbox{8}{version} & \bitbox{8}{payload ID} &
        \bitbox{16}{packet length} \\
        \bitbox{32}{latitude} \\
        \bitbox{32}{longitude} \\
        \bitbox{16}{altitude} & \bitbox[lrt]{16}{}
    \end{rightwordgroup} \\
    \begin{leftwordgroup}{Payload}
        \wordbox[lrt]{1}{data} \\
        \skippedwords \\
        \wordbox[lrb]{1}{}
    \end{leftwordgroup}
\end{bytefield}
\end{center}

\subsection{Initial Implementation and Testing}

\section{Bus and Flight Software}

\subsection{Requirements}

\subsection{Protocol Specification}

\subsection{Arduino Implementation}

\section{Flight Software}

\section{Ground Segment}

\subsection{Reception Binary Stream Extraction}

\section{Testing}

\subsection{Component Testing}

\subsection{Integrated Testing}

\subsubsection{Day in the life Testing}

\subsubsection{Resilience Testing}

\newpage
\chapter{Results}
\label{sec:reults}

\newpage
\chapter{Discussion}
\label{sec:discussion}

\newpage
\chapter{Conclusion}
\label{sec:conclusion}

\newpage
\printbibliography[heading=bibnumbered]

\end{document}

### 2017-03-30 paper.tex
Add link layer info

@@
 \chapter{Introduction}
-\label{sec:introduction}
+\label{ch:introduction}
 
@@
 \chapter{Literature Review}
-\label{sec:literature-review}
+\label{ch:literature-review}
 
@@
 \chapter{Proposed Solution and Methodology}
-\label{sec:methodology}
+\label{ch:methodology}
 
@@
 \subsection{Requirements}
+\label{ssec:requirements}
 
@@
 
-As stated in chapter \ref{sec:literature-review}, the United Kingdom's Office
+As stated in chapter \ref{ch:literature-review}, the United Kingdom's Office
 of Communications restrict what frequencies can be used for airborne
@@
 
-\subsubsection{Forward Error Correction}
+While RTTY was not considered sufficient to cover fulfil the requirements of
+the whole protocol, as shown in section \ref{ssec:requirements}, it fulfils the
+simple requirements of the Link Layer: it can be used to send 8-bit bytes, one
+by one, over a radio link.
+
+RTTY functions like UART, over radio. Instead of using logical electrical levels
+(5V for logical high, 0V for logical low), it uses Audio Frequency Shift Keying
+(\textit{AFSK}): two audio tones separated by a frequency shift act as two
+logical levels. When nothing is being transmitted, the line is held high. The
+start of a byte is signalled by the \textit{start bit} (the line goes low),
+followed by the eight \textit{data bits}, and one or two \textit{stop bits} (the
+line is held low). After a byte has been sent, the line is held high again until
+the next byte, or the next byte starts immediately.
+
+%TODO: replace mentions of UART with RS-232 // NOPE, RS232 diff. than TTL
+%TODO: references for DominoEx and Olivia
+Other existing protocols exist that can transmit data over radio audio signals,
+like DominoEx or Olivia, sometimes with better transmission rates. RTTY was
+preferred over them because the ESP8266 board used as OBC in the AHABus
+prototype provides hardware support for UART, making it easier to implement and
+reliable. Since the link layer is transparent to the frame and packet layers,
+RTTY could be swapped for another protocol in the future with no impact on
+those.
 
@@
 
-\begin{center}
+\begin{figure}[H]
     \begin{bytefield}{16}
@@
         \end{leftwordgroup} \\
-    \begin{rightwordgroup}{Error Correction}
+        \begin{rightwordgroup}{FEC}
             \wordbox[lrt]{1}{checksum} \\
@@
     \end{bytefield}
-\end{center}
+    \centering
+    \caption{AHABus Frame Format}
+\end{figure}
 
@@
 
-\begin{center}
+\begin{figure}[H]
     \begin{bytefield}{32}
@@
 \chapter{Results}
-\label{sec:reults}
+\label{ch:reults}
 
@@
 \chapter{Discussion}
-\label{sec:discussion}
+\label{ch:discussion}
 
@@
 \chapter{Conclusion}
-\label{sec:conclusion}
+\label{ch:conclusion}
 

### 2017-04-01 paper.tex
Keep adding to frame format

@@
 
-The platform, or \textit{Payload Bus} is composed of the following components hardware and software components:
+The platform, or \textit{Payload Bus} is composed of the following components 
+hardware and software components:
 
@@
 
-%TODO: mention one-way
-
 As stated in chapter \ref{ch:literature-review}, the United Kingdom's Office
@@
 that would have fulfilled most of the requirements, in use in other countries
-like AX.25.
+like AX.25. \todo{mention Tx only, no Rx}
 
@@
 of the protocol evolved or had to be changed during the prototyping process.
-% TODO: explain more? less?
+\todo{Might not be necessary to say that here, given further sections}
 
@@
 
+\todo{Add some details about simplicity of the network topology}
+
 \begin{description}
@@
 data units, and provides reliability and error correction to higher-level 
-layers.
+layers.\todo{Add vocabulary maybe (ancillary)}
 
@@
 
+The design process and specification of each of those three layers is explained
+in more details in the following sections.
+
 \subsubsection{Link Layer}
+\label{sssec:link-layer}
+
+\todo{add an introductory sentence?}
 
@@
 followed by the eight \textit{data bits}, and one or two \textit{stop bits} (the
-line is held low). After a byte has been sent, the line is held high again until
-the next byte, or the next byte starts immediately.
+line is held high). After a byte has been sent, the line is held high again
+until the next byte, or the next byte starts immediately.
 
-%TODO: replace mentions of UART with RS-232 // NOPE, RS232 diff. than TTL
-%TODO: references for DominoEx and Olivia
 Other existing protocols exist that can transmit data over radio audio signals,
+\todo{references for DominoEx and Olivia}
 like DominoEx or Olivia, sometimes with better transmission rates. RTTY was
@@
 
+The valid encoding and decoding of a RTTY signal depends on the chosen baud
+rate (the number of times the logical signal changes level). During tests,
+the highest baud rate that allowed clear decoding of radio signals at 10mW
+power rating was 200bauds. Since each byte requires eleven bits to be sent
+(one start bit, eight data bits and two stop bits), this corresponds to a
+\(\frac{200}{11}=18.18\) bytes per second bandwidth.
+
 \subsubsection{Frame Layer}
+
+The role of the frame layer is to provide higher-level layers with a reliable
+data stream: the Link Layer discussed in section \ref{sssec:link-layer} will
+send data bytes one after another, but it does not guarantee that the bytes are
+received, or that noise on the channel (in this case, radio waves) have not
+damaged the data between transmission and reception. Those are the two main
+potential types of issues that the Frame Layer was designed to address.
+
+Preventing complete data loss on a radio channel is only possible to a degree:
+the signal-to-noise ratio is limited by multiple factors including transmission
+power – limited in the United Kingdom – distance, other neighbouring
+transmissions and line-of-sight. Those issues could have been mitigated in part
+through solutions that were outside of this project's scope, like antennae
+design. While this meant that some data loss could not be prevented, a solution
+was required to detect occurrences of data loss, and identify which part of the
+data was missing.
+
+Identifying lost sections of an arbitrary stream of bytes without any form of
+formatting is not possible: since it is not known in advance what data should
+be expected, there is no model to which actual received data can be compared.
+For this reason, the AHABus radio protocol was designed as a framed protocol:
+the stream of data is divided in chunks of a set number of bytes, encapsulated
+in \textit{frames} that contain the a sequence number used to track the position
+of the chunk in the data stream.
+
+The AHABus frame design was inspired by other frame-based protocols like the
+CCSDS Space Packet/Frame protocol. Each frame is composed of two section: a
+header, that contains metadata, and a payload containing the data carried by
+that frame.
+
+The only piece of metadata strictly required to be carried in a frame was its
+sequence number. The frame format was designed to use a 16-bit integer to
+represent the sequence number, as it allowed a large number of frames to be
+sent in-between wrap-arounds of the number (when it becomes larger than 65,535).
+
+To ensure that frames can be identified in the future, even if the format were
+to evolve, the protocol version number was also added to the header format. At
+the time of this writing, the protocol version was 0. If a future frame decoder
+were to receive a frame indicating an old protocol version, it could be at
+least signalled to the user, if not handled gracefully.
+
+\todo{explain requirement for sync marker}
+
+The resulting first version of the frame format is presented in figure
+\ref{fig:frame-fmt-orig}.
+
+\begin{figure}[H]
+    \begin{bytefield}{24}
+        \bitheader{0,7,8,15,16,23} \\
+        \bitbox{8}{version} & \bitbox{16}{sequence number} \\
+            \wordbox[lrt]{1}{payload data} \\
+            \skippedwords \\
+            \wordbox[lrb]{1}{}
+    \end{bytefield}
+    \centering
+    \caption{Original AHABus Frame Format}
+    \label{fig:frame-fmt-orig}
+\end{figure}
+
+\todo{explain move away from double-marker towards sequence + mark}
+
+Encapsulating the transmitted data adds overhead. If \(n\) is the size of a
+chunk in bytes, and each frame adds \(m\) bytes of metadata, then only a
+percentage \(\frac{n}{n+m}\) of the stream contains actual data. For reasons
+explained further, the AHABus frames were chosen to be 256-byte long. As shown
+previously, each frame also includes.
 
@@
         \end{rightwordgroup} \\
-        \begin{leftwordgroup}{Payload}
+        \begin{rightwordgroup}{Payload}
             \wordbox[lrt]{1}{data} \\
@@
             \wordbox[lrb]{1}{}
-        \end{leftwordgroup} \\
+        \end{rightwordgroup} \\
         \begin{rightwordgroup}{FEC}
@@
 
+% Topic list for this:
+%   * Provide reliability:
+%       * Data arrives whole
+%       * We receive valid data
+%   * receive all data: we chunk the data stream, number the chunks         v
+%   * we tag data with housekeeping things                                  v~
+%   * If frames too small, housekeeping/data ratio is shit
+%   * Need a way to detect the start + end of a frame.
+%   * Explain that end of frame marker complicated, going with count?
+%
+%   * Validity checked (and corrected using FEC)
+%   * FEC: went with Reed-Solomon for simplicity + availability
+%   * FEC means we need small blocks of data -> Frames
+%
+%   * FEC forces us to change from double marker inside to sync sequence + frame
+
 \subsubsection{Packet Layer}
@@
         \bitheader{0,7,8,15,16,23,24,31} \\
-        \begin{rightwordgroup}{Header}
+        \begin{leftwordgroup}{Header}
             \bitbox{8}{version} & \bitbox{8}{payload ID} &
@@
             \bitbox{16}{altitude} & \bitbox[lrt]{16}{}
-        \end{rightwordgroup} \\
+        \end{leftwordgroup} \\
         \begin{leftwordgroup}{Payload}
@@
 
+\subsection{Architecture}
+
+\subsection{User Interface}
+
 \section{Ground Segment}
 
-\subsection{Reception Binary Stream Extraction}
+\subsection{Extracting Raw Binary Stream}
+
+\subsection{Packet}
 
@@
 \chapter{Results}
-\label{ch:reults}
+\label{ch:results}
 

### 2017-04-02 paper.tex
Add language on sync bytes

@@
 
-Designing and building a communication layer that can be relied up was an
+Designing and building a communication protocol that can be relied up was an
 important requirement for the rest of the project. Transmitting as much as
@@
 \begin{figure}[H]
-    \begin{bytefield}{24}
-        \bitheader{0,7,8,15,16,23} \\
-        \bitbox{8}{version} & \bitbox{16}{sequence number} \\
-            \wordbox[lrt]{1}{payload data} \\
-            \skippedwords \\
-            \wordbox[lrb]{1}{}
+    \begin{center}
+    \begin{bytefield}[bitwidth=0.65em]{64}
+        \bitheader{0,8,24,40} \\
+        \bitbox{8}{version} &
+        \bitbox{16}{sequence number} &
+        \bitbox{40}{data}
     \end{bytefield}
+    \end{center}
     \centering
-    \caption{Original AHABus Frame Format}
+    \caption{AHABus Frame Format – Second Iteration}
     \label{fig:frame-fmt-orig}
+\end{figure}
+
+This preliminary design proved impossible to work with. While it encapsulates
+the required data and metadata, it does not provide any way to identify the
+start or the end of a frame in a raw binary stream, which would have rendered
+parsing an incoming frame stream impossible.
+
+To solve this issue, a ``start sequence'' was added at the very beginning of
+the frame. When parsing frames from a binary stream, the decoder must ignore
+bytes until the start sequence (a single byte of value \texttt{0xAA} followed
+by a single byte of value \texttt{0x5A}) occurs. Once the frame sequence is
+detected, the frame's end is detected once 256 bytes have been received.
+
+\begin{figure}[H]
+    \begin{center}
+    \begin{bytefield}[bitwidth=0.65em]{64}
+        \bitheader{0,8,16,24,40} \\
+        \bitbox{8}{0xAA} & \bitbox{8}{0x5A} & \bitbox{8}{version} &
+        \bitbox{16}{sequence number} &
+        \bitbox{24}{data}
+    \end{bytefield}
+    \end{center}
+    \centering
+    \caption{AHABus Frame Format – Second Iteration}
+    \label{fig:frame-fmt-2}
+\end{figure}
+
+% Go to the magic website http://madebyevan.com/fsm/
+\begin{figure}[H]
+    \begin{center}
+    \begin{tikzpicture}[scale=0.2]
+    \tikzstyle{every node}+=[inner sep=0pt]
+    \draw [black] (23.9,-13.4) circle (3);
+    \draw (23.9,-13.4) node {$Wait$};
+    \draw [black] (53.5,-13.4) circle (3);
+    \draw (53.5,-13.4) node {$Sync$};
+    \draw [black] (38.6,-30.7) circle (3);
+    \draw (38.6,-30.7) node {$Frame$};
+    \draw [black] (26.719,-12.375) arc (107.79277:72.20723:39.209);
+    \fill [black] (50.68,-12.37) -- (50.07,-11.65) -- (49.77,-12.61);
+    \draw (38.7,-10) node [above] {$byte=AA$};
+    \draw [black] (53.11,-16.371) arc (-12.09889:-69.37588:18.591);
+    \fill [black] (41.48,-29.87) -- (42.41,-30.06) -- (42.05,-29.12);
+    \draw (49.56,-26.05) node [right] {$byte=5A$};
+    \draw [black] (35.918,-29.359) arc (-119.94439:-159.34576:25.489);
+    \fill [black] (24.79,-16.26) -- (24.61,-17.19) -- (25.54,-16.84);
+    \draw (28.67,-25.22) node [left] {$read=256$};
+    \draw [black] (51.972,-15.968) arc (-3.02212:-291.02212:2.25);
+    \draw (47.22,-18.39) node [left] {$byte=AA$};
+    \fill [black] (50.53,-13.75) -- (49.76,-13.21) -- (49.71,-14.21);
+    \draw [black] (25.724,-11.023) arc (137.60109:42.39891:17.571);
+    \fill [black] (25.72,-11.02) -- (26.63,-10.77) -- (25.89,-10.1);
+    \draw (38.7,-4.8) node [above] {$byte!=AA$};
+    \end{tikzpicture}
+    \end{center}
+    \centering
+    \caption{AHABus Frame Synchronisation}
 \end{figure}
@@
 \newpage
-\printbibliography[heading=bibnumbered]
+\printbibliography[heading=bibnumbered, title=References]
 

### 2017-04-03 paper.tex
Finish first Frame draft

@@
 \tableofcontents
+\setlength{\parskip}{1em}
 
@@
 that would have fulfilled most of the requirements, in use in other countries
-like AX.25. \todo{mention Tx only, no Rx}
+like AX.25.
+% TODO: mention Tx only, no Rx
 
@@
 of the protocol evolved or had to be changed during the prototyping process.
-\todo{Might not be necessary to say that here, given further sections}
+% TODO: Might not be necessary to say that here, given further sections
 
@@
 
-\todo{Add some details about simplicity of the network topology}
+% TODO: Add some details about simplicity of the network topology
 
@@
 data units, and provides reliability and error correction to higher-level 
-layers.\todo{Add vocabulary maybe (ancillary)}
+layers.
+% TODO: Add vocabulary maybe (ancillary)
 
@@
 
-\todo{add an introductory sentence?}
+% TODO: add an introductory sentence?
 
@@
 
+% TODO: references for DominoEx and Olivia
+
 Other existing protocols exist that can transmit data over radio audio signals,
-\todo{references for DominoEx and Olivia}
 like DominoEx or Olivia, sometimes with better transmission rates. RTTY was
@@
 
-\todo{explain requirement for sync marker}
+% TODO: explain requirement for sync marker
 
@@
     \begin{center}
-    \begin{bytefield}[bitwidth=0.65em]{64}
+    \begin{bytefield}[bitwidth=0.5em]{64}
         \bitheader{0,8,24,40} \\
         \bitbox{8}{version} &
-        \bitbox{16}{sequence number} &
-        \bitbox{40}{data}
+        \bitbox{16}{seq. number} &
+        \bitbox{40}{data[253]}
     \end{bytefield}
@@
     \centering
-    \caption{AHABus Frame Format – Second Iteration}
+    \caption{AHABus Frame Format – Version 1}
     \label{fig:frame-fmt-orig}
@@
     \begin{center}
-    \begin{bytefield}[bitwidth=0.65em]{64}
+    \begin{bytefield}[bitwidth=0.5em]{64}
         \bitheader{0,8,16,24,40} \\
-        \bitbox{8}{0xAA} & \bitbox{8}{0x5A} & \bitbox{8}{version} &
-        \bitbox{16}{sequence number} &
-        \bitbox{24}{data}
+        \bitbox{8}{\texttt{0xAA}} & \bitbox{8}{\texttt{0x5A}} & \bitbox{8}{version} &
+        \bitbox{16}{seq. number} &
+        \bitbox{24}{data[251]}
     \end{bytefield}
@@
     \centering
-    \caption{AHABus Frame Format – Second Iteration}
+    \caption{AHABus Frame Format – Version 2}
     \label{fig:frame-fmt-2}
@@
 
-% Go to the magic website http://madebyevan.com/fsm/
+The second version of the frame format shown in figure \ref{fig:frame-fmt-2}
+fulfils one of its two design goals (detection of data loss). However, it does
+not provide a way to detect and correct errors due to noise or interferences on
+the radio channel.
+
+The error-correction requirement led to the use of a Forward Error Correction
+(\textit{FEC}) algorithm. FEC algorithms function by adding some amount of data
+\(D\) to each message sent \(M\). \(D\) is computed by the encoder. The decoder
+uses \(D\) to detect whether errors were introduced anywhere in the received
+block \(M+D\), and correct those errors if their number is below the
+algorithm's maximum.
+
+% TODO: block diagram of the encoding-decoding chain
+
+For the purpose of this project, Reed-Solomon(255,223) was chosen for multiple
+reasons:
+
+\begin{itemize}
+
+\item The Reed-Solomon family of algorithms is widespread in its use, notably
+in space communications systems – the Voyager probes used Reed-Solomon codes to
+encode their images (\cite{Wicker1994}), the Space Data Links frames (devised
+by ESA and CCSDS) use a variant of Reed-Solomon to provide FEC
+(\cite{EuropeanCooperationforSpaceStandardization2010}) and the SSDV protocol,
+used in High-Altitude Balloon missions to transmit low-resolution images,
+uses RS(255,223) with 256-byte frames (\cite{UKHAS2016}).
+
+\item There were existing, well tested open-source implementation of that
+specific variant of the Reed-Solomon encoders and decoders, which meant that
+more time could be spent working on other parts of the project.
+
+% TODO: Ask Ian, transparent might not be the right technical term
+
+\item Reed-Solomon codes are transparent: the original data is not modified for
+the purposes of error correction. This means that even if too many errors are
+present for the FEC decoder to correct, the data can be manually inspected and
+some of it can potentially be saved.
+
+\end{itemize}
+
+When applied to 8-bit \textit{symbols} (in this project's case, bytes),
+Reed-Solomon codes can only handle blocks (Data and Error-Correction Codes) of
+a length up to \(2^{8}-1 = 255\) bytes. The chosen variant, \(RS(223,255)\)
+works on 223-byte messages, and adds 32 bytes of Error-Correction Codes at the
+end to produce 255-byte blocks. This prompted slight changes in the frame
+format: each frame being 256-byte long, one byte would not have been covered
+by error correction.
+
+The chosen solution was to not include the start sequence as part of the
+error-corrected message. Since proper detection of that sequence is required
+for the decoder to start parsing a frame, it would have been impossible to apply
+FEC to a frame whose frame start sequence is corrupted.
+
+\begin{figure}[H]
+    \begin{center}
+    \begin{bytefield}[bitwidth=1.8em]{20}
+        \bitbox{8}{frame} &
+        \bitbox{8}{frame} &
+        \bitbox{4}{...}
+    \end{bytefield}
+    \end{center}
+    \centering
+    \caption{Frame stream with embedded start sequence}
+    \label{fig:frame-sync-emb}
+\end{figure}
+
+The start sequence design had to be changed to occupy a single frame byte, as
+opposed to two bytes in version 2 of the frame format. Using a single byte to
+detect the start of a frame was not deemed sufficient, since the probably of
+that exact value occurring elsewhere in the binary stream would have been
+fairly high. To mitigate this, the third and final version of the frame format
+uses a single frame start marker (\texttt{0x5A}) at the beginning of each
+frame, and each frame must be preceded by at least one synchronisation byte
+(\texttt{0xAA}), inserted in the binary stream as shown in the lower half of 
+figure \ref{fig:frame-sync-extra}. The following 223 bytes in the frame
+(starting with the protocol version number) are error-corrected, with the
+frame's 32 trailing bytes.
+
+\begin{figure}[H]
+    \begin{center}
+    \begin{bytefield}[bitwidth=1.8em]{20}
+        \bitbox{1}{\texttt{AA}} &
+        \bitbox{8}{frame} &
+        \bitbox{1}{\texttt{AA}} &
+        \bitbox{8}{frame} &
+        \bitbox{2}{...}
+    \end{bytefield}
+    \end{center}
+    \centering
+    \caption{Frame stream with external synchronisation byte}
+    \label{fig:frame-sync-extra}
+\end{figure}
+
+While this approach may seem wasteful since at least one extra byte must be
+inserted in between each frame, it is to be noted that the same synchronisation
+byte was previously part of each frame's 256 bytes, leading to a smaller amount
+of data being encapsulated in each frame.
+
 \begin{figure}[H]
@@
     \fill [black] (50.68,-12.37) -- (50.07,-11.65) -- (49.77,-12.61);
-    \draw (38.7,-10) node [above] {$byte=AA$};
+    \draw (38.7,-10) node [above] {$\texttt{byte=0xAA}$};
     \draw [black] (53.11,-16.371) arc (-12.09889:-69.37588:18.591);
     \fill [black] (41.48,-29.87) -- (42.41,-30.06) -- (42.05,-29.12);
-    \draw (49.56,-26.05) node [right] {$byte=5A$};
+    \draw (49.56,-26.05) node [right] {$\texttt{byte=0x5A}$};
     \draw [black] (35.918,-29.359) arc (-119.94439:-159.34576:25.489);
     \fill [black] (24.79,-16.26) -- (24.61,-17.19) -- (25.54,-16.84);
-    \draw (28.67,-25.22) node [left] {$read=256$};
+    \draw (28.67,-25.22) node [left] {$\texttt{read=256}$};
     \draw [black] (51.972,-15.968) arc (-3.02212:-291.02212:2.25);
-    \draw (47.22,-18.39) node [left] {$byte=AA$};
+    \draw (47.22,-18.39) node [left] {$\texttt{byte=0xAA}$};
     \fill [black] (50.53,-13.75) -- (49.76,-13.21) -- (49.71,-14.21);
@@
     \fill [black] (25.72,-11.02) -- (26.63,-10.77) -- (25.89,-10.1);
-    \draw (38.7,-4.8) node [above] {$byte!=AA$};
+    \draw (38.7,-4.8) node [above] {$\texttt{byte!=0xAA}$};
     \end{tikzpicture}
@@
     \centering
-    \caption{AHABus Frame Synchronisation}
+    \caption{AHABus Frame Synchronisation State Machine}
+    \label{fig:frame-fsm}
 \end{figure}
 
-\todo{explain move away from double-marker towards sequence + mark}
+Figure \ref{fig:frame-fsm} presents the finite state machine that AHABus frame
+decoders must implement to detect frames in a binary stream, and figure
+\ref{fig:frame-fmt-3} the final version of the AHABus frame format.
 
-Encapsulating the transmitted data adds overhead. If \(n\) is the size of a
-chunk in bytes, and each frame adds \(m\) bytes of metadata, then only a
-percentage \(\frac{n}{n+m}\) of the stream contains actual data. For reasons
-explained further, the AHABus frames were chosen to be 256-byte long. As shown
-previously, each frame also includes.
+% Go to the magic website http://madebyevan.com/fsm/
 
@@
         \begin{rightwordgroup}{Header}
-            \bitbox{8}{01011010} & \bitbox{8}{version} \\
+            \bitbox{8}{\texttt{0x5A}} & \bitbox{8}{version} \\
             \bitbox{16}{sequence number}
@@
     \centering
-    \caption{AHABus Frame Format}
+    \caption{AHABus Frame Format - Version 3}
+    \label{fig:frame-fmt-3}
 \end{figure}
+
+Encapsulating the transmitted data adds overhead. If \(n\) is the size of a
+chunk in bytes, and each frame adds \(m\) bytes of metadata, then only a
+percentage \(\frac{n}{n+m}\) of the stream contains actual data. For reasons
+explained further, the AHABus frames were chosen to be 256-byte long. As shown
+previously, each frame also includes.
 

### 2017-04-03 paper.tex
Add basic packet stuf

@@
 
+% TODO: add something about:
+%       - hardware choices
+%       - software choices
+%       - language choices
+%
+%       - overall arch (flight platform, ground segmetn)
+
 The proposed solution for this project is inspired by the bus model that
@@
 \begin{enumerate}
-\item provide a system to transmit binary data.
-\item minimise the amount of overhead and bandwidth required.
-\item allow multiple payload's data to be sent on the same link.
-\item provide stable transmissions on a low-power link.
-\item correct as many transmission-induced data errors as possible.
-\item allow ground crew to track the balloon's location over time.
+\item provide a system to transmit binary data.% DATA LINK
+\item minimise the amount of overhead and bandwidth required.% ALL
+\item allow multiple payload's data to be sent on the same link.% PACKET LAYER
+\item provide stable transmissions on a low-power link.% FRAME LAYER
+\item correct as many transmission-induced data errors as possible.% FRAME LAYER
+\item allow ground crew to track the balloon's location over time.% PACKET LAYER
 \end{enumerate}
@@
 
-% TODO: references for DominoEx and Olivia
-
 Other existing protocols exist that can transmit data over radio audio signals,
@@
 \subsubsection{Frame Layer}
+\label{sssec:frame-layer}
 
@@
 
-% TODO: explain requirement for sync marker
-
 The resulting first version of the frame format is presented in figure
@@
 
-This preliminary design proved impossible to work with. While it encapsulates
+This preliminary design proved impossible to work with: while it encapsulates
 the required data and metadata, it does not provide any way to identify the
@@
     \begin{bytefield}[bitwidth=1.8em]{20}
+        \bitbox{1}{...} &
         \bitbox{8}{frame} &
         \bitbox{8}{frame} &
-        \bitbox{4}{...}
+        \bitbox{3}{...}
     \end{bytefield}
@@
     \begin{bytefield}[bitwidth=1.8em]{20}
+        \bitbox{1}{...} &
         \bitbox{1}{\texttt{AA}} &
@@
         \bitbox{8}{frame} &
-        \bitbox{2}{...}
+        \bitbox{1}{...}
     \end{bytefield}
@@
         \bitheader{0,7,8,15} \\
-        \begin{rightwordgroup}{Header}
+        \begin{leftwordgroup}{Header}
             \bitbox{8}{\texttt{0x5A}} & \bitbox{8}{version} \\
             \bitbox{16}{sequence number}
-        \end{rightwordgroup} \\
-        \begin{rightwordgroup}{Payload}
+        \end{leftwordgroup} \\
+        \begin{leftwordgroup}{Payload}
             \wordbox[lrt]{1}{data} \\
@@
             \wordbox[lrb]{1}{}
-        \end{rightwordgroup} \\
-        \begin{rightwordgroup}{FEC}
+        \end{leftwordgroup} \\
+        \begin{leftwordgroup}{FEC}
             \wordbox[lrt]{1}{checksum} \\
@@
             \wordbox[lrb]{1}{}
-        \end{rightwordgroup}
+        \end{leftwordgroup}
     \end{bytefield}
@@
 \end{figure}
+
+%TODO: what's that worth? shouldn't it be at the end
 
@@
 
-% Topic list for this:
-%   * Provide reliability:
-%       * Data arrives whole
-%       * We receive valid data
-%   * receive all data: we chunk the data stream, number the chunks         v
-%   * we tag data with housekeeping things                                  v~
-%   * If frames too small, housekeeping/data ratio is shit
-%   * Need a way to detect the start + end of a frame.
-%   * Explain that end of frame marker complicated, going with count?
-%
-%   * Validity checked (and corrected using FEC)
-%   * FEC: went with Reed-Solomon for simplicity + availability
-%   * FEC means we need small blocks of data -> Frames
-%
-%   * FEC forces us to change from double marker inside to sync sequence + frame
+\subsubsection{Packet Layer}
 
-\subsubsection{Packet Layer}
+The Packet Layer relies on the stable data stream provided by the Link and Frame
+Layers, and was designed to allow the multiplexing of multiple payload's data as
+well as the transmission of ancillary data required for the overall flight
+management (platform's GNSS co-ordinates and altitude).
+
+To achieve those goals, the AHABus protocol uses packets similar to frames. Each
+packet encapsulates a chunk of data retrieved from a single payload. To that
+end, packets were designed to be of variable length, to avoid unnecessary
+padding, which would have wasted bandwidth. The data - referred to as the
+packet's \textit{payload} – is preceded by the packet's header, which contains
+information necessary to attribute the packet's data to a payload, indicate
+the length of the packet, and carry the AHABus platform's location and altitude.
+
+This design achieves multiplexing through time division: packets are chunked
+and sent in frames one after the other, and can be dispatched to their
+respective payload teams once decoded at the ground station. The main issue with
+that method is that the protocol itself does not ensure that each payload gets
+an even or fair share of the total bandwidth. As discussed in section
+\ref{sec:flight-software}, this was solved in the flight software design.
 
@@
             \bitbox{8}{version} & \bitbox{8}{payload ID} &
-            \bitbox{16}{packet length} \\
+            \bitbox{16}{length} \\
             \bitbox{32}{latitude} \\
@@
         \begin{leftwordgroup}{Payload}
-            \wordbox[lrt]{1}{data} \\
-            \skippedwords \\
-            \wordbox[lrb]{1}{}
+            \wordbox[lrb]{3}{data[length]}
         \end{leftwordgroup}
@@
     \centering
-    \caption{AHABus Frame Format}
+    \caption{AHABus Packet Format - Version 1}
+    \label{fig:packet-fmt-1}
 \end{figure}
+
+Figure \ref{fig:packet-fmt-1} shows the first and current version of the AHABus
+packet format. Like frame discussed in section \ref{sssec:frame-layer}, each
+packet carries a version flag which was added to allow backwards compatibility,
+were the protocol to evolve in the future.
+
+The encoding of the metadata was chosen based on the physical constraints of
+the information:
+
+\begin{itemize}
+\item Latitude and Longitudes can be both positive (North, East) or Negative
+(South, West). At the equator, \({0.0001}^\circ\) \footnotemark of
+longitude maps to roughly 11 metres, and that distance decreases when moving
+towards the poles, which was deemed precise enough for the project's purpose.
+To avoid conflicts in floating-point mathematics conventions, and allow packet
+encoders to be written for embedded platform without hardware floating point
+support, latitude and longitude were designed to be multiplied by \(10,000\)
+and transmitted as 32-bit precision signed integers.
+
+\item Altitude is positive for the whole mission duration, and cannot exceed
+\(40,000m\). Since a sub-metre precision is not required, and would hardly be
+achievable with civilian GNSS receivers, it is encoded as a 16-bit precision
+unsigned integer.
+\end{itemize}
+
+\footnotetext{
+Earth circumference at the equator: \(40\time10^{6}m\).
+\(D_{0.0001^{\circ}} = 0.0001\frac{40\time10^{6}}{360}=11.1...m\)
+}
 
@@
 
-\section{Bus and Flight Software}
+\section{Payload Data Bus}
 
@@
 \section{Flight Software}
+\label{sec:flight-software}
 

### 2017-04-04 paper.tex
Add first packet layout draft

@@
 \item Latitude and Longitudes can be both positive (North, East) or Negative
-(South, West). At the equator, \({0.0001}^\circ\) \footnotemark of
-longitude maps to roughly 11 metres, and that distance decreases when moving
-towards the poles, which was deemed precise enough for the project's purpose.
-To avoid conflicts in floating-point mathematics conventions, and allow packet
-encoders to be written for embedded platform without hardware floating point
-support, latitude and longitude were designed to be multiplied by \(10,000\)
-and transmitted as 32-bit precision signed integers.
+(South, West). At the equator, \({0.0001}^\circ\) of longitude maps to roughly
+11 metres\footnotemark , and that distance decreases when moving towards the
+poles, which was deemed precise enough for the project's purpose. To avoid
+conflicts in floating-point mathematics conventions, and allow packet encoders
+to be written for embedded platform without hardware floating point support,
+latitude and longitude were designed to be multiplied by 10,000 and transmitted
+as 32-bit precision signed integers.
 
-\item Altitude is positive for the whole mission duration, and cannot exceed
-\(40,000m\). Since a sub-metre precision is not required, and would hardly be
+\item Altitude is above zero for the whole mission duration, and cannot exceed
+50,000m. Since a sub-metre precision is not required, and would hardly be
 achievable with civilian GNSS receivers, it is encoded as a 16-bit precision
@@
 \footnotetext{
-Earth circumference at the equator: \(40\time10^{6}m\).
+Earth circumference at the equator: \(40\time10^{6}\)m.
 \(D_{0.0001^{\circ}} = 0.0001\frac{40\time10^{6}}{360}=11.1...m\)
 }
+
+The decision was made during the design process to align packet boundaries with
+frame boundaries. This means that every time a new packet is sent through the
+radio channel starts at the start of a new frame, instead of being inserted
+in the middle of a frame, where the last packet stopped.
+
+This approach leads to a loss in bandwidth: if a packet's length is not a 
+multiple of the amount of data that a frame carry, the packet's last frame
+will be padded with zeros instead of containing the start of the next packet.
+However, it was deemed an acceptable tradeoff because of the issues that
+non-aligned (contiguous) packets would present:
+
+\begin{itemize}
+\item If a packet's end does not match a frame's end, that frame cannot be
+sent until the start of the next packet is known. This meant that the
+transmission of most packet's last frames could be delayed for sometimes several
+seconds until another packet is processed.
+
+\item In the case where a frame is lost due to channel noise or interferences,
+finding the start of the next packet would have been complicated, or required
+a synchronisation mechanism similar to the one designed for AHABus frames.
+Aligning packets instead makes such a system unnecessary, reducing the packet
+header size and the complexity of packet decoders.
+\end{itemize}
+
+% TODO: Add diagram explaining aligment
 

### 2017-04-04 glossaries.tex
Add glossary

\usepackage{datatool}
\usepackage[acronym,toc]{glossaries}
\makeglossaries

\newacronym{hab}{HAB}{High-Altitude Balloon}
\newacronym{gnss}{GNSS}{Global Navigation Satellite System}
\newacronym{obc}{OBC}{On Board Computer}
\newacronym{ahabus}{AHABus}{Abertay High-Altitude Bus}
\newacronym{fec}{FEC}{Forward Error Correction}
\newacronym{i2c}{I\textsuperscript{2}C}{Inter-Integrated Circuit}
\newacronym{ccsds}{CCSDS}{Consultative Committee for Space Data Systems}
\newacronym{ecc}{ECC}{Error-Correction Codes}
\newacronym{rtty}{RTTY}{Radio-Teletype}

### 2017-04-04 paper.tex
Add glossary

@@
 \input{header.tex}
+\input{glossaries.tex}
 
 \begin{document}
+
+\frontmatter
 
@@
 \begin{abstract}
-
 \end{abstract}
 
-\newpage
 \tableofcontents
@@
 
-\newpage
+\listoffigures
+
+\listoftables
+
+\printglossary
+\printglossary[type=\acronymtype]
+
+\mainmatter
+
 \chapter{Introduction}
@@
 
-Most HAB missions led by universities, schools or amateur are built around
-commercially available small computers or micro-controller boards: Redland
-Green School students used a Raspberry Pi (\cite{rpi2014}) running
-python scripts as an On Board Computer (\textit{OBC}) for their HAB mission
+Most \acrshort{hab} missions led by universities, schools or amateur are built 
+around commercially available small computers or micro-controller boards:
+Redland Green School students used a Raspberry Pi (\cite{rpi2014}) running
+python scripts as an \acrfull{obc} for their HAB mission
 (\cite{Hinschelwood2015}), while other projects have been ran on AVR
@@
 constraints, these missions usually transmit their telemetry as text using the
-Radio-Teletype (RTTY) protocol.
+\acrfull{rtty} protocol.
 
-While most HAB projects make use of commercial-off-the-shelf components, the
-finished payloads are usually custom, single-purpose designs that are little,
-if at all modular. In the design document for the Titan-1 HAB mission, Bombasaro
-describes a hardware bus based on tightly coupled sensors which communicate
-over two different data buses, SPI and I2C, and a custom-made, single-purpose
-flight software (\cite{Bombasaro2015}).
+While most \acrshort{hab} projects make use of commercial-off-the-shelf
+components, the finished payloads are usually custom, single-purpose designs
+that are little, if at all modular. In the design document for the Titan-1 HAB
+mission, Bombasaro describes a hardware bus based on tightly coupled sensors
+which communicate over two different data buses, SPI and \acrshort{i2c}, and a
+custom-made, single-purpose flight software (\cite{Bombasaro2015}).
 
@@
 is a based on a custom made circuit board, it is designed to provide a standard
-power interface, as well as access to an I2C standard bus
-(\cite{NXPSemiconductors2014}) to each payload (\cite{Volstad2011}). The I2C
-protocol was chosen because of its low power consumption, and because if only
-requires two lines (clock and data). Thomas Clausen describes in his 2001 paper
-how a simple packet protocol built on top of I2C itself is used for data
-transfer and error detection in Aalborg University's CubeSats
-(\cite{Clausen2001}).
+power interface, as well as access to an \acrshort{i2c} standard bus
+(\cite{NXPSemiconductors2014}) to each payload (\cite{Volstad2011}). The
+\acrshort{i2c} protocol was chosen because of its low power consumption, and
+because if only requires two lines (clock and data). Thomas Clausen describes
+in his 2001 paper how a simple packet protocol built on top of \acrshort{i2c}
+itself is used for data transfer and error detection in Aalborg University's
+CubeSats (\cite{Clausen2001}).
 
@@
 There are however some technologies that can be adapted from satellites to be
-used in HAB applications. The University of Arizona uses a custom packet format
-over 434MHz radio, containing raw binary values for each instrument's
-measurements (\cite{Eatchel2002}). The BRITE-Austria CubeSat mission uses the
-AX.25 packet radio protocol (used by amateur radio users), which allows the
-use of off-the-shelf transmission and reception hardware rather than custom-made
-circuits (\cite{Traussnig2007}).
+used in \acrshort{hab} applications. The University of Arizona uses a custom
+packet format over 434MHz radio, containing raw binary values for each
+instrument's measurements (\cite{Eatchel2002}). The BRITE-Austria CubeSat
+mission uses the AX.25 packet radio protocol (used by amateur radio users),
+which allows the use of off-the-shelf transmission and reception hardware
+rather than custom-made circuits (\cite{Traussnig2007}).
 
@@
 limits imposed in the United Kingdom, such a selective telemetry system cannot
-be relied upon for HAB missions.
+be relied upon for \acrshort{hab} missions.
 
@@
 remotely poll their memory (\cite{parkes2005}). A similar system is used to
-fetch data from ROM chips over I2C and could be used to build a simple data bus.
+fetch data from ROM chips over \acrshort{i2c} and could be used to build a
+simple data bus.
 
-The CCSDS Space Packet protocol defines a packet format used
-to encapsulate application data (from instruments for example) that can be sent
+The \acrshort{ccsds} Space Packet protocol defines a packet format used to
+encapsulate application data (from instruments for example) that can be sent
 over a ``Space Link'', an analogue of the OSI model's link layer
-(\cite{Stallings1987}) using CCSDS frames to encapsulate packets originating
-from multiple instruments, ground stations and spacecrafts (\cite{ccsds2003}).
+(\cite{Stallings1987}) using \acrshort{ccsds} frames to encapsulate packets
+originating from multiple instruments, ground stations and spacecrafts
+(\cite{ccsds2003}).
 
-While HAB applications do not require the level of complexity of CCSDS
-protocols — addressing is not needed since the network only contains two
-endpoints, communicating in a single direction — some details of the Space
-Packet protocol are worth reusing (Forward Error Correction, different
-Application IDs for each payload).
+While \acrshort{hab} applications do not require the level of complexity of
+\acrshort{ccsds} protocols — addressing is not needed since the network only
+contains two endpoints, communicating in a single direction — some details of
+the Space Packet protocol are worth reusing (Forward Error Correction,
+different Application IDs for each payload).
 
@@
 (\textit{FEC}) algorithm. FEC algorithms function by adding some amount of data
-\(D\) to each message sent \(M\). \(D\) is computed by the encoder. The decoder
+\(D\) to each message sent \(M\). In this case, the message includes both the
+frame's header and its payload. \(D\) is computed by the encoder. The decoder
 uses \(D\) to detect whether errors were introduced anywhere in the received
@@
 the purposes of error correction. This means that even if too many errors are
-present for the FEC decoder to correct, the data can be manually inspected and
-some of it can potentially be saved.
+present for the \acrshort{fec} decoder to correct, the data can be manually
+inspected and some of it can potentially be saved.
 
@@
 
-Figure \ref{fig:frame-fsm} presents the finite state machine that AHABus frame
-decoders must implement to detect frames in a binary stream, and figure
-\ref{fig:frame-fmt-3} the final version of the AHABus frame format.
+Figure \ref{fig:frame-fsm} presents the finite state machine that
+\acrshort{ahabus} frame decoders must implement to detect frames in a binary
+stream, and figure \ref{fig:frame-fmt-3} the final version of the
+\acrshort{ahabus} frame format.
 
@@
         \begin{leftwordgroup}{Payload}
-            \wordbox[lrt]{1}{data} \\
-            \skippedwords \\
-            \wordbox[lrb]{1}{}
+            \wordbox{2}{data[]}
         \end{leftwordgroup} \\
         \begin{leftwordgroup}{FEC}
-            \wordbox[lrt]{1}{checksum} \\
-            \skippedwords \\
-            \wordbox[lrb]{1}{}
+            \wordbox{2}{checksum[32]}
         \end{leftwordgroup}
@@
     \centering
-    \caption{AHABus Frame Format - Version 3}
+    \caption{\acrshort{ahabus} Frame Format - Version 3}
     \label{fig:frame-fmt-3}
@@
 percentage \(\frac{n}{n+m}\) of the stream contains actual data. For reasons
-explained further, the AHABus frames were chosen to be 256-byte long. As shown
-previously, each frame also includes.
+explained further, the \acrshort{ahabus} frames were chosen to be 256-byte
+long. As shown previously, each frame also includes.
 
@@
 
+To evaluate the protocol's viability, a reference implementation of an encoder
+and decoder was written. The goal was to
+
 \section{Payload Data Bus}
@@
 
-\newpage
-\printbibliography[heading=bibnumbered, title=References]
+\printbibliography[heading=bibintoc, title=References]
 

### 2017-04-06 paper.tex
various fixes

@@
 
+To avoid possible byte order conflicts between the architectures used on the
+flight platform and the ground station, multi-bytes values referred to in the
+specification should always be transmitted as big-endian (most significant byte
+first).
+
 The design process and specification of each of those three layers is explained
@@
 
-The valid encoding and decoding of a RTTY signal depends on the chosen baud
-rate (the number of times the logical signal changes level). During tests,
-the highest baud rate that allowed clear decoding of radio signals at 10mW
-power rating was 200bauds. Since each byte requires eleven bits to be sent
-(one start bit, eight data bits and two stop bits), this corresponds to a
-\(\frac{200}{11}=18.18\) bytes per second bandwidth.
+The valid encoding and decoding of a \acrshort{rtty} signal depends on the
+chosen baud rate (the number of times the logical signal changes level). During
+tests, the highest baud rate that allowed clear decoding of radio signals at
+\SI{10}{mW} power rating was \SI{200}{bps}. Since each byte requires eleven
+bits to be sent (one start bit, eight data bits and two stop bits), this
+corresponds to a \(\frac{200}{11}=\SI{18.18}{Bps}\) bandwidth.
 
@@
 encode their images (\cite{Wicker1994}), the Space Data Links frames (devised
-by ESA and CCSDS) use a variant of Reed-Solomon to provide FEC
-(\cite{EuropeanCooperationforSpaceStandardization2010}) and the SSDV protocol,
-used in High-Altitude Balloon missions to transmit low-resolution images,
-uses RS(255,223) with 256-byte frames (\cite{UKHAS2016}).
+by \acrshort{esa} and \acrshort{ccsds}) use a variant of Reed-Solomon to
+provide \acrshort{fec} (\cite{EuropeanCooperationforSpaceStandardization2010})
+and the SSDV protocol, used in High-Altitude Balloon missions to transmit
+low-resolution images, uses RS(255,223) with 256-byte frames (\cite{UKHAS2016}).
 
@@
 for the decoder to start parsing a frame, it would have been impossible to apply
-FEC to a frame whose frame start sequence is corrupted.
+\acrshort{fec} to a frame whose frame start sequence is corrupted.
 
@@
 Figure \ref{fig:frame-fsm} presents the finite state machine that
-\acrshort{ahabus} frame decoders must implement to detect frames in a binary
+AHABus frame decoders must implement to detect frames in a binary
 stream, and figure \ref{fig:frame-fmt-3} the final version of the
-\acrshort{ahabus} frame format.
+AHABus frame format.
 
@@
     \centering
-    \caption{\acrshort{ahabus} Frame Format - Version 3}
+    \caption{AHABus Frame Format - Version 3}
     \label{fig:frame-fmt-3}
@@
 percentage \(\frac{n}{n+m}\) of the stream contains actual data. For reasons
-explained further, the \acrshort{ahabus} frames were chosen to be 256-byte
+explained further, the AHABus frames were chosen to be 256-byte
 long. As shown previously, each frame also includes.
@@
 \item Latitude and Longitudes can be both positive (North, East) or Negative
-(South, West). At the equator, \({0.0001}^\circ\) of longitude maps to roughly
-11 metres\footnotemark , and that distance decreases when moving towards the
-poles, which was deemed precise enough for the project's purpose. To avoid
-conflicts in floating-point mathematics conventions, and allow packet encoders
-to be written for embedded platform without hardware floating point support,
-latitude and longitude were designed to be multiplied by 10,000 and transmitted
-as 32-bit precision signed integers.
+(South, West). At the equator, \SI{0.0001}{\degree} of longitude maps to
+roughly eleven metres\footnotemark , and that distance decreases when moving
+towards the poles, which was deemed precise enough for the project's purpose.
+To avoid conflicts in floating-point mathematics conventions, and allow packet
+encoders to be written for embedded platform without hardware floating point
+support, latitude and longitude were designed to be multiplied by 10,000 and
+transmitted as 32-bit precision signed integers.
 
 \item Altitude is above zero for the whole mission duration, and cannot exceed
-50,000m. Since a sub-metre precision is not required, and would hardly be
-achievable with civilian GNSS receivers, it is encoded as a 16-bit precision
-unsigned integer.
+\SI{50000}{\meter}. Since a sub-metre precision is not required, and would
+hardly be achievable with civilian GNSS receivers, it is encoded as a 16-bit
+precision unsigned integer.
+
 \end{itemize}
@@
 \footnotetext{
-Earth circumference at the equator: \(40\time10^{6}\)m.
-\(D_{0.0001^{\circ}} = 0.0001\frac{40\time10^{6}}{360}=11.1...m\)
+Earth circumference at the equator: \SI{40e6}{\meter}.
+\(D_{\SI{0.0001}{\degree}} = 0.0001\times\frac{40\times{10^6}}{360}=\SI{11.1}{\meter}\)
 }
@@
 
-To evaluate the protocol's viability, a reference implementation of an encoder
-and decoder was written. The goal was to
+To evaluate the protocol's viability, a full demonstration of an AHABus radio
+stack was written. Its main goals were to provide a reference implementation of
+a compliant encoder and decoder, validate the protocol's design process, and
+test the algorithms used in a controlled environment before integrated testing.
 
-\section{Payload Data Bus}
+The C language (ISO/IEC 9899:1999 standard) was chosen for the reference
+implementation because of its widespread use, and its availability on most
+micro-controller and embedded platforms, which would be used as AHABus flight
+platforms. To render the code as portable and reusable as possible, care was
+taken to limit the dependencies to a subset of the C language standard library,
+and avoid use of dynamic memory allocation, which is often inefficient or
+unavailable on embedded platforms.
+
+To increase portability and ease of use further, the encoder and decoder read
+and write data using user-supplied callback. This allows developers to import
+the implementation files and use the library without editing, at the expense of
+optimisation.
+% TODO: Explain?
+
+\subsubsection{Encoder}
+
+Figure \ref{fig:encoder-flow} illustrate the data flow in the encoder.
+In an attempt to use as little memory as possible, no intermediate
+representation of the packet is created or stored. Instead, the header and
+first chunk of the packet's data is written directly to a frame buffer. Once
+the frame is fully constructed, its binary representation is written using the
+user-provided callback. The buffer is then cleared, and the process is repeated
+until all of the packet's payload has been framed.
+
+\begin{figure}[H]
+\includegraphics[width=0.6\textwidth]{encoder-flow.pdf}
+\centering
+\caption{Encoder Data Flow}
+\label{fig:encoder-flow}
+\end{figure}
+
+For testing purposes, a simple program was built that takes in user input,
+uses the encoder to package that data and outputs the frame stream to the
+UNIX standard output (\texttt{stdout}).
+
+% TODO: more details? how much is required?
+% TODO: add reference to appendix with code
+
+\subsubsection{Decoder}
+
+\subsubsection{Initial Testing}
+
+\section{Payload Bus}
+\label{sec:payload-bus}
+
+The Payload Bus is the hardware and software interface that is used by the
+\acrfull{obc} to communicate with each payload carried on a mission.
 
 \subsection{Requirements}
+
+\subsection{Initial Prototypes}
+
+Before a protocol was chosen, two different preliminary designs approaches were
+prototyped and
+
+\subsubsection
 
@@
 \section{Ground Segment}
+\label{sec:ground-segment}
 
@@
 \section{Testing}
+\label{sec:testing}
 

### 2017-04-06 paper.tex
Add Serial Payload Bus

@@
 
-Before a protocol was chosen, two different preliminary designs approaches were
-prototyped and
+Before a protocol was chosen, two different preliminary designs were devised
+and experimented with outside of the full flight platform to ensure they were
+viable, and which of the two was worth developing further.
 
-\subsubsection
+\subsubsection{Serial Bus Interface}
+
+The first experimental bus interface was designed around a Serial interface.
+Serial interfaces (based on \acrshort{uart}) are only meant to connect two
+devices, and require two data wires (one for each transmission direction). To
+avoid requiring a pair of input/output pins on the \acrlong{obc} for each
+connected payload, and simplify the electronic layout, all payloads were wired
+in parallels on the same reception (Rx) and transmission (Rx) data lines. When
+a payload had any data ready to be sent, it would transmit it using
+\acrshort{uart} to the \acrshort{obc}, which would then store it in memory.
+
+This configuration proved unusable. Because \acrshort{uart} is not designed
+to handle multiple devices on the same connection endpoint, data would be
+corrupted whenever two or more payloads were attempting to send data to the
+\acrshort{obc}.
+
+The chosen solution for this problem was to make four bus lines available to
+each payload. In addition to the common Rx/Tx lines, each interface is composed
+of a Communication Request Line (CR) and a Communication Enable Line (CE). With
+the new design, the communication sequence had to be changed to the following:
+
+\begin{enumerate}
+\item When a payload is ready to send data, it pulls its CR line high.
+\item The \acrshort{obc} lists payloads that have raised CR. For each:
+\begin{enumerate}
+    \item The \acrshort{obc} pulls the payload's CE line high.
+    \item The payload sends its data to the \acrshort{obc}.
+    \item The \acrshort{obc} pulls the payload's CE line low.
+    \item The payload pulls its CR line low.
+    \item The \acrshort{obc} processes the received data.
+\end{enumerate}
+\end{enumerate}
+
+To ensure that non-cooperative payload could not attempt to send data when the
+\acrshort{obc} had not addressed them through pulling the CE line high, a logic
+AND gate (two NAND gates chained together as shown in figure \ref{fig:uart-nor})
+was placed between the payload's Tx line and its CE line. With this design,
+any signal on the Tx line could not get past the gate unless the payload was
+being addressed, thus preventing a payload from corrupting another's signal.
+
+\begin{figure}[H]
+\includegraphics[width=0.6\textwidth]{serial-nand.png}
+\centering
+\caption{Payload Tx line protection}
+\label{fig:uart-nor}
+\end{figure}
+
+This solution was tested using a Raspberry Pi 3 for the \acrlong{obc}, and an
+Arduino UNO board for the payload. An unforeseen electrical engineering issue
+required slight redesign of the system: on both the Arduino or the Raspberry
+Pi, a small voltage, even inferior to the logic high (\SI{3.3}{V} on Raspberry
+Pi, \SI{5}{V} on the Arduino UNO) was enough for the line to be read as a
+logical high. This caused certain bursts of noise in the lines to be detected
+by the \acrshort{obc} as a high CR line, even though the payload had not
+raised it. This was fixed by using inverted logic: a payload would signal 
+available data by pulling the \(\overline{CE}\) line low, and the \acrshort{obc}
+would pull \(\overline{CR}\) low when addressing a payload.
+
+\begin{figure}[H]
+\includegraphics[width=0.6\textwidth]{serial-wires.jpg}
+\centering
+\caption{Payload Serial Bus test setup}
+\label{fig:uart-nor}
+\end{figure}
+
+The system worked with the experimental setup, and the \acrlong{obc} and the
+OBC were able to exchange data. However, the design was deemed flawed because
+of the hardware requirements: while the Rx and Tx lines are shared between all
+payloads, each still requires two I/O pins on the \acrshort{obc} for the 
+\(\overline{CE}\) and \(\overline{CR}\) lines. With most micro-controllers
+exposing between 5 and 15 I/O pins, this would have strongly limited how many
+payloads could have been connected at the same time.
+
+\subsubsection{\acrshort{i2c} Bus Interface}
 

### 2017-04-07 chapters/02-lit-review.tex
move chapters ot chapters/

--- 2017-04-04 paper.tex
@@
-\input{header.tex}
-\input{glossaries.tex}
-
-\begin{document}
-
-\frontmatter
-
-\begin{titlepage}
-\maketitle
-\end{titlepage}
-
-\begin{abstract}
-\end{abstract}
-
-\tableofcontents
-\setlength{\parskip}{1em}
-
-\listoffigures
-
-\listoftables
-
-\printglossary
-\printglossary[type=\acronymtype]
-
-\mainmatter
-
-\chapter{Introduction}
-\label{ch:introduction}
-
-\newpage
 \chapter{Literature Review}
@@
 \section{Analysis of the Problem}
-
-\newpage
-\chapter{Proposed Solution and Methodology}
-\label{ch:methodology}
-
-% TODO: add something about:
-%       - hardware choices
-%       - software choices
-%       - language choices
-%
-%       - overall arch (flight platform, ground segmetn)
-
-The proposed solution for this project is inspired by the bus model that
-supports the architecture of most commercial satellites. A standard software
-and hardware platform provides common facilities required on all High-Altitude
-flights – position tracking, internal data transfers and radio telemetry to the
-ground station – allowing payloads (scientific instruments for example) to
-only generate data and make it available to the platform.
-
-\begin{figure}[H]
-\includegraphics[width=0.35\textwidth]{payloadbus.pdf}
-\centering
-\caption{Payload bus architecture}
-\end{figure}
-
-The platform, or \textit{Payload Bus} is composed of the following components 
-hardware and software components:
-
-\begin{description}
-
-\item[On Board Computer:] the computer that runs the flight software and
-controls the data bus and all communications units (also called
-\textit{Bus Controller}).
-
-\item[Flight Software:] the firmware installed on the \textit{OBC}, that
-controls the bus and the execution of the mission.
-
-\item[Power Bus:] the hardware interface that provides electrical power to
-payloads.
-
-\item[Data Bus:] the hardware interface and software protocol through which
-payloads communicate data to the flight computer.
-
-\item[Navigation System:] the hardware (GNSS receiver) and software module used
-by the flight software to keep track of the balloon's geographic coordinates and
-altitude.
-
-\item[Radio System:] the hardware (transmitter), protocols and software module
-used by the \textit{OBC} to transmit collected data and metadata to the
-ground station.
-
-\item[Ground Segment:] the software required to record and decode the signal
-received from the platform.
-
-\end{description}
-
-The following chapter describes the design processes used to develop and
-integrate those components into a functioning prototype of the Abertay
-High-Altitude Bus (\textit{AHABus}), and the methods used to test the viability
-and performance of the design.
-
-\section{Radio Communications Protocol}
-
-This section describes the design process that led to the radio communications
-protocol used by the AHABus platform to send payload data and flight metadata
-to the ground station. Its role is to ensure that arbitrary binary data – each
-payload can use their own data format – as stably and reliably as possible. As
-discussed in the following subsections, the design and implementation choice
-were also driven by legal constraints: in the United Kingdom, airborne radio
-transmitters are limited in frequency band (70cm wavelength) and power
-(transmission power inferior to 10mW).
-
-\subsection{Requirements}
-\label{ssec:requirements}
-
-Designing and building a communication protocol that can be relied up was an
-important requirement for the rest of the project. Transmitting as much as
-possible of a mission's scientific data is primordial in case the platform
-is carried too far by winds and cannot be recovered after landing.
-
-From the constraints discussed above, the requirements below were established
-for the radio communications system:
-
-\begin{enumerate}
-\item provide a system to transmit binary data.% DATA LINK
-\item minimise the amount of overhead and bandwidth required.% ALL
-\item allow multiple payload's data to be sent on the same link.% PACKET LAYER
-\item provide stable transmissions on a low-power link.% FRAME LAYER
-\item correct as many transmission-induced data errors as possible.% FRAME LAYER
-\item allow ground crew to track the balloon's location over time.% PACKET LAYER
-\end{enumerate}
-
-As stated in chapter \ref{ch:literature-review}, the United Kingdom's Office
-of Communications restrict what frequencies can be used for airborne
-radio transmissions, and how much power the transmissions can be sent at.
-Those restrictions constrain what existing protocols can be used to transmit
-arbitrary data over radio. This disqualifies existing packet radio protocols
-that would have fulfilled most of the requirements, in use in other countries
-like AX.25.
-% TODO: mention Tx only, no Rx
-
-Most UK-launched High Altitude Balloon missions make use of Radio-Teletype
-(\textit{RTTY}) to send ASCII-encoded text. This was considered insufficient
-for the purpose of a common platform for two main reasons: First, ASCII does
-not handle binary data by default -- only some byte values would be displayed
-as ASCII, while the rest would translate to non-printing characters. This leads
-to data loss, unless the binary data is encoded in an ASCII-compatible format
-like base64, which would require more bandwidth to send a similar amount of
-data. Furthermore, ASCII alone does not provide either a way to multiplex data
-coming from different payloads, or any kind of error checking or error
-correction. This means that such facilities must be built on top of the
-plain-text protocol.
-
-RTTY itself functions in the same way as wired UART serial links. It can provide
-some level of error checking if the transmission of a parity bit is enabled, but
-it does not carry enough information to correct errors. Additionally, RTTY is
-purely a link layer, and does not provide any kind of encapsulation necessary
-for multiplexing data from different payloads.
-
-\subsection{Protocol Specification}
-
-The AHABus radio protocol was designed to match the requirements for the radio
-system as closely as possible. While the specification presented in this section
-is the last version, which was implemented in the AHABus prototype, some parts
-of the protocol evolved or had to be changed during the prototyping process.
-% TODO: Might not be necessary to say that here, given further sections
-
-The protocol was designed in layers, some equivalent to the ones present in the
-OSI model (\cite{Stallings1987}). This approach was chosen because it allows
-some level of modularity – if a layer proves problematic, it can be swapped
-for another with minimal impact on the other layers. Each layer fulfils a
-specific purpose:
-
-% TODO: Add some details about simplicity of the network topology
-
-\begin{description}
-\item[Link Layer:] The lowest-level layer's purpose is to establish a one-way
-binary stream between the two end points of the radio link.
-
-\item[Frame Layer:] The frame layer relies on the Link Layer to transmit its
-data units, and provides reliability and error correction to higher-level 
-layers.
-% TODO: Add vocabulary maybe (ancillary)
-
-\item[Packet Layer:] The packet layer is the highest-level layer. It provides
-a way for multiple payloads' data to share bandwidth as well as metadata
-transmission.
-\end{description}
-
-The design process and specification of each of those three layers is explained
-in more details in the following sections.
-
-\subsubsection{Link Layer}
-\label{sssec:link-layer}
-
-% TODO: add an introductory sentence?
-
-While RTTY was not considered sufficient to cover fulfil the requirements of
-the whole protocol, as shown in section \ref{ssec:requirements}, it fulfils the
-simple requirements of the Link Layer: it can be used to send 8-bit bytes, one
-by one, over a radio link.
-
-RTTY functions like UART, over radio. Instead of using logical electrical levels
-(5V for logical high, 0V for logical low), it uses Audio Frequency Shift Keying
-(\textit{AFSK}): two audio tones separated by a frequency shift act as two
-logical levels. When nothing is being transmitted, the line is held high. The
-start of a byte is signalled by the \textit{start bit} (the line goes low),
-followed by the eight \textit{data bits}, and one or two \textit{stop bits} (the
-line is held high). After a byte has been sent, the line is held high again
-until the next byte, or the next byte starts immediately.
-
-Other existing protocols exist that can transmit data over radio audio signals,
-like DominoEx or Olivia, sometimes with better transmission rates. RTTY was
-preferred over them because the ESP8266 board used as OBC in the AHABus
-prototype provides hardware support for UART, making it easier to implement and
-reliable. Since the link layer is transparent to the frame and packet layers,
-RTTY could be swapped for another protocol in the future with no impact on
-those.
-
-The valid encoding and decoding of a RTTY signal depends on the chosen baud
-rate (the number of times the logical signal changes level). During tests,
-the highest baud rate that allowed clear decoding of radio signals at 10mW
-power rating was 200bauds. Since each byte requires eleven bits to be sent
-(one start bit, eight data bits and two stop bits), this corresponds to a
-\(\frac{200}{11}=18.18\) bytes per second bandwidth.
-
-\subsubsection{Frame Layer}
-\label{sssec:frame-layer}
-
-The role of the frame layer is to provide higher-level layers with a reliable
-data stream: the Link Layer discussed in section \ref{sssec:link-layer} will
-send data bytes one after another, but it does not guarantee that the bytes are
-received, or that noise on the channel (in this case, radio waves) have not
-damaged the data between transmission and reception. Those are the two main
-potential types of issues that the Frame Layer was designed to address.
-
-Preventing complete data loss on a radio channel is only possible to a degree:
-the signal-to-noise ratio is limited by multiple factors including transmission
-power – limited in the United Kingdom – distance, other neighbouring
-transmissions and line-of-sight. Those issues could have been mitigated in part
-through solutions that were outside of this project's scope, like antennae
-design. While this meant that some data loss could not be prevented, a solution
-was required to detect occurrences of data loss, and identify which part of the
-data was missing.
-
-Identifying lost sections of an arbitrary stream of bytes without any form of
-formatting is not possible: since it is not known in advance what data should
-be expected, there is no model to which actual received data can be compared.
-For this reason, the AHABus radio protocol was designed as a framed protocol:
-the stream of data is divided in chunks of a set number of bytes, encapsulated
-in \textit{frames} that contain the a sequence number used to track the position
-of the chunk in the data stream.
-
-The AHABus frame design was inspired by other frame-based protocols like the
-CCSDS Space Packet/Frame protocol. Each frame is composed of two section: a
-header, that contains metadata, and a payload containing the data carried by
-that frame.
-
-The only piece of metadata strictly required to be carried in a frame was its
-sequence number. The frame format was designed to use a 16-bit integer to
-represent the sequence number, as it allowed a large number of frames to be
-sent in-between wrap-arounds of the number (when it becomes larger than 65,535).
-
-To ensure that frames can be identified in the future, even if the format were
-to evolve, the protocol version number was also added to the header format. At
-the time of this writing, the protocol version was 0. If a future frame decoder
-were to receive a frame indicating an old protocol version, it could be at
-least signalled to the user, if not handled gracefully.
-
-The resulting first version of the frame format is presented in figure
-\ref{fig:frame-fmt-orig}.
-
-\begin{figure}[H]
-    \begin{center}
-    \begin{bytefield}[bitwidth=0.5em]{64}
-        \bitheader{0,8,24,40} \\
-        \bitbox{8}{version} &
-        \bitbox{16}{seq. number} &
-        \bitbox{40}{data[253]}
-    \end{bytefield}
-    \end{center}
-    \centering
-    \caption{AHABus Frame Format – Version 1}
-    \label{fig:frame-fmt-orig}
-\end{figure}
-
-This preliminary design proved impossible to work with: while it encapsulates
-the required data and metadata, it does not provide any way to identify the
-start or the end of a frame in a raw binary stream, which would have rendered
-parsing an incoming frame stream impossible.
-
-To solve this issue, a ``start sequence'' was added at the very beginning of
-the frame. When parsing frames from a binary stream, the decoder must ignore
-bytes until the start sequence (a single byte of value \texttt{0xAA} followed
-by a single byte of value \texttt{0x5A}) occurs. Once the frame sequence is
-detected, the frame's end is detected once 256 bytes have been received.
-
-\begin{figure}[H]
-    \begin{center}
-    \begin{bytefield}[bitwidth=0.5em]{64}
-        \bitheader{0,8,16,24,40} \\
-        \bitbox{8}{\texttt{0xAA}} & \bitbox{8}{\texttt{0x5A}} & \bitbox{8}{version} &
-        \bitbox{16}{seq. number} &
-        \bitbox{24}{data[251]}
-    \end{bytefield}
-    \end{center}
-    \centering
-    \caption{AHABus Frame Format – Version 2}
-    \label{fig:frame-fmt-2}
-\end{figure}
-
-The second version of the frame format shown in figure \ref{fig:frame-fmt-2}
-fulfils one of its two design goals (detection of data loss). However, it does
-not provide a way to detect and correct errors due to noise or interferences on
-the radio channel.
-
-The error-correction requirement led to the use of a Forward Error Correction
-(\textit{FEC}) algorithm. FEC algorithms function by adding some amount of data
-\(D\) to each message sent \(M\). In this case, the message includes both the
-frame's header and its payload. \(D\) is computed by the encoder. The decoder
-uses \(D\) to detect whether errors were introduced anywhere in the received
-block \(M+D\), and correct those errors if their number is below the
-algorithm's maximum.
-
-% TODO: block diagram of the encoding-decoding chain
-
-For the purpose of this project, Reed-Solomon(255,223) was chosen for multiple
-reasons:
-
-\begin{itemize}
-
-\item The Reed-Solomon family of algorithms is widespread in its use, notably
-in space communications systems – the Voyager probes used Reed-Solomon codes to
-encode their images (\cite{Wicker1994}), the Space Data Links frames (devised
-by ESA and CCSDS) use a variant of Reed-Solomon to provide FEC
-(\cite{EuropeanCooperationforSpaceStandardization2010}) and the SSDV protocol,
-used in High-Altitude Balloon missions to transmit low-resolution images,
-uses RS(255,223) with 256-byte frames (\cite{UKHAS2016}).
-
-\item There were existing, well tested open-source implementation of that
-specific variant of the Reed-Solomon encoders and decoders, which meant that
-more time could be spent working on other parts of the project.
-
-% TODO: Ask Ian, transparent might not be the right technical term
-
-\item Reed-Solomon codes are transparent: the original data is not modified for
-the purposes of error correction. This means that even if too many errors are
-present for the \acrshort{fec} decoder to correct, the data can be manually
-inspected and some of it can potentially be saved.
-
-\end{itemize}
-
-When applied to 8-bit \textit{symbols} (in this project's case, bytes),
-Reed-Solomon codes can only handle blocks (Data and Error-Correction Codes) of
-a length up to \(2^{8}-1 = 255\) bytes. The chosen variant, \(RS(223,255)\)
-works on 223-byte messages, and adds 32 bytes of Error-Correction Codes at the
-end to produce 255-byte blocks. This prompted slight changes in the frame
-format: each frame being 256-byte long, one byte would not have been covered
-by error correction.
-
-The chosen solution was to not include the start sequence as part of the
-error-corrected message. Since proper detection of that sequence is required
-for the decoder to start parsing a frame, it would have been impossible to apply
-FEC to a frame whose frame start sequence is corrupted.
-
-\begin{figure}[H]
-    \begin{center}
-    \begin{bytefield}[bitwidth=1.8em]{20}
-        \bitbox{1}{...} &
-        \bitbox{8}{frame} &
-        \bitbox{8}{frame} &
-        \bitbox{3}{...}
-    \end{bytefield}
-    \end{center}
-    \centering
-    \caption{Frame stream with embedded start sequence}
-    \label{fig:frame-sync-emb}
-\end{figure}
-
-The start sequence design had to be changed to occupy a single frame byte, as
-opposed to two bytes in version 2 of the frame format. Using a single byte to
-detect the start of a frame was not deemed sufficient, since the probably of
-that exact value occurring elsewhere in the binary stream would have been
-fairly high. To mitigate this, the third and final version of the frame format
-uses a single frame start marker (\texttt{0x5A}) at the beginning of each
-frame, and each frame must be preceded by at least one synchronisation byte
-(\texttt{0xAA}), inserted in the binary stream as shown in the lower half of 
-figure \ref{fig:frame-sync-extra}. The following 223 bytes in the frame
-(starting with the protocol version number) are error-corrected, with the
-frame's 32 trailing bytes.
-
-\begin{figure}[H]
-    \begin{center}
-    \begin{bytefield}[bitwidth=1.8em]{20}
-        \bitbox{1}{...} &
-        \bitbox{1}{\texttt{AA}} &
-        \bitbox{8}{frame} &
-        \bitbox{1}{\texttt{AA}} &
-        \bitbox{8}{frame} &
-        \bitbox{1}{...}
-    \end{bytefield}
-    \end{center}
-    \centering
-    \caption{Frame stream with external synchronisation byte}
-    \label{fig:frame-sync-extra}
-\end{figure}
-
-While this approach may seem wasteful since at least one extra byte must be
-inserted in between each frame, it is to be noted that the same synchronisation
-byte was previously part of each frame's 256 bytes, leading to a smaller amount
-of data being encapsulated in each frame.
-
-\begin{figure}[H]
-    \begin{center}
-    \begin{tikzpicture}[scale=0.2]
-    \tikzstyle{every node}+=[inner sep=0pt]
-    \draw [black] (23.9,-13.4) circle (3);
-    \draw (23.9,-13.4) node {$Wait$};
-    \draw [black] (53.5,-13.4) circle (3);
-    \draw (53.5,-13.4) node {$Sync$};
-    \draw [black] (38.6,-30.7) circle (3);
-    \draw (38.6,-30.7) node {$Frame$};
-    \draw [black] (26.719,-12.375) arc (107.79277:72.20723:39.209);
-    \fill [black] (50.68,-12.37) -- (50.07,-11.65) -- (49.77,-12.61);
-    \draw (38.7,-10) node [above] {$\texttt{byte=0xAA}$};
-    \draw [black] (53.11,-16.371) arc (-12.09889:-69.37588:18.591);
-    \fill [black] (41.48,-29.87) -- (42.41,-30.06) -- (42.05,-29.12);
-    \draw (49.56,-26.05) node [right] {$\texttt{byte=0x5A}$};
-    \draw [black] (35.918,-29.359) arc (-119.94439:-159.34576:25.489);
-    \fill [black] (24.79,-16.26) -- (24.61,-17.19) -- (25.54,-16.84);
-    \draw (28.67,-25.22) node [left] {$\texttt{read=256}$};
-    \draw [black] (51.972,-15.968) arc (-3.02212:-291.02212:2.25);
-    \draw (47.22,-18.39) node [left] {$\texttt{byte=0xAA}$};
-    \fill [black] (50.53,-13.75) -- (49.76,-13.21) -- (49.71,-14.21);
-    \draw [black] (25.724,-11.023) arc (137.60109:42.39891:17.571);
-    \fill [black] (25.72,-11.02) -- (26.63,-10.77) -- (25.89,-10.1);
-    \draw (38.7,-4.8) node [above] {$\texttt{byte!=0xAA}$};
-    \end{tikzpicture}
-    \end{center}
-    \centering
-    \caption{AHABus Frame Synchronisation State Machine}
-    \label{fig:frame-fsm}
-\end{figure}
-
-Figure \ref{fig:frame-fsm} presents the finite state machine that
-\acrshort{ahabus} frame decoders must implement to detect frames in a binary
-stream, and figure \ref{fig:frame-fmt-3} the final version of the
-\acrshort{ahabus} frame format.
-
-% Go to the magic website http://madebyevan.com/fsm/
-
-\begin{figure}[H]
-    \begin{bytefield}{16}
-        \bitheader{0,7,8,15} \\
-        \begin{leftwordgroup}{Header}
-            \bitbox{8}{\texttt{0x5A}} & \bitbox{8}{version} \\
-            \bitbox{16}{sequence number}
-        \end{leftwordgroup} \\
-        \begin{leftwordgroup}{Payload}
-            \wordbox{2}{data[]}
-        \end{leftwordgroup} \\
-        \begin{leftwordgroup}{FEC}
-            \wordbox{2}{checksum[32]}
-        \end{leftwordgroup}
-    \end{bytefield}
-    \centering
-    \caption{\acrshort{ahabus} Frame Format - Version 3}
-    \label{fig:frame-fmt-3}
-\end{figure}
-
-%TODO: what's that worth? shouldn't it be at the end
-
-Encapsulating the transmitted data adds overhead. If \(n\) is the size of a
-chunk in bytes, and each frame adds \(m\) bytes of metadata, then only a
-percentage \(\frac{n}{n+m}\) of the stream contains actual data. For reasons
-explained further, the \acrshort{ahabus} frames were chosen to be 256-byte
-long. As shown previously, each frame also includes.
-
-\subsubsection{Packet Layer}
-
-The Packet Layer relies on the stable data stream provided by the Link and Frame
-Layers, and was designed to allow the multiplexing of multiple payload's data as
-well as the transmission of ancillary data required for the overall flight
-management (platform's GNSS co-ordinates and altitude).
-
-To achieve those goals, the AHABus protocol uses packets similar to frames. Each
-packet encapsulates a chunk of data retrieved from a single payload. To that
-end, packets were designed to be of variable length, to avoid unnecessary
-padding, which would have wasted bandwidth. The data - referred to as the
-packet's \textit{payload} – is preceded by the packet's header, which contains
-information necessary to attribute the packet's data to a payload, indicate
-the length of the packet, and carry the AHABus platform's location and altitude.
-
-This design achieves multiplexing through time division: packets are chunked
-and sent in frames one after the other, and can be dispatched to their
-respective payload teams once decoded at the ground station. The main issue with
-that method is that the protocol itself does not ensure that each payload gets
-an even or fair share of the total bandwidth. As discussed in section
-\ref{sec:flight-software}, this was solved in the flight software design.
-
-\begin{figure}[H]
-    \begin{bytefield}{32}
-        \bitheader{0,7,8,15,16,23,24,31} \\
-        \begin{leftwordgroup}{Header}
-            \bitbox{8}{version} & \bitbox{8}{payload ID} &
-            \bitbox{16}{length} \\
-            \bitbox{32}{latitude} \\
-            \bitbox{32}{longitude} \\
-            \bitbox{16}{altitude} & \bitbox[lrt]{16}{}
-        \end{leftwordgroup} \\
-        \begin{leftwordgroup}{Payload}
-            \wordbox[lrb]{3}{data[length]}
-        \end{leftwordgroup}
-    \end{bytefield}
-    \centering
-    \caption{AHABus Packet Format - Version 1}
-    \label{fig:packet-fmt-1}
-\end{figure}
-
-Figure \ref{fig:packet-fmt-1} shows the first and current version of the AHABus
-packet format. Like frame discussed in section \ref{sssec:frame-layer}, each
-packet carries a version flag which was added to allow backwards compatibility,
-were the protocol to evolve in the future.
-
-The encoding of the metadata was chosen based on the physical constraints of
-the information:
-
-\begin{itemize}
-\item Latitude and Longitudes can be both positive (North, East) or Negative
-(South, West). At the equator, \({0.0001}^\circ\) of longitude maps to roughly
-11 metres\footnotemark , and that distance decreases when moving towards the
-poles, which was deemed precise enough for the project's purpose. To avoid
-conflicts in floating-point mathematics conventions, and allow packet encoders
-to be written for embedded platform without hardware floating point support,
-latitude and longitude were designed to be multiplied by 10,000 and transmitted
-as 32-bit precision signed integers.
-
-\item Altitude is above zero for the whole mission duration, and cannot exceed
-50,000m. Since a sub-metre precision is not required, and would hardly be
-achievable with civilian GNSS receivers, it is encoded as a 16-bit precision
-unsigned integer.
-\end{itemize}
-
-\footnotetext{
-Earth circumference at the equator: \(40\time10^{6}\)m.
-\(D_{0.0001^{\circ}} = 0.0001\frac{40\time10^{6}}{360}=11.1...m\)
-}
-
-The decision was made during the design process to align packet boundaries with
-frame boundaries. This means that every time a new packet is sent through the
-radio channel starts at the start of a new frame, instead of being inserted
-in the middle of a frame, where the last packet stopped.
-
-This approach leads to a loss in bandwidth: if a packet's length is not a 
-multiple of the amount of data that a frame carry, the packet's last frame
-will be padded with zeros instead of containing the start of the next packet.
-However, it was deemed an acceptable tradeoff because of the issues that
-non-aligned (contiguous) packets would present:
-
-\begin{itemize}
-\item If a packet's end does not match a frame's end, that frame cannot be
-sent until the start of the next packet is known. This meant that the
-transmission of most packet's last frames could be delayed for sometimes several
-seconds until another packet is processed.
-
-\item In the case where a frame is lost due to channel noise or interferences,
-finding the start of the next packet would have been complicated, or required
-a synchronisation mechanism similar to the one designed for AHABus frames.
-Aligning packets instead makes such a system unnecessary, reducing the packet
-header size and the complexity of packet decoders.
-\end{itemize}
-
-% TODO: Add diagram explaining aligment
-
-\subsection{Initial Implementation and Testing}
-
-To evaluate the protocol's viability, a reference implementation of an encoder
-and decoder was written. The goal was to
-
-\section{Payload Data Bus}
-
-\subsection{Requirements}
-
-\subsection{Protocol Specification}
-
-\subsection{Arduino Implementation}
-
-\section{Flight Software}
-\label{sec:flight-software}
-
-\subsection{Architecture}
-
-\subsection{User Interface}
-
-\section{Ground Segment}
-
-\subsection{Extracting Raw Binary Stream}
-
-\subsection{Packet}
-
-\section{Testing}
-
-\subsection{Component Testing}
-
-\subsection{Integrated Testing}
-
-\subsubsection{Day in the life Testing}
-
-\subsubsection{Resilience Testing}
-
-\newpage
-\chapter{Results}
-\label{ch:results}
-
-\newpage
-\chapter{Discussion}
-\label{ch:discussion}
-
-\newpage
-\chapter{Conclusion}
-\label{ch:conclusion}
-
-\printbibliography[heading=bibintoc, title=References]
-
-\end{document}

### 2017-04-07 chapters/03-methodology.tex
move chapters ot chapters/

--- 2017-04-06 paper.tex
@@
-\input{header.tex}
-\input{glossaries.tex}
-
-\begin{document}
-
-\frontmatter
-
-\begin{titlepage}
-\maketitle
-\end{titlepage}
-
-\begin{abstract}
-\end{abstract}
-
-\tableofcontents
-\setlength{\parskip}{1em}
-
-\listoffigures
-
-\listoftables
-
-\printglossary
-\printglossary[type=\acronymtype]
-
-\mainmatter
-
-\chapter{Introduction}
-\label{ch:introduction}
-
-\newpage
-\chapter{Literature Review}
-\label{ch:literature-review}
-
-Most \acrshort{hab} missions led by universities, schools or amateur are built 
-around commercially available small computers or micro-controller boards:
-Redland Green School students used a Raspberry Pi (\cite{rpi2014}) running
-python scripts as an \acrfull{obc} for their HAB mission
-(\cite{Hinschelwood2015}), while other projects have been ran on AVR
-micro-controller based boards like the Arduino Mega
-(\cite{AtmelCorporation2015}).
-
-In the United Kingdom, radio licenses do not cover airborne transmission, making
-packet radio protocols like AX.25 unusable for HAB missions 
-(\cite{ukhasradio2016}). However, the Office of Communications allows the use
-of certain \textit{license-exempt} frequencies like the 70cm wavelength band,
-provided that the transmission power does not exceed 10mW (\cite{Ofcom2014}).
-For this reason, most missions launched from the United Kingdom use the
-Radiometrix NTX2 transmitter (\cite{radiometrix2012}). Because of the
-constraints, these missions usually transmit their telemetry as text using the
-\acrfull{rtty} protocol.
-
-While most \acrshort{hab} projects make use of commercial-off-the-shelf
-components, the finished payloads are usually custom, single-purpose designs
-that are little, if at all modular. In the design document for the Titan-1 HAB
-mission, Bombasaro describes a hardware bus based on tightly coupled sensors
-which communicate over two different data buses, SPI and \acrshort{i2c}, and a
-custom-made, single-purpose flight software (\cite{Bombasaro2015}).
-
-Because of the risks and cost involved, nano-satellites rely more on modular
-designs, allowing each team to work on their module independently. In his paper,
-Volstad describes the design of the data bus of the NTNU CubeSat: while the OBC
-is a based on a custom made circuit board, it is designed to provide a standard
-power interface, as well as access to an \acrshort{i2c} standard bus
-(\cite{NXPSemiconductors2014}) to each payload (\cite{Volstad2011}). The
-\acrshort{i2c} protocol was chosen because of its low power consumption, and
-because if only requires two lines (clock and data). Thomas Clausen describes
-in his 2001 paper how a simple packet protocol built on top of \acrshort{i2c}
-itself is used for data transfer and error detection in Aalborg University's
-CubeSats (\cite{Clausen2001}).
-
-Satellite missions last a lot longer than High-Altitude Balloons' and the
-requirements for radio transmissions differ: while satellites can only
-communicate when their orbit passes above a ground station and require precise
-speed and data volume planning, as described by Sandy Anthunes in his book
-(\cite{Antunes2015}), non-floating HAB missions allow for line-of-sight
-communications from liftoff to late into the descent of the payload.
-
-There are however some technologies that can be adapted from satellites to be
-used in \acrshort{hab} applications. The University of Arizona uses a custom
-packet format over 434MHz radio, containing raw binary values for each
-instrument's measurements (\cite{Eatchel2002}). The BRITE-Austria CubeSat
-mission uses the AX.25 packet radio protocol (used by amateur radio users),
-which allows the use of off-the-shelf transmission and reception hardware
-rather than custom-made circuits (\cite{Traussnig2007}).
-
-Some nano-satellites use custom telemetry data format, which allows the team
-to minimise the volume of data to be sent over radio. The Planetary Society's
-Lightsail mission has an uplink connection that allows the ground station
-to request specific logs or data. By default, The spacecraft only communicates
-a custom beacon containing a summary of the its housekeeping data
-(\cite{planetary2016}). Uplink being impractical with the transmission power
-limits imposed in the United Kingdom, such a selective telemetry system cannot
-be relied upon for \acrshort{hab} missions.
-
-To improve collaboration and allow the deployment of large networks of
-satellites, probes and other space vehicles, the Consultative Committee for
-Space Data Systems has defined multiple communication standards that for
-the different types of networks encountered in spacecrafts.
-
-SpaceWire defines how instruments' data can be accessed by having the OBC
-remotely poll their memory (\cite{parkes2005}). A similar system is used to
-fetch data from ROM chips over \acrshort{i2c} and could be used to build a
-simple data bus.
-
-The \acrshort{ccsds} Space Packet protocol defines a packet format used to
-encapsulate application data (from instruments for example) that can be sent
-over a ``Space Link'', an analogue of the OSI model's link layer
-(\cite{Stallings1987}) using \acrshort{ccsds} frames to encapsulate packets
-originating from multiple instruments, ground stations and spacecrafts
-(\cite{ccsds2003}).
-
-While \acrshort{hab} applications do not require the level of complexity of
-\acrshort{ccsds} protocols — addressing is not needed since the network only
-contains two endpoints, communicating in a single direction — some details of
-the Space Packet protocol are worth reusing (Forward Error Correction,
-different Application IDs for each payload).
-
-\section{Analysis of the Problem}
-
-\newpage
 \chapter{Proposed Solution and Methodology}
@@
 \centering
-\caption{Payload bus architecture}
+\caption{Platform architecture}
 \end{figure}
@@
 
-\section{Radio Communications Protocol}
+\section{Radio Communications}
 
@@
 for multiplexing data from different payloads.
+
+\subsection{Hardware Setup}
 
@@
 
+The second experimental setup was much simpler: instead of using an hybrid
+solution like the one described in the previous section, it relied on the
+\acrshort{i2c} standard. The protocol was chosen because it provides a two-way
+communication interface over two wires, for up to 127 devices. The major
+difference with a \acrshort{uart} serial interface is that \acrshort{i2c} is
+asymmetric: one device acts as the bus controllers, and passive devices can
+only receive or transmit data when addressed by it.
+
 \subsection{Protocol Specification}
@@
 \subsubsection{Resilience Testing}
-
-\newpage
-\chapter{Results}
-\label{ch:results}
-
-\newpage
-\chapter{Discussion}
-\label{ch:discussion}
-
-\newpage
-\chapter{Conclusion}
-\label{ch:conclusion}
-
-\printbibliography[heading=bibintoc, title=References]
-
-\end{document}

### 2017-04-07 chapters/04-results.tex
move chapters ot chapters/

\chapter{Results}
\label{ch:results}

\section{Component Testing}

\section{Day In The Life Testing}

The Day-in-the-Life test consisted of a full simulated \acrlong{hab} mission,
from lift-off to landing. To obtain data similar to what would be encountered
during a real flight, the GNSS receiver was replaced by an Arduino board 
sending simulated geographic co-ordinates and altitude.

Figure \ref{fig:dit-alt} shows the altitude of the AHABus platform reported by
the received headers over the approximately two hours and twenty minutes of the
simulation. The altitude marks are matched to the time at which each packet
is received, and not emitted.

\begin{figure}[H]
\begin{tikzpicture}
\begin{axis}
\addplot table [
xlabel=Mission Elapsed Time,
ylabel=Altitude,
x=met, y=alt, col sep=comma
] {data/02-ditl/loc.csv};
\end{axis}
\end{tikzpicture}
\centering
\caption{DITL Test - Altitude over time}
\label{fig:dit-alt}
\end{figure}

The mission platform was composed of the AHABus \acrlong{obc}, and two simulated
payloads: payload 10 at priority level 2, and payload 20 at priority level 1.
Table \ref{tab:ditl-cfg}.

\begin{table}[H]
\caption {DITL Test - FCORE configuration}
\label{tab:ditl-cfg}
\begin{center}
\begin{tabular}{ |c|c c c| }
\hline
payload & priority & core frequency & polling interval \\
\hline
GNSS & - & -
10 & 2 & \SI{5.082}{\hertz} & \SI{102}{\second} \\
20 & 1 & \SI{10.164}{\hertz} & \SI{51}{\second} \\
\hline 
\end{tabular}
\end{center}
\end{table}

\section{Radio Stress-Testing}

### 2017-04-07 paper.tex
move chapters ot chapters/

--- 2017-04-04 paper.tex
@@
 \mainmatter
-
-\chapter{Introduction}
-\label{ch:introduction}
+\input{chapters/01-introduction.tex}
 
 \newpage
-\chapter{Literature Review}
-\label{ch:literature-review}
-
-Most \acrshort{hab} missions led by universities, schools or amateur are built 
-around commercially available small computers or micro-controller boards:
-Redland Green School students used a Raspberry Pi (\cite{rpi2014}) running
-python scripts as an \acrfull{obc} for their HAB mission
-(\cite{Hinschelwood2015}), while other projects have been ran on AVR
-micro-controller based boards like the Arduino Mega
-(\cite{AtmelCorporation2015}).
-
-In the United Kingdom, radio licenses do not cover airborne transmission, making
-packet radio protocols like AX.25 unusable for HAB missions 
-(\cite{ukhasradio2016}). However, the Office of Communications allows the use
-of certain \textit{license-exempt} frequencies like the 70cm wavelength band,
-provided that the transmission power does not exceed 10mW (\cite{Ofcom2014}).
-For this reason, most missions launched from the United Kingdom use the
-Radiometrix NTX2 transmitter (\cite{radiometrix2012}). Because of the
-constraints, these missions usually transmit their telemetry as text using the
-\acrfull{rtty} protocol.
-
-While most \acrshort{hab} projects make use of commercial-off-the-shelf
-components, the finished payloads are usually custom, single-purpose designs
-that are little, if at all modular. In the design document for the Titan-1 HAB
-mission, Bombasaro describes a hardware bus based on tightly coupled sensors
-which communicate over two different data buses, SPI and \acrshort{i2c}, and a
-custom-made, single-purpose flight software (\cite{Bombasaro2015}).
-
-Because of the risks and cost involved, nano-satellites rely more on modular
-designs, allowing each team to work on their module independently. In his paper,
-Volstad describes the design of the data bus of the NTNU CubeSat: while the OBC
-is a based on a custom made circuit board, it is designed to provide a standard
-power interface, as well as access to an \acrshort{i2c} standard bus
-(\cite{NXPSemiconductors2014}) to each payload (\cite{Volstad2011}). The
-\acrshort{i2c} protocol was chosen because of its low power consumption, and
-because if only requires two lines (clock and data). Thomas Clausen describes
-in his 2001 paper how a simple packet protocol built on top of \acrshort{i2c}
-itself is used for data transfer and error detection in Aalborg University's
-CubeSats (\cite{Clausen2001}).
-
-Satellite missions last a lot longer than High-Altitude Balloons' and the
-requirements for radio transmissions differ: while satellites can only
-communicate when their orbit passes above a ground station and require precise
-speed and data volume planning, as described by Sandy Anthunes in his book
-(\cite{Antunes2015}), non-floating HAB missions allow for line-of-sight
-communications from liftoff to late into the descent of the payload.
-
-There are however some technologies that can be adapted from satellites to be
-used in \acrshort{hab} applications. The University of Arizona uses a custom
-packet format over 434MHz radio, containing raw binary values for each
-instrument's measurements (\cite{Eatchel2002}). The BRITE-Austria CubeSat
-mission uses the AX.25 packet radio protocol (used by amateur radio users),
-which allows the use of off-the-shelf transmission and reception hardware
-rather than custom-made circuits (\cite{Traussnig2007}).
-
-Some nano-satellites use custom telemetry data format, which allows the team
-to minimise the volume of data to be sent over radio. The Planetary Society's
-Lightsail mission has an uplink connection that allows the ground station
-to request specific logs or data. By default, The spacecraft only communicates
-a custom beacon containing a summary of the its housekeeping data
-(\cite{planetary2016}). Uplink being impractical with the transmission power
-limits imposed in the United Kingdom, such a selective telemetry system cannot
-be relied upon for \acrshort{hab} missions.
-
-To improve collaboration and allow the deployment of large networks of
-satellites, probes and other space vehicles, the Consultative Committee for
-Space Data Systems has defined multiple communication standards that for
-the different types of networks encountered in spacecrafts.
-
-SpaceWire defines how instruments' data can be accessed by having the OBC
-remotely poll their memory (\cite{parkes2005}). A similar system is used to
-fetch data from ROM chips over \acrshort{i2c} and could be used to build a
-simple data bus.
-
-The \acrshort{ccsds} Space Packet protocol defines a packet format used to
-encapsulate application data (from instruments for example) that can be sent
-over a ``Space Link'', an analogue of the OSI model's link layer
-(\cite{Stallings1987}) using \acrshort{ccsds} frames to encapsulate packets
-originating from multiple instruments, ground stations and spacecrafts
-(\cite{ccsds2003}).
-
-While \acrshort{hab} applications do not require the level of complexity of
-\acrshort{ccsds} protocols — addressing is not needed since the network only
-contains two endpoints, communicating in a single direction — some details of
-the Space Packet protocol are worth reusing (Forward Error Correction,
-different Application IDs for each payload).
-
-\section{Analysis of the Problem}
+\input{chapters/02-lit-review.tex}
 
 \newpage
-\chapter{Proposed Solution and Methodology}
-\label{ch:methodology}
-
-% TODO: add something about:
-%       - hardware choices
-%       - software choices
-%       - language choices
-%
-%       - overall arch (flight platform, ground segmetn)
-
-The proposed solution for this project is inspired by the bus model that
-supports the architecture of most commercial satellites. A standard software
-and hardware platform provides common facilities required on all High-Altitude
-flights – position tracking, internal data transfers and radio telemetry to the
-ground station – allowing payloads (scientific instruments for example) to
-only generate data and make it available to the platform.
-
-\begin{figure}[H]
-\includegraphics[width=0.35\textwidth]{payloadbus.pdf}
-\centering
-\caption{Payload bus architecture}
-\end{figure}
-
-The platform, or \textit{Payload Bus} is composed of the following components 
-hardware and software components:
-
-\begin{description}
-
-\item[On Board Computer:] the computer that runs the flight software and
-controls the data bus and all communications units (also called
-\textit{Bus Controller}).
-
-\item[Flight Software:] the firmware installed on the \textit{OBC}, that
-controls the bus and the execution of the mission.
-
-\item[Power Bus:] the hardware interface that provides electrical power to
-payloads.
-
-\item[Data Bus:] the hardware interface and software protocol through which
-payloads communicate data to the flight computer.
-
-\item[Navigation System:] the hardware (GNSS receiver) and software module used
-by the flight software to keep track of the balloon's geographic coordinates and
-altitude.
-
-\item[Radio System:] the hardware (transmitter), protocols and software module
-used by the \textit{OBC} to transmit collected data and metadata to the
-ground station.
-
-\item[Ground Segment:] the software required to record and decode the signal
-received from the platform.
-
-\end{description}
-
-The following chapter describes the design processes used to develop and
-integrate those components into a functioning prototype of the Abertay
-High-Altitude Bus (\textit{AHABus}), and the methods used to test the viability
-and performance of the design.
-
-\section{Radio Communications Protocol}
-
-This section describes the design process that led to the radio communications
-protocol used by the AHABus platform to send payload data and flight metadata
-to the ground station. Its role is to ensure that arbitrary binary data – each
-payload can use their own data format – as stably and reliably as possible. As
-discussed in the following subsections, the design and implementation choice
-were also driven by legal constraints: in the United Kingdom, airborne radio
-transmitters are limited in frequency band (70cm wavelength) and power
-(transmission power inferior to 10mW).
-
-\subsection{Requirements}
-\label{ssec:requirements}
-
-Designing and building a communication protocol that can be relied up was an
-important requirement for the rest of the project. Transmitting as much as
-possible of a mission's scientific data is primordial in case the platform
-is carried too far by winds and cannot be recovered after landing.
-
-From the constraints discussed above, the requirements below were established
-for the radio communications system:
-
-\begin{enumerate}
-\item provide a system to transmit binary data.% DATA LINK
-\item minimise the amount of overhead and bandwidth required.% ALL
-\item allow multiple payload's data to be sent on the same link.% PACKET LAYER
-\item provide stable transmissions on a low-power link.% FRAME LAYER
-\item correct as many transmission-induced data errors as possible.% FRAME LAYER
-\item allow ground crew to track the balloon's location over time.% PACKET LAYER
-\end{enumerate}
-
-As stated in chapter \ref{ch:literature-review}, the United Kingdom's Office
-of Communications restrict what frequencies can be used for airborne
-radio transmissions, and how much power the transmissions can be sent at.
-Those restrictions constrain what existing protocols can be used to transmit
-arbitrary data over radio. This disqualifies existing packet radio protocols
-that would have fulfilled most of the requirements, in use in other countries
-like AX.25.
-% TODO: mention Tx only, no Rx
-
-Most UK-launched High Altitude Balloon missions make use of Radio-Teletype
-(\textit{RTTY}) to send ASCII-encoded text. This was considered insufficient
-for the purpose of a common platform for two main reasons: First, ASCII does
-not handle binary data by default -- only some byte values would be displayed
-as ASCII, while the rest would translate to non-printing characters. This leads
-to data loss, unless the binary data is encoded in an ASCII-compatible format
-like base64, which would require more bandwidth to send a similar amount of
-data. Furthermore, ASCII alone does not provide either a way to multiplex data
-coming from different payloads, or any kind of error checking or error
-correction. This means that such facilities must be built on top of the
-plain-text protocol.
-
-RTTY itself functions in the same way as wired UART serial links. It can provide
-some level of error checking if the transmission of a parity bit is enabled, but
-it does not carry enough information to correct errors. Additionally, RTTY is
-purely a link layer, and does not provide any kind of encapsulation necessary
-for multiplexing data from different payloads.
-
-\subsection{Protocol Specification}
-
-The AHABus radio protocol was designed to match the requirements for the radio
-system as closely as possible. While the specification presented in this section
-is the last version, which was implemented in the AHABus prototype, some parts
-of the protocol evolved or had to be changed during the prototyping process.
-% TODO: Might not be necessary to say that here, given further sections
-
-The protocol was designed in layers, some equivalent to the ones present in the
-OSI model (\cite{Stallings1987}). This approach was chosen because it allows
-some level of modularity – if a layer proves problematic, it can be swapped
-for another with minimal impact on the other layers. Each layer fulfils a
-specific purpose:
-
-% TODO: Add some details about simplicity of the network topology
-
-\begin{description}
-\item[Link Layer:] The lowest-level layer's purpose is to establish a one-way
-binary stream between the two end points of the radio link.
-
-\item[Frame Layer:] The frame layer relies on the Link Layer to transmit its
-data units, and provides reliability and error correction to higher-level 
-layers.
-% TODO: Add vocabulary maybe (ancillary)
-
-\item[Packet Layer:] The packet layer is the highest-level layer. It provides
-a way for multiple payloads' data to share bandwidth as well as metadata
-transmission.
-\end{description}
-
-The design process and specification of each of those three layers is explained
-in more details in the following sections.
-
-\subsubsection{Link Layer}
-\label{sssec:link-layer}
-
-% TODO: add an introductory sentence?
-
-While RTTY was not considered sufficient to cover fulfil the requirements of
-the whole protocol, as shown in section \ref{ssec:requirements}, it fulfils the
-simple requirements of the Link Layer: it can be used to send 8-bit bytes, one
-by one, over a radio link.
-
-RTTY functions like UART, over radio. Instead of using logical electrical levels
-(5V for logical high, 0V for logical low), it uses Audio Frequency Shift Keying
-(\textit{AFSK}): two audio tones separated by a frequency shift act as two
-logical levels. When nothing is being transmitted, the line is held high. The
-start of a byte is signalled by the \textit{start bit} (the line goes low),
-followed by the eight \textit{data bits}, and one or two \textit{stop bits} (the
-line is held high). After a byte has been sent, the line is held high again
-until the next byte, or the next byte starts immediately.
-
-Other existing protocols exist that can transmit data over radio audio signals,
-like DominoEx or Olivia, sometimes with better transmission rates. RTTY was
-preferred over them because the ESP8266 board used as OBC in the AHABus
-prototype provides hardware support for UART, making it easier to implement and
-reliable. Since the link layer is transparent to the frame and packet layers,
-RTTY could be swapped for another protocol in the future with no impact on
-those.
-
-The valid encoding and decoding of a RTTY signal depends on the chosen baud
-rate (the number of times the logical signal changes level). During tests,
-the highest baud rate that allowed clear decoding of radio signals at 10mW
-power rating was 200bauds. Since each byte requires eleven bits to be sent
-(one start bit, eight data bits and two stop bits), this corresponds to a
-\(\frac{200}{11}=18.18\) bytes per second bandwidth.
-
-\subsubsection{Frame Layer}
-\label{sssec:frame-layer}
-
-The role of the frame layer is to provide higher-level layers with a reliable
-data stream: the Link Layer discussed in section \ref{sssec:link-layer} will
-send data bytes one after another, but it does not guarantee that the bytes are
-received, or that noise on the channel (in this case, radio waves) have not
-damaged the data between transmission and reception. Those are the two main
-potential types of issues that the Frame Layer was designed to address.
-
-Preventing complete data loss on a radio channel is only possible to a degree:
-the signal-to-noise ratio is limited by multiple factors including transmission
-power – limited in the United Kingdom – distance, other neighbouring
-transmissions and line-of-sight. Those issues could have been mitigated in part
-through solutions that were outside of this project's scope, like antennae
-design. While this meant that some data loss could not be prevented, a solution
-was required to detect occurrences of data loss, and identify which part of the
-data was missing.
-
-Identifying lost sections of an arbitrary stream of bytes without any form of
-formatting is not possible: since it is not known in advance what data should
-be expected, there is no model to which actual received data can be compared.
-For this reason, the AHABus radio protocol was designed as a framed protocol:
-the stream of data is divided in chunks of a set number of bytes, encapsulated
-in \textit{frames} that contain the a sequence number used to track the position
-of the chunk in the data stream.
-
-The AHABus frame design was inspired by other frame-based protocols like the
-CCSDS Space Packet/Frame protocol. Each frame is composed of two section: a
-header, that contains metadata, and a payload containing the data carried by
-that frame.
-
-The only piece of metadata strictly required to be carried in a frame was its
-sequence number. The frame format was designed to use a 16-bit integer to
-represent the sequence number, as it allowed a large number of frames to be
-sent in-between wrap-arounds of the number (when it becomes larger than 65,535).
-
-To ensure that frames can be identified in the future, even if the format were
-to evolve, the protocol version number was also added to the header format. At
-the time of this writing, the protocol version was 0. If a future frame decoder
-were to receive a frame indicating an old protocol version, it could be at
-least signalled to the user, if not handled gracefully.
-
-The resulting first version of the frame format is presented in figure
-\ref{fig:frame-fmt-orig}.
-
-\begin{figure}[H]
-    \begin{center}
-    \begin{bytefield}[bitwidth=0.5em]{64}
-        \bitheader{0,8,24,40} \\
-        \bitbox{8}{version} &
-        \bitbox{16}{seq. number} &
-        \bitbox{40}{data[253]}
-    \end{bytefield}
-    \end{center}
-    \centering
-    \caption{AHABus Frame Format – Version 1}
-    \label{fig:frame-fmt-orig}
-\end{figure}
-
-This preliminary design proved impossible to work with: while it encapsulates
-the required data and metadata, it does not provide any way to identify the
-start or the end of a frame in a raw binary stream, which would have rendered
-parsing an incoming frame stream impossible.
-
-To solve this issue, a ``start sequence'' was added at the very beginning of
-the frame. When parsing frames from a binary stream, the decoder must ignore
-bytes until the start sequence (a single byte of value \texttt{0xAA} followed
-by a single byte of value \texttt{0x5A}) occurs. Once the frame sequence is
-detected, the frame's end is detected once 256 bytes have been received.
-
-\begin{figure}[H]
-    \begin{center}
-    \begin{bytefield}[bitwidth=0.5em]{64}
-        \bitheader{0,8,16,24,40} \\
-        \bitbox{8}{\texttt{0xAA}} & \bitbox{8}{\texttt{0x5A}} & \bitbox{8}{version} &
-        \bitbox{16}{seq. number} &
-        \bitbox{24}{data[251]}
-    \end{bytefield}
-    \end{center}
-    \centering
-    \caption{AHABus Frame Format – Version 2}
-    \label{fig:frame-fmt-2}
-\end{figure}
-
-The second version of the frame format shown in figure \ref{fig:frame-fmt-2}
-fulfils one of its two design goals (detection of data loss). However, it does
-not provide a way to detect and correct errors due to noise or interferences on
-the radio channel.
-
-The error-correction requirement led to the use of a Forward Error Correction
-(\textit{FEC}) algorithm. FEC algorithms function by adding some amount of data
-\(D\) to each message sent \(M\). In this case, the message includes both the
-frame's header and its payload. \(D\) is computed by the encoder. The decoder
-uses \(D\) to detect whether errors were introduced anywhere in the received
-block \(M+D\), and correct those errors if their number is below the
-algorithm's maximum.
-
-% TODO: block diagram of the encoding-decoding chain
-
-For the purpose of this project, Reed-Solomon(255,223) was chosen for multiple
-reasons:
-
-\begin{itemize}
-
-\item The Reed-Solomon family of algorithms is widespread in its use, notably
-in space communications systems – the Voyager probes used Reed-Solomon codes to
-encode their images (\cite{Wicker1994}), the Space Data Links frames (devised
-by ESA and CCSDS) use a variant of Reed-Solomon to provide FEC
-(\cite{EuropeanCooperationforSpaceStandardization2010}) and the SSDV protocol,
-used in High-Altitude Balloon missions to transmit low-resolution images,
-uses RS(255,223) with 256-byte frames (\cite{UKHAS2016}).
-
-\item There were existing, well tested open-source implementation of that
-specific variant of the Reed-Solomon encoders and decoders, which meant that
-more time could be spent working on other parts of the project.
-
-% TODO: Ask Ian, transparent might not be the right technical term
-
-\item Reed-Solomon codes are transparent: the original data is not modified for
-the purposes of error correction. This means that even if too many errors are
-present for the \acrshort{fec} decoder to correct, the data can be manually
-inspected and some of it can potentially be saved.
-
-\end{itemize}
-
-When applied to 8-bit \textit{symbols} (in this project's case, bytes),
-Reed-Solomon codes can only handle blocks (Data and Error-Correction Codes) of
-a length up to \(2^{8}-1 = 255\) bytes. The chosen variant, \(RS(223,255)\)
-works on 223-byte messages, and adds 32 bytes of Error-Correction Codes at the
-end to produce 255-byte blocks. This prompted slight changes in the frame
-format: each frame being 256-byte long, one byte would not have been covered
-by error correction.
-
-The chosen solution was to not include the start sequence as part of the
-error-corrected message. Since proper detection of that sequence is required
-for the decoder to start parsing a frame, it would have been impossible to apply
-FEC to a frame whose frame start sequence is corrupted.
-
-\begin{figure}[H]
-    \begin{center}
-    \begin{bytefield}[bitwidth=1.8em]{20}
-        \bitbox{1}{...} &
-        \bitbox{8}{frame} &
-        \bitbox{8}{frame} &
-        \bitbox{3}{...}
-    \end{bytefield}
-    \end{center}
-    \centering
-    \caption{Frame stream with embedded start sequence}
-    \label{fig:frame-sync-emb}
-\end{figure}
-
-The start sequence design had to be changed to occupy a single frame byte, as
-opposed to two bytes in version 2 of the frame format. Using a single byte to
-detect the start of a frame was not deemed sufficient, since the probably of
-that exact value occurring elsewhere in the binary stream would have been
-fairly high. To mitigate this, the third and final version of the frame format
-uses a single frame start marker (\texttt{0x5A}) at the beginning of each
-frame, and each frame must be preceded by at least one synchronisation byte
-(\texttt{0xAA}), inserted in the binary stream as shown in the lower half of 
-figure \ref{fig:frame-sync-extra}. The following 223 bytes in the frame
-(starting with the protocol version number) are error-corrected, with the
-frame's 32 trailing bytes.
-
-\begin{figure}[H]
-    \begin{center}
-    \begin{bytefield}[bitwidth=1.8em]{20}
-        \bitbox{1}{...} &
-        \bitbox{1}{\texttt{AA}} &
-        \bitbox{8}{frame} &
-        \bitbox{1}{\texttt{AA}} &
-        \bitbox{8}{frame} &
-        \bitbox{1}{...}
-    \end{bytefield}
-    \end{center}
-    \centering
-    \caption{Frame stream with external synchronisation byte}
-    \label{fig:frame-sync-extra}
-\end{figure}
-
-While this approach may seem wasteful since at least one extra byte must be
-inserted in between each frame, it is to be noted that the same synchronisation
-byte was previously part of each frame's 256 bytes, leading to a smaller amount
-of data being encapsulated in each frame.
-
-\begin{figure}[H]
-    \begin{center}
-    \begin{tikzpicture}[scale=0.2]
-    \tikzstyle{every node}+=[inner sep=0pt]
-    \draw [black] (23.9,-13.4) circle (3);
-    \draw (23.9,-13.4) node {$Wait$};
-    \draw [black] (53.5,-13.4) circle (3);
-    \draw (53.5,-13.4) node {$Sync$};
-    \draw [black] (38.6,-30.7) circle (3);
-    \draw (38.6,-30.7) node {$Frame$};
-    \draw [black] (26.719,-12.375) arc (107.79277:72.20723:39.209);
-    \fill [black] (50.68,-12.37) -- (50.07,-11.65) -- (49.77,-12.61);
-    \draw (38.7,-10) node [above] {$\texttt{byte=0xAA}$};
-    \draw [black] (53.11,-16.371) arc (-12.09889:-69.37588:18.591);
-    \fill [black] (41.48,-29.87) -- (42.41,-30.06) -- (42.05,-29.12);
-    \draw (49.56,-26.05) node [right] {$\texttt{byte=0x5A}$};
-    \draw [black] (35.918,-29.359) arc (-119.94439:-159.34576:25.489);
-    \fill [black] (24.79,-16.26) -- (24.61,-17.19) -- (25.54,-16.84);
-    \draw (28.67,-25.22) node [left] {$\texttt{read=256}$};
-    \draw [black] (51.972,-15.968) arc (-3.02212:-291.02212:2.25);
-    \draw (47.22,-18.39) node [left] {$\texttt{byte=0xAA}$};
-    \fill [black] (50.53,-13.75) -- (49.76,-13.21) -- (49.71,-14.21);
-    \draw [black] (25.724,-11.023) arc (137.60109:42.39891:17.571);
-    \fill [black] (25.72,-11.02) -- (26.63,-10.77) -- (25.89,-10.1);
-    \draw (38.7,-4.8) node [above] {$\texttt{byte!=0xAA}$};
-    \end{tikzpicture}
-    \end{center}
-    \centering
-    \caption{AHABus Frame Synchronisation State Machine}
-    \label{fig:frame-fsm}
-\end{figure}
-
-Figure \ref{fig:frame-fsm} presents the finite state machine that
-\acrshort{ahabus} frame decoders must implement to detect frames in a binary
-stream, and figure \ref{fig:frame-fmt-3} the final version of the
-\acrshort{ahabus} frame format.
-
-% Go to the magic website http://madebyevan.com/fsm/
-
-\begin{figure}[H]
-    \begin{bytefield}{16}
-        \bitheader{0,7,8,15} \\
-        \begin{leftwordgroup}{Header}
-            \bitbox{8}{\texttt{0x5A}} & \bitbox{8}{version} \\
-            \bitbox{16}{sequence number}
-        \end{leftwordgroup} \\
-        \begin{leftwordgroup}{Payload}
-            \wordbox{2}{data[]}
-        \end{leftwordgroup} \\
-        \begin{leftwordgroup}{FEC}
-            \wordbox{2}{checksum[32]}
-        \end{leftwordgroup}
-    \end{bytefield}
-    \centering
-    \caption{\acrshort{ahabus} Frame Format - Version 3}
-    \label{fig:frame-fmt-3}
-\end{figure}
-
-%TODO: what's that worth? shouldn't it be at the end
-
-Encapsulating the transmitted data adds overhead. If \(n\) is the size of a
-chunk in bytes, and each frame adds \(m\) bytes of metadata, then only a
-percentage \(\frac{n}{n+m}\) of the stream contains actual data. For reasons
-explained further, the \acrshort{ahabus} frames were chosen to be 256-byte
-long. As shown previously, each frame also includes.
-
-\subsubsection{Packet Layer}
-
-The Packet Layer relies on the stable data stream provided by the Link and Frame
-Layers, and was designed to allow the multiplexing of multiple payload's data as
-well as the transmission of ancillary data required for the overall flight
-management (platform's GNSS co-ordinates and altitude).
-
-To achieve those goals, the AHABus protocol uses packets similar to frames. Each
-packet encapsulates a chunk of data retrieved from a single payload. To that
-end, packets were designed to be of variable length, to avoid unnecessary
-padding, which would have wasted bandwidth. The data - referred to as the
-packet's \textit{payload} – is preceded by the packet's header, which contains
-information necessary to attribute the packet's data to a payload, indicate
-the length of the packet, and carry the AHABus platform's location and altitude.
-
-This design achieves multiplexing through time division: packets are chunked
-and sent in frames one after the other, and can be dispatched to their
-respective payload teams once decoded at the ground station. The main issue with
-that method is that the protocol itself does not ensure that each payload gets
-an even or fair share of the total bandwidth. As discussed in section
-\ref{sec:flight-software}, this was solved in the flight software design.
-
-\begin{figure}[H]
-    \begin{bytefield}{32}
-        \bitheader{0,7,8,15,16,23,24,31} \\
-        \begin{leftwordgroup}{Header}
-            \bitbox{8}{version} & \bitbox{8}{payload ID} &
-            \bitbox{16}{length} \\
-            \bitbox{32}{latitude} \\
-            \bitbox{32}{longitude} \\
-            \bitbox{16}{altitude} & \bitbox[lrt]{16}{}
-        \end{leftwordgroup} \\
-        \begin{leftwordgroup}{Payload}
-            \wordbox[lrb]{3}{data[length]}
-        \end{leftwordgroup}
-    \end{bytefield}
-    \centering
-    \caption{AHABus Packet Format - Version 1}
-    \label{fig:packet-fmt-1}
-\end{figure}
-
-Figure \ref{fig:packet-fmt-1} shows the first and current version of the AHABus
-packet format. Like frame discussed in section \ref{sssec:frame-layer}, each
-packet carries a version flag which was added to allow backwards compatibility,
-were the protocol to evolve in the future.
-
-The encoding of the metadata was chosen based on the physical constraints of
-the information:
-
-\begin{itemize}
-\item Latitude and Longitudes can be both positive (North, East) or Negative
-(South, West). At the equator, \({0.0001}^\circ\) of longitude maps to roughly
-11 metres\footnotemark , and that distance decreases when moving towards the
-poles, which was deemed precise enough for the project's purpose. To avoid
-conflicts in floating-point mathematics conventions, and allow packet encoders
-to be written for embedded platform without hardware floating point support,
-latitude and longitude were designed to be multiplied by 10,000 and transmitted
-as 32-bit precision signed integers.
-
-\item Altitude is above zero for the whole mission duration, and cannot exceed
-50,000m. Since a sub-metre precision is not required, and would hardly be
-achievable with civilian GNSS receivers, it is encoded as a 16-bit precision
-unsigned integer.
-\end{itemize}
-
-\footnotetext{
-Earth circumference at the equator: \(40\time10^{6}\)m.
-\(D_{0.0001^{\circ}} = 0.0001\frac{40\time10^{6}}{360}=11.1...m\)
-}
-
-The decision was made during the design process to align packet boundaries with
-frame boundaries. This means that every time a new packet is sent through the
-radio channel starts at the start of a new frame, instead of being inserted
-in the middle of a frame, where the last packet stopped.
-
-This approach leads to a loss in bandwidth: if a packet's length is not a 
-multiple of the amount of data that a frame carry, the packet's last frame
-will be padded with zeros instead of containing the start of the next packet.
-However, it was deemed an acceptable tradeoff because of the issues that
-non-aligned (contiguous) packets would present:
-
-\begin{itemize}
-\item If a packet's end does not match a frame's end, that frame cannot be
-sent until the start of the next packet is known. This meant that the
-transmission of most packet's last frames could be delayed for sometimes several
-seconds until another packet is processed.
-
-\item In the case where a frame is lost due to channel noise or interferences,
-finding the start of the next packet would have been complicated, or required
-a synchronisation mechanism similar to the one designed for AHABus frames.
-Aligning packets instead makes such a system unnecessary, reducing the packet
-header size and the complexity of packet decoders.
-\end{itemize}
-
-% TODO: Add diagram explaining aligment
-
-\subsection{Initial Implementation and Testing}
-
-To evaluate the protocol's viability, a reference implementation of an encoder
-and decoder was written. The goal was to
-
-\section{Payload Data Bus}
-
-\subsection{Requirements}
-
-\subsection{Protocol Specification}
-
-\subsection{Arduino Implementation}
-
-\section{Flight Software}
-\label{sec:flight-software}
-
-\subsection{Architecture}
-
-\subsection{User Interface}
-
-\section{Ground Segment}
-
-\subsection{Extracting Raw Binary Stream}
-
-\subsection{Packet}
-
-\section{Testing}
-
-\subsection{Component Testing}
-
-\subsection{Integrated Testing}
-
-\subsubsection{Day in the life Testing}
-
-\subsubsection{Resilience Testing}
+\input{chapters/03-methodology.tex}
 
 \newpage
-\chapter{Results}
-\label{ch:results}
+\input{chapters/04-results.tex}
 
 \newpage
-\chapter{Discussion}
-\label{ch:discussion}
+\input{chapters/05-discussion.tex}
 
 \newpage
-\chapter{Conclusion}
-\label{ch:conclusion}
+\input{chapters/06-conclusion.tex}
 

### 2017-04-07 data/02-ditl/cfg.c

@@ -1,0 +1,8 @@
+// FCORE Configuration File - FCORE Test 1
+// Generated on 2017-03-29 by fcore_config.rb
+#include "fcore.h"
+
+const FCPayload fcore_payloads[FCORE_PAYLOAD_COUNT] = {
+    {10, "zero-mass spectrometer", 2, STATUS_BOOT},
+    {20, "thermometer", 1, STATUS_BOOT},
+};

### 2017-04-07 data/02-ditl/cfg.h

@@ -1,0 +1,14 @@
+// FCORE Configuration File - FCORE Test 1
+// Generated on 2017-03-29 by fcore_config.rb
+#ifndef FCORE_MAIN_INCLUDE
+#error "direct inclusion of generated configuration file"
+#endif
+
+#define FCORE_MISSION_NAME    "FCORE Test 1"
+#define FCORE_CONFIG_GENDATE  "2017-03-29"
+#define FCORE_PAYLOAD_COUNT   (2)
+#define FCORE_BUS_MAXDATA     (420)
+#define FCORE_BUS_FREQ        (10.164040404040405)
+#define FCORE_BUS_INTERVAL    (51)
+
+extern const FCPayload fcore_payloads[FCORE_PAYLOAD_COUNT];

### 2017-04-07 data/02-ditl/cfg.c
Add tests results

@@ -1,8 +1,0 @@
-// FCORE Configuration File - FCORE Test 1
-// Generated on 2017-03-29 by fcore_config.rb
-#include "fcore.h"
-
-const FCPayload fcore_payloads[FCORE_PAYLOAD_COUNT] = {
-    {10, "zero-mass spectrometer", 2, STATUS_BOOT},
-    {20, "thermometer", 1, STATUS_BOOT},
-};

### 2017-04-07 data/02-ditl/cfg.h

@@ -1,14 +1,0 @@
-// FCORE Configuration File - FCORE Test 1
-// Generated on 2017-03-29 by fcore_config.rb
-#ifndef FCORE_MAIN_INCLUDE
-#error "direct inclusion of generated configuration file"
-#endif
-
-#define FCORE_MISSION_NAME    "FCORE Test 1"
-#define FCORE_CONFIG_GENDATE  "2017-03-29"
-#define FCORE_PAYLOAD_COUNT   (2)
-#define FCORE_BUS_MAXDATA     (420)
-#define FCORE_BUS_FREQ        (10.164040404040405)
-#define FCORE_BUS_INTERVAL    (51)
-
-extern const FCPayload fcore_payloads[FCORE_PAYLOAD_COUNT];

### 2017-04-08 chapters/04-results.tex
Add stress tests results

@@
 \begin{tikzpicture}
-\begin{axis}
-\addplot table [
-xlabel=Mission Elapsed Time,
-ylabel=Altitude,
-x=met, y=alt, col sep=comma
-] {data/02-ditl/loc.csv};
+\begin{axis} [
+xlabel=Mission Elapsed Time (\SI{}{\second}),
+ylabel=Altitude (\SI{}{\metre}),
+mark size=1.0pt,
+mark repeat={5},
+legend style ={
+    at={(0.98,0.95)}, 
+    anchor=north east, draw=black, 
+    fill=white,align=left
+}
+]
+\addplot+ table [
+x=MET, y=altitude, col sep=comma
+] {data/test-ditl/loc.csv};
 \end{axis}
@@
 
-The mission platform was composed of the AHABus \acrlong{obc}, and two simulated
-payloads: payload 10 at priority level 2, and payload 20 at priority level 1.
-Table \ref{tab:ditl-cfg}.
+% The mission platform was composed of the AHABus \acrlong{obc}, and two simulated
+% payloads: payload 10 at priority level 2, and payload 20 at priority level 1.
+% Table \ref{tab:ditl-cfg}.
+%
+% \begin{table}[H]
+% \begin{center}
+% \begin{tabular}{ |c|c c c| }
+% \hline
+% payload & priority & core frequency & polling interval \\
+% \hline
+% GNSS & - & - & - \\
+% 10 & 2 & \SI{5.082}{\hertz} & \SI{102}{\second} \\
+% 20 & 1 & \SI{10.164}{\hertz} & \SI{51}{\second} \\
+% \hline
+% \end{tabular}
+% \end{center}
+% \caption {DITL Test - FCORE configuration}
+% \label{tab:ditl-cfg}
+% \end{table}
+
+Figure \ref{fig:ditl-bandwidth} shows that the average data rate used by the
+platform stabilised early into the flight close to \SI{10}{bps}, or
+\SI{110}{bps} (out of a maximum bandwidth of \SI{200}{bps}). Over the full
+simulation's duration, an average of \SI{0.63}{Bps} were lost due to
+transmission errors.
+
+\begin{figure}[H]
+\begin{tikzpicture}
+\begin{axis} [
+xlabel=Mission Elapsed Time (\SI{}{\second}),
+ylabel=Data Rate (\SI{}{Bps}),
+mark size=1.0pt,
+mark repeat={5},
+legend style ={
+    at={(0.98,0.95)}, 
+    anchor=north east, draw=black, 
+    fill=white,align=left
+}
+]
+\addplot+ table [
+x=MET, y=error rate, col sep=comma
+] {data/test-ditl/fcore.log.rxstats.csv};
+
+\addplot+ table [
+x=MET, y=bandwidth, col sep=comma
+] {data/test-ditl/fcore.log.rxstats.csv};
+
+\legend{Erroneous, Total}
+\end{axis}
+\end{tikzpicture}
+\centering
+\caption{DITL Test - Average data rates}
+\label{fig:ditl-bandwidth}
+\end{figure}
+
+%
+%
+% \begin{figure}[H]
+% \begin{tikzpicture}
+% \begin{axis} [
+% xlabel=Mission Elapsed Time (\SI{}{\second}),
+% ylabel=Received Data (\SI{}{B}),
+% mark size=1.0pt,
+% mark repeat={5},
+% legend style ={
+%     at={(0.02,0.95)},
+%     anchor=north west, draw=black,
+%     fill=white,align=left
+% }
+% ]
+% \addplot+ table [
+% x=MET, y=received, col sep=comma
+% ] {data/test-ditl/fcore.log.rxstats.csv};
+%
+% \addplot+ table [
+% x=MET, y=valid, col sep=comma
+% ] {data/test-ditl/fcore.log.rxstats.csv};
+%
+% \addplot+ table [
+% x=MET, y=invalid, col sep=comma
+% ] {data/test-ditl/fcore.log.rxstats.csv};
+%
+% \legend{Total, Valid Frame Data, Invalid Frame Data}
+% \end{axis}
+% \end{tikzpicture}
+% \centering
+% \caption{DITL Test - Received data}
+% \label{fig:ditl-rxstats}
+% \end{figure}
+
+For the simulated mission, FCORE was configured to transmit a health packet
+every \SI{120}{\second}, reporting the status of the GNSS receiver and the
+two payloads. Table \ref{tab:ditl-health} presents the data extracted from those
+packets: both payloads were available and responding to the bus controller more
+than \SI{90}{\percent} of the time. The actual percentage could differ by a few
+percentage points since the system reports health at fairly large time
+intervals. It is interesting to note that the GNSS receiver is marked as
+recovering in nearly \SI{3}{\percent} of the health messages, indicating an
+inability from the \acrshort{ahabus} platform to parse a valid location from
+its input.
 
 \begin{table}[H]
-\caption {DITL Test - FCORE configuration}
-\label{tab:ditl-cfg}
 \begin{center}
-\begin{tabular}{ |c|c c c| }
+\begin{tabular}{ |c|c c c c| }
 \hline
-payload & priority & core frequency & polling interval \\
+payload & up count & down count & recovering count & uprate \\
 \hline
-GNSS & - & -
-10 & 2 & \SI{5.082}{\hertz} & \SI{102}{\second} \\
-20 & 1 & \SI{10.164}{\hertz} & \SI{51}{\second} \\
+GNSS       & 67 & 0 & 2 & 97.1\% \\
+payload 10 & 64 & 0 & 5 & 92.8\% \\
+payload 20 & 63 & 0 & 6 & 91.3\% \\
 \hline 
@@
 \end{center}
+\caption {DITL Test - Platform health data}
+\label{tab:ditl-health}
 \end{table}
@@
 \section{Radio Stress-Testing}
+
+The final series of tests focused on the radio communications system's
+performance and ability to correct noise-induced errors in the received data.
+As discussed in section \ref{sec:radio-coms}, the \acrfull{ecc} embedded in
+each frame can correct up to sixteen erroneous bytes per frame of 256 bytes.
+A modified version of FCORE was used to corrupt given amounts of bytes before
+transmission and thus simulate a noise radio channel. The three tests were run
+for ten minutes each with increasing error simulation rates below, near and
+above the theoretical limit of the \acrshort{fec} algorithm, at $\frac{8}{256}$,
+$\frac{14}{256}$ and $\frac{18}{256}$ error rates.
+
+Figure \ref{fig:stress-8.256} shows that when the byte error rate is well below
+(half) the theoretical maximum acceptable, the \acrshort{fec} algorithm performs
+well and only a single frame was lost, due to an antenna disconnection.
+
+\begin{figure}[H]
+\begin{tikzpicture}
+\begin{axis} [
+xlabel=Mission Elapsed Time (\SI{}{\second}),
+ylabel=Received Data (\SI{}{\byte}),
+mark size=1.0pt,
+legend style ={
+    at={(0.02,0.95)}, 
+    anchor=north west, draw=black, 
+    fill=white,align=left
+}
+]
+\addplot+ table [
+x=MET, y=received, col sep=comma
+] {data/test-rs8.256/fcore.log.rxstats.csv};
+
+\addplot+ table [
+x=MET, y=invalid, col sep=comma
+] {data/test-rs8.256/fcore.log.rxstats.csv};
+
+\addplot+ table [
+x=MET, y=fixed, col sep=comma
+] {data/test-rs8.256/fcore.log.rxstats.csv};
+
+\legend{Total, Erroneous, Fixed}
+\end{axis}
+\end{tikzpicture}
+\centering
+\caption{$\frac{8}{256}$ Error Test - Received Data}
+\label{fig:stress-8.256}
+\end{figure}
+
+When the average byte error rate, as shown in figure \ref{fig:stress-14.256},
+some frames can be corrected, while some where the error rate is higher than
+the average are lost. The fluctuations in actual error rate are due to the use
+of the C language's \texttt{rand()} function. %remote or explain
+
+\begin{figure}[H]
+\begin{tikzpicture}
+\begin{axis} [
+xlabel=Mission Elapsed Time (\SI{}{\second}),
+ylabel=Received Data (\SI{}{\byte}),
+mark size=1.0pt,
+legend style ={
+    at={(0.02,0.95)}, 
+    anchor=north west, draw=black, 
+    fill=white,align=left
+}
+]
+\addplot+ table [
+x=MET, y=received, col sep=comma
+] {data/test-rs14.256/fcore.log.rxstats.csv};
+
+\addplot+ table [
+x=MET, y=invalid, col sep=comma
+] {data/test-rs14.256/fcore.log.rxstats.csv};
+
+\addplot+ table [
+x=MET, y=fixed, col sep=comma
+] {data/test-rs14.256/fcore.log.rxstats.csv};
+
+\legend{Total, Erroneous, Fixed}
+\end{axis}
+\end{tikzpicture}
+\centering
+\caption{$\frac{14}{256}$ Error Test - Received Data}
+\label{fig:stress-14.256}
+\end{figure}
+
+Finally, an average error rate above the theoretical maximum of \SI{16}{\byte}
+causes most of the frames to be lost as shown in figure \ref{fig:stress-18.256}.
+
+\begin{figure}[H]
+\begin{tikzpicture}
+\begin{axis} [
+xlabel=Mission Elapsed Time (\SI{}{\second}),
+ylabel=Received Data (\SI{}{\byte}),
+mark size=1.0pt,
+legend style ={
+    at={(0.02,0.95)}, 
+    anchor=north west, draw=black, 
+    fill=white,align=left
+}
+]
+\addplot+ table [
+x=MET, y=received, col sep=comma
+] {data/test-rs18.256/fcore.log.rxstats.csv};
+
+\addplot+ table [
+x=MET, y=invalid, col sep=comma
+] {data/test-rs18.256/fcore.log.rxstats.csv};
+
+\addplot+ table [
+x=MET, y=fixed, col sep=comma
+] {data/test-rs18.256/fcore.log.rxstats.csv};
+
+\legend{Total, Erroneous, Fixed}
+\end{axis}
+\end{tikzpicture}
+\centering
+\caption{$\frac{18}{256}$ Error Test - Received Data}
+\label{fig:stress-18.256}
+\end{figure}

### 2017-04-08 chapters/04-results.tex
move temp to out directory

@@
 
-\section{Day In The Life Testing}
+\section{Integrated Testing}
+
+\subsection{Day In The Life Testing}
 
@@
 xlabel=Mission Elapsed Time (\SI{}{\second}),
-ylabel=Data Rate (\SI{}{Bps}),
+ylabel=Received Data (\SI{}{\byte}),
 mark size=1.0pt,
@@
 legend style ={
-    at={(0.98,0.95)}, 
-    anchor=north east, draw=black, 
+    at={(0.02,0.95)}, 
+    anchor=north west, draw=black, 
     fill=white,align=left
@@
 ]
+
 \addplot+ table [
-x=MET, y=error rate, col sep=comma
+x=MET, y=received, col sep=comma
 ] {data/test-ditl/fcore.log.rxstats.csv};
@@
 \addplot+ table [
-x=MET, y=bandwidth, col sep=comma
+x=MET, y=invalid, col sep=comma
 ] {data/test-ditl/fcore.log.rxstats.csv};
 
-\legend{Erroneous, Total}
+\legend{Total, Erroneous}
 \end{axis}
@@
 \centering
-\caption{DITL Test - Average data rates}
+\caption{DITL Test - Received frame data}
 \label{fig:ditl-bandwidth}
 \end{figure}
-
-%
-%
-% \begin{figure}[H]
-% \begin{tikzpicture}
-% \begin{axis} [
-% xlabel=Mission Elapsed Time (\SI{}{\second}),
-% ylabel=Received Data (\SI{}{B}),
-% mark size=1.0pt,
-% mark repeat={5},
-% legend style ={
-%     at={(0.02,0.95)},
-%     anchor=north west, draw=black,
-%     fill=white,align=left
-% }
-% ]
-% \addplot+ table [
-% x=MET, y=received, col sep=comma
-% ] {data/test-ditl/fcore.log.rxstats.csv};
-%
-% \addplot+ table [
-% x=MET, y=valid, col sep=comma
-% ] {data/test-ditl/fcore.log.rxstats.csv};
-%
-% \addplot+ table [
-% x=MET, y=invalid, col sep=comma
-% ] {data/test-ditl/fcore.log.rxstats.csv};
-%
-% \legend{Total, Valid Frame Data, Invalid Frame Data}
-% \end{axis}
-% \end{tikzpicture}
-% \centering
-% \caption{DITL Test - Received data}
-% \label{fig:ditl-rxstats}
-% \end{figure}
 
@@
 
-\section{Radio Stress-Testing}
+\subsection{Radio Stress-Testing}
 
@@
 
-\addplot+ table [
-x=MET, y=fixed, col sep=comma
-] {data/test-rs8.256/fcore.log.rxstats.csv};
-
-\legend{Total, Erroneous, Fixed}
+\legend{Total, Erroneous}
 \end{axis}
@@
 
-\addplot+ table [
-x=MET, y=fixed, col sep=comma
-] {data/test-rs14.256/fcore.log.rxstats.csv};
-
-\legend{Total, Erroneous, Fixed}
+\legend{Total, Erroneous}
 \end{axis}
@@
 
-\addplot+ table [
-x=MET, y=fixed, col sep=comma
-] {data/test-rs18.256/fcore.log.rxstats.csv};
-
-\legend{Total, Erroneous, Fixed}
+\legend{Total, Erroneous}
 \end{axis}
@@
 \end{figure}
+
+Figure \ref{fig:stress-fec} shows that in all three tests, the amount of bytes
+that were corrected using the frames' \acrlong{ecc} was similar. As expected,
+the test with the highest byte error rate shows a slightly inferior correction
+rate – frames that contain more than 16 erroneous bytes cannot be corrected at
+all. Table \ref{tab:stress-fec} shows that the actual error correction rates
+are close to the theoretical byte error rates induced at transmission as long
+as the those theoretical rates are below the theoretical maximum.
+
+\begin{figure}[H]
+\begin{tikzpicture}
+\begin{axis} [
+xlabel=Mission Elapsed Time (\SI{}{\second}),
+ylabel=Error-Corrected Data (\SI{}{\byte}),
+mark size=1.0pt,
+legend style ={
+    at={(0.02,0.95)}, 
+    anchor=north west, draw=black, 
+    fill=white,align=left
+}
+]
+
+\addplot+ table [
+x=MET, y=fixed, col sep=comma
+] {data/test-rs8.256/fcore.log.rxstats.csv};
+
+\addplot+ table [
+x=MET, y=fixed, col sep=comma
+] {data/test-rs14.256/fcore.log.rxstats.csv};
+
+\addplot+ table [
+x=MET, y=fixed, col sep=comma
+] {data/test-rs18.256/fcore.log.rxstats.csv};
+
+\legend{$\frac{8}{256}$, $\frac{14}{256}$, $\frac{18}{256}$}
+\end{axis}
+\end{tikzpicture}
+\centering
+\caption{$\frac{14}{256}$ Error Tests - Error Correction Performance}
+\label{fig:stress-fec}
+\end{figure}

### 2017-04-08 chapters/03-methodology.tex
Add stress tests results
re-structure paper

@@
 
-% TODO: add something about:
-%       - hardware choices
-%       - software choices
-%       - language choices
-%
-%       - overall arch (flight platform, ground segmetn)
+\section{Proposed Solution}
 
@@
 
+\section{Environment}
+
+\section{Development Approach}
+
 \section{Radio Communications}
+\label{sec:radio-coms}
 
@@
 Arduino UNO board for the payload. An unforeseen electrical engineering issue
-required slight redesign of the system: on both the Arduino or the Raspberry
+required a slight redesign of the system: on both the Arduino or the Raspberry
 Pi, a small voltage, even inferior to the logic high (\SI{3.3}{V} on Raspberry
@@
 logical high. This caused certain bursts of noise in the lines to be detected
-by the \acrshort{obc} as a high CR line, even though the payload had not
+by the \acrshort{obc} as a high CR signal, even though the payload had not
 raised it. This was fixed by using inverted logic: a payload would signal 
@@
 
-\subsection{Arduino Implementation}
+% TODO
 
-\section{Flight Software}
+\section{Flight Platform}
 \label{sec:flight-software}
 
-\subsection{Architecture}
+\subsection{Hardware Prototype}
 
-\subsection{User Interface}
+\subsection{Flight Software}
+
+\subsubsection{Payload Priority System}
+
+\subsubsection{Mission Design and Configuration Files}
 

### 2017-04-08 chapters/04-results.tex
re-structure paper

@@
 rate – frames that contain more than 16 erroneous bytes cannot be corrected at
-all. Table \ref{tab:stress-fec} shows that the actual error correction rates
-are close to the theoretical byte error rates induced at transmission as long
-as the those theoretical rates are below the theoretical maximum.
+all.
 
@@
 \end{figure}
+
+Table \ref{tab:stress-fec} shows that the reported error correction rates
+are slightly less than the theoretical error rates induced at transmission as
+long as the these theoretical rates are below the maximum acceptable rate, and
+fall well below when the error rate is above the limit.
+
+\begin{table}[H]
+\begin{center}
+\begin{tabular}{ |c|c c| }
+\hline
+test & Correction Rate & Theoretical Error Rate \\
+\hline
+$\frac{8}{256}$  & \SI{2.826}{\percent} & \SI{3.125}{\percent} \\ [0.2em]
+$\frac{14}{256}$ & \SI{3.655}{\percent} & \SI{5.469}{\percent} \\ [0.2em]
+$\frac{18}{256}$ & \SI{2.514}{\percent} & \SI{7.031}{\percent} \\ [0.2em]
+\hline 
+\end{tabular}
+\end{center}
+\caption {DITL Test - Platform health data}
+\label{tab:stress-fec}
+\end{table}

### 2017-04-08 chapters/05-discussion.tex
re-structure paper

\chapter{Discussion}
\label{ch:discussion}

\section{Introduction}

\section{Project Evaluation}

\subsection{Overview}

\subsection{Radio Communications System}

\subsection{Payload Bus}

\subsection{Ground Segment}

\section{Analysis of Test Results}

\subsection{Day-in-the-Life Test}

\subsection{Radio System Stress Tests}

\section{Further Work} % << mix in with the rest?

### 2017-04-08 paper.tex
Add Aknowledgements

@@
-\input{header.tex}
+\documentclass[11pt]{report}
+\usepackage[utf8]{inputenc}
+\usepackage[english]{babel}
+\usepackage{amsmath}
+\usepackage{amsfonts}
+\usepackage{titlesec}
+\usepackage{float}
+\usepackage{color}
+\usepackage{listings}
+\usepackage{lmodern}
+\usepackage{graphicx}
+\usepackage{bytefield}
+\usepackage{pgfplots}
+\usepackage{csvsimple}
+\usepackage[detect-weight=true, binary-units=true]{siunitx}
+\usepackage{tikz}
+\usepackage[justification=centering,labelfont={small,bf},font={small,bf}]{caption}
+\usepackage[scaled]{beramono} % BitStream Vera font (code)
+\usepackage[T1]{fontenc}
+\usepackage[a4paper, margin=2.54cm]{geometry}
+\usepackage[nottoc]{tocbibind}
+
+%\renewcommand{\familydefault}{\sfdefault}
+\pgfplotsset{width=10cm,height=6cm,compat=newest}
+\graphicspath{{figures/}}
+
+\usepackage[
+style=authoryear,
+citestyle=authoryear,
+backend=biber,
+firstinits=true
+]{biblatex}
+\DeclareNameAlias{author}{last-first}
+
+\addbibresource{../library.bib}
+\DefineBibliographyStrings{english}{%
+  references = {References},
+}
+
+\lstset{basicstyle=\footnotesize\ttfamily,breaklines=true}
+
+\author{Amy Parent}
+\title{AHABus \& FCORE:\\Modular Bus for High-Altitude Balloon Missions}
+\date{April 2017}
+
+% Parameters
+
+\setlength{\baselineskip}{1.5em}
+
+% Title
+
+\newcommand\frontmatter{%
+    \cleardoublepage
+    \pagenumbering{roman}}
+
+\newcommand\mainmatter{%
+    \cleardoublepage
+    \pagenumbering{arabic}}
+
+\makeatletter
+\def\maketitle{%
+    \vspace*{10em}
+	\begin{center}%
+	\let \footnote \thanks
+        \vskip 5em%
+		{\LARGE \textbf{\@title} \par}%
+		\vskip 1.5em%
+		{\large \@author}%
+		\vskip 1em%
+        {\large BSc (Hons) Computing - \@date\\} %
+		\vskip 1em%
+		{\normalsize School of Arts, Media and Games\\}%
+		\vskip .1em%
+		{\normalsize University of Abertay Dundee\\}%
+		\vskip .1em%
+		{\normalsize Dundee, DD1 1HG, UK}%
+	\end{center}
+}
+\makeatother
+
 \input{glossaries.tex}
@@
 
+\chapter*{Acknowledgements}
+\addcontentsline{toc}{chapter}{Acknowledgements}
+
+I would like to thank my supervisor, Dr. Ian Ferguson, for the support and
+insight he gave throughout this research project, and as Dr. Natalie Coull for
+her advice during the proposal phase of the project.
+
 \mainmatter

### 2017-04-08 chapters/04-results.tex
Add Aknowledgements
Better table styles

@@
 
-\section{Component Testing}
+The following section presents the results of the tests done as part of the
+evaluation process of the project, as discussed in section \ref{sec:testing}.
 
-\section{Integrated Testing}
+% \section{Component Testing}
+%
+% \section{Integrated Testing}
 
-\subsection{Day In The Life Testing}
+\section{Day In The Life Testing}
+\label{ssec:results-ditl}
 
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
 \begin{tikzpicture}
@@
 
-% The mission platform was composed of the AHABus \acrlong{obc}, and two simulated
-% payloads: payload 10 at priority level 2, and payload 20 at priority level 1.
-% Table \ref{tab:ditl-cfg}.
-%
-% \begin{table}[H]
-% \begin{center}
-% \begin{tabular}{ |c|c c c| }
-% \hline
-% payload & priority & core frequency & polling interval \\
-% \hline
-% GNSS & - & - & - \\
-% 10 & 2 & \SI{5.082}{\hertz} & \SI{102}{\second} \\
-% 20 & 1 & \SI{10.164}{\hertz} & \SI{51}{\second} \\
-% \hline
-% \end{tabular}
-% \end{center}
-% \caption {DITL Test - FCORE configuration}
-% \label{tab:ditl-cfg}
-% \end{table}
+The mission platform was composed of the AHABus \acrlong{obc}, and two simulated
+payloads: payload 10 at priority level 2, and payload 20 at priority level 1.
+Table \ref{tab:ditl-cfg} shows the core frequency and polling intervals for
+each payload, generated from the mission file by the FCORE configuration tool.
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{c||c c c}
+
+payload & priority & core frequency & polling interval \\ \hline
+10 & 2 & \SI{5.082}{\hertz} & \SI{102}{\second} \\
+20 & 1 & \SI{10.164}{\hertz} & \SI{51}{\second} \\
+
+\end{tabular}
+\end{center}
+\caption {DITL Test - FCORE configuration}
+\label{tab:ditl-cfg}
+\end{table}
 
@@
 
-\begin{table}[H]
+\begin{table}[h]
 \begin{center}
-\begin{tabular}{ |c|c c c c| }
-\hline
-payload & up count & down count & recovering count & uprate \\
-\hline
+\begin{tabular}{c||c c c|c}
+
+payload & up count & down count & recovering count & uptime ratio \\ \hline
 GNSS       & 67 & 0 & 2 & 97.1\% \\
-payload 10 & 64 & 0 & 5 & 92.8\% \\
-payload 20 & 63 & 0 & 6 & 91.3\% \\
-\hline 
+payload 10 & 64 & 0 & 5 & \SI{92.8}{\percent} \\
+payload 20 & 63 & 0 & 6 & \SI{91.3}{\percent} \\
+
 \end{tabular}
@@
 
-\subsection{Radio Stress-Testing}
+The number of valid and invalid packets received for each payload are given in
+table \ref{tab:ditl-packet-count}, along with the expected number of packets
+calculated from the FCORE configuration polling intervals given in table
+\ref{tab:ditl-cfg} and the mission duration of \SI{8482}{\second}. The ratio of
+received packets to expected packets matches the uptime ratios reported by
+FCORE health packets, given in table \ref{tab:ditl-health}.
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{c||c c|c c|c}
+
+payload & valid packets & invalid packets & total & expected & ratio \\
+\hline
+payload 10 & 71 & 2 & 79 & 83 & \SI{95.2}{\percent} \\
+payload 20 & 138 & 6 & 144 & 166 & \SI{86.7}{\percent} \\
+
+\end{tabular}
+\end{center}
+\caption {DITL Test - Received packets}
+\label{tab:ditl-health}
+\end{table}
+
+\section{Radio System Resilience Testing}
+\label{ssec:results-resilience}
 
@@
 
-\begin{table}[H]
+\begin{table}[h]
 \begin{center}
-\begin{tabular}{ |c|c c| }
-\hline
-test & Correction Rate & Theoretical Error Rate \\
-\hline
+\begin{tabular}{c||c c}
+
+test & Correction Rate & Theoretical Error Rate \\ \hline
 $\frac{8}{256}$  & \SI{2.826}{\percent} & \SI{3.125}{\percent} \\ [0.2em]
@@
 $\frac{18}{256}$ & \SI{2.514}{\percent} & \SI{7.031}{\percent} \\ [0.2em]
-\hline 
+
 \end{tabular}

### 2017-04-08 chapters/04-results.tex
Better data parsing

@@
 mark size=1.0pt,
-mark repeat={5},
+mark repeat={10},
 legend style ={

### 2017-04-09 chapters/03-methodology.tex
Add Aknowledgements
Better table styles
Start of discussion

@@
 
-\subsection{Protocol Specification}
+\subsection{AHABus Radio Protocol}
 
@@
 To avoid possible byte order conflicts between the architectures used on the
-flight platform and the ground station, multi-bytes values referred to in the
+flight platform and the ground station, multi-byte values referred to in the
 specification should always be transmitted as big-endian (most significant byte
@@
 The start sequence design had to be changed to occupy a single frame byte, as
-opposed to two bytes in version 2 of the frame format. Using a single byte to
-detect the start of a frame was not deemed sufficient, since the probably of
-that exact value occurring elsewhere in the binary stream would have been
-fairly high. To mitigate this, the third and final version of the frame format
-uses a single frame start marker (\texttt{0x5A}) at the beginning of each
-frame, and each frame must be preceded by at least one synchronisation byte
-(\texttt{0xAA}), inserted in the binary stream as shown in the lower half of 
-figure \ref{fig:frame-sync-extra}. The following 223 bytes in the frame
+opposed to two bytes in the second version of the frame format. Using a single
+byte to detect the start of a frame was not deemed sufficient, since the
+probably of that exact value occurring elsewhere in the binary stream would
+have been fairly high. To mitigate this, the third and final version of the
+frame format uses a single frame start marker (\texttt{0x5A}) at the beginning
+of each frame, and each frame must be preceded by at least one synchronisation
+byte (\texttt{0xAA}), inserted in the binary stream as shown in the lower half
+of figure \ref{fig:frame-sync-extra}. The following 223 bytes in the frame
 (starting with the protocol version number) are error-corrected, with the
@@
 
-\subsection{Protocol Specification}
+\subsection{AHABus Payload Bus Protocol}
 
@@
 \subsection{Hardware Prototype}
+\label{sec:flight-hardware}
 
@@
 \subsubsection{Mission Design and Configuration Files}
+\label{sssec:conf-file}
 
@@
 
-\section{Testing}
+\section{Evaluation}
 \label{sec:testing}
 
-\subsection{Component Testing}
+\subsection{Day in the life Testing}
+\label{ssec:method-ditl}
 
-\subsection{Integrated Testing}
-
-\subsubsection{Day in the life Testing}
-
-\subsubsection{Resilience Testing}
+\subsection{Resilience Testing}
+\label{ssec:method-resilience}

### 2017-04-09 chapters/05-discussion.tex
Add Aknowledgements
Start of discussion

@@
 
-\section{Introduction}
+The aim of the project was to design and build a prototype of a High-Altitude
+Balloon modular platform, similar in principle to buses used in the satellite
+industry, and evaluate both the feasibility of such a platform and its
+performance. The platform was required to allow for stable internal
+communications between the payloads and the \acrlong{obc}, tracking of
+ancillary data (\acrshort{gnss} co-ordinates and altitude) and reliable
+transmission of the collected telemetry to one or multiple ground stations 
+through a low-power radio channel. In this chapter, the \acrshort{ahabus}
+platform prototype is evaluated against these aims, first qualitatively through
+general observations and critical analysis, then in quantitatively, using the
+testing results presented in chapter \ref{ch:results}. This chapter finally
+presents the further work that would be required for the platform to fully meet
+its design goals.
 
@@
 
-\subsection{Overview}
+\subsection{Platform}
+
+As a software platform, the \acrshort{ahabus} prototype fulfils most of its
+goals: the flight \acrshort{fcore} was complete enough to support full-flight
+simulations, manage the bus, \acrshort{gnss} and radio communications without
+failure. During the testing phase, no major software design error were found,
+and the only instances where failures were found to be caused by the flight
+software itself. Though quite cumbersome to work with, and lacking in
+documentation, the \acrshort{fcore} configuration tool discussed in section
+\ref{sssec:conf-file} proved reliable, and the resulting \acrshort{fcore} builds
+did not suffer from erroneous bandwidth usage.
+
+There are, however, areas that could be improved upon: the software design of
+\acrshort{fcore} makes use of numerous data buffers for the different
+transmission layers. Those buffers are required because of the slow speeds that
+those communications protocol use, but are at the moment not optimised to be as
+small as required – instead, all buffers can hold \SI{4096}{\byte}. This would
+prevent the current source code to be ported to architectures that provide less
+dynamic memory than the chosen ESP8266 micro-controller does.
+
+Due to lack of time, very little work was done on the hardware aspect of
+\acrshort{ahabus}: testing was done using a development board instead of the
+bare micro-controller for the \acrlong{obc}, and the power and payload bus were
+little more than wires routed through a breadboard. An improved development
+platform using a printed circuit board was designed as presented in
+\ref{sec:flight-hardware}, but could not be fully assembled and tested in time.
 
@@
 
-\subsection{Radio System Stress Tests}
+\subsection{Radio System Resilience Tests}
 

### 2017-04-09 chapters/03-methodology.tex
Add AHABus Payload Bus Protocol

@@
 
-\begin{enumerate}
+\begin{itemize}
 \item provide a system to transmit binary data.% DATA LINK
@@
 \item allow ground crew to track the balloon's location over time.% PACKET LAYER
-\end{enumerate}
+\end{itemize}
 
@@
 uses the encoder to package that data and outputs the frame stream to the
-UNIX standard output (\texttt{stdout}).
+UNIX standard output (\texttt{stdout}) so that the test decoder, discussed in
+section \ref{sssec:rtx-decoder}, could parse it as input.
 
@@
 \subsubsection{Decoder}
+\label{sssec:rtx-decoder}
 
@@
 \subsection{Requirements}
+
+The requirements established for the payload bus were fairly simple: its only
+role is to collect available data from each payload and make it available to
+the flight software to process and transmit through the radio system.
+Nonetheless a formal list of requirements was established:
+
+\begin{itemize}
+\item Provide a channel between payloads and the \acrshort{obc}.
+\item Prevent multiple payloads from transmitting simultaneously.
+\item Allow for different numbers of payloads with minimum changes.
+\item Minimise the amount of Input/Output pins required on the \acrshort{obc}.
+\end{itemize}
 
@@
 each payload. In addition to the common Rx/Tx lines, each interface is composed
-of a Communication Request Line (CR) and a Communication Enable Line (CE). With
-the new design, the communication sequence had to be changed to the following:
+of a Communication Request Line ($\overline{CR}$) and a Communication Enable
+Line ($\overline{CE}$). With this design, the communication sequence had to
+be changed to the following:
 
 \begin{enumerate}
-\item When a payload is ready to send data, it pulls its CR line high.
+\item When a payload is ready to send data, it pulls its $\overline{CR}$ line
+low.
 \item The \acrshort{obc} lists payloads that have raised CR. For each:
 \begin{enumerate}
-    \item The \acrshort{obc} pulls the payload's CE line high.
+    \item The \acrshort{obc} pulls the payload's $\overline{CE}$ line low.
     \item The payload sends its data to the \acrshort{obc}.
-    \item The \acrshort{obc} pulls the payload's CE line low.
-    \item The payload pulls its CR line low.
+    \item The \acrshort{obc} release the payload's $\overline{CE}$ line to high.
+    \item The payload releases its $\overline{CR}$ line to high.
     \item The \acrshort{obc} processes the received data.
@@
 AND gate (two NAND gates chained together as shown in figure \ref{fig:uart-nor})
-was placed between the payload's Tx line and its CE line. With this design,
-any signal on the Tx line could not get past the gate unless the payload was
-being addressed, thus preventing a payload from corrupting another's signal.
+was placed between the payload's Tx line and its $\overline{CE}$ line. With
+this design, any signal on the Tx line could not get past the gate unless the
+payload was being addressed, thus preventing a payload from corrupting
+another's signal.
 
@@
 \centering
-\caption{Payload Tx line protection}
+\caption{Payload Tx Line Protection}
 \label{fig:uart-nor}
@@
 This solution was tested using a Raspberry Pi 3 for the \acrlong{obc}, and an
-Arduino UNO board for the payload. An unforeseen electrical engineering issue
-required a slight redesign of the system: on both the Arduino or the Raspberry
-Pi, a small voltage, even inferior to the logic high (\SI{3.3}{V} on Raspberry
-Pi, \SI{5}{V} on the Arduino UNO) was enough for the line to be read as a
-logical high. This caused certain bursts of noise in the lines to be detected
-by the \acrshort{obc} as a high CR signal, even though the payload had not
-raised it. This was fixed by using inverted logic: a payload would signal 
-available data by pulling the \(\overline{CE}\) line low, and the \acrshort{obc}
-would pull \(\overline{CR}\) low when addressing a payload.
+Arduino UNO board for the payload. The system worked with the experimental
+setup, and the \acrlong{obc} and the OBC were able to exchange data. However,
+the design was deemed flawed because of the hardware requirements: while the Rx
+and Tx lines are shared between all payloads, each still requires two I/O pins
+on the \acrshort{obc} for the $\overline{CE}$ and $\overline{CR}$ lines. With
+most micro-controllers exposing between 5 and 15 I/O pins, this would have
+strongly limited how many payloads could have been connected at the same time.
 
@@
 \centering
-\caption{Payload Serial Bus test setup}
+\caption{Payload Serial Bus Test Setup}
 \label{fig:uart-nor}
@@
 
-The system worked with the experimental setup, and the \acrlong{obc} and the
-OBC were able to exchange data. However, the design was deemed flawed because
-of the hardware requirements: while the Rx and Tx lines are shared between all
-payloads, each still requires two I/O pins on the \acrshort{obc} for the 
-\(\overline{CE}\) and \(\overline{CR}\) lines. With most micro-controllers
-exposing between 5 and 15 I/O pins, this would have strongly limited how many
-payloads could have been connected at the same time.
+\subsubsection{\acrshort{i2c} Bus Interface}
 
-\subsubsection{\acrshort{i2c} Bus Interface}
+% FIXME: move that to results?
 
@@
 
-% TODO
+After the two experiments described in in the previous sections, \acrshort{i2c}
+was chosen as the payload bus's underlying protocol: it allows a large number of
+payloads to be supported with only two lines (clock line, SCL, and data line,
+SDA), and its controller/controllee paradigm allows the \acrshort{obc} to be
+have complete control over which payload is allowed to transmit, and when.
+
+The AHABus Payload Bus protocol was designed to work with a virtual register
+system: each payload can be thought of like a memory array, and the bus
+controller (the \acrshort{obc}) writes and reads data to and from addresses on
+each payload. Addressing a specific payload on the bus, replying to the bus
+controller and read/write operations are handled by the underlying
+\acrshort{i2c} protocol (\cite{NXPSemiconductors2014}). Table
+\ref{tab:bus-protocol-registers} gives the addresses, sizes and purposes of
+each register. Multi-byte values are always encoded in big-endian order.
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{r||l l l}
+
+address & size & name & description \\ \hline
+\texttt{0x01} & 1 & Con. Flag & Set to \texttt{0xFF} by the \acrshort{obc} when  a connection is open. \\
+\texttt{0x02} & 2 & Data Length & Number of data bytes available \\
+\texttt{0x10} & - & data & Payload data exposed to the \acrshort{obc}
+
+\end{tabular}
+\end{center}
+\caption {AHABus Payload Bus Protocol - Virtual Registers}
+\label{tab:bus-protocol-registers}
+\end{table}
+
+When the bus controller starts polling a payload, a connection is opened by
+writing \texttt{0xFF} to the Connection Flag register. To avoid data races,
+payloads cannot change any of the registers values while the connection flag
+is set. The controller then fetches two bytes from the Data Length register. If
+there is no data to be collected (zero length), the controller closes the
+connection by writing \texttt{0x00} in the connection flag.
+
+If a non-zero amount of data is available, the controller can start requesting
+it by sending \acrshort{i2c} read requests from address \texttt{0x10} onwards.
+To be compatible with devices that only provide small transmissions buffers,
+the controller cannot fetch more than 32 bytes at the same time. If there are
+more than 32 bytes of data available, the controller will send multiple read
+requests with increasing offsets from the base data address.
+
+Once the data has been successfully collected, the controller closes the
+connection by writing \texttt{0x00} in the connection flag.
 

### 2017-04-09 chapters/03-methodology.tex
Begin FCORE

@@
 difference with a \acrshort{uart} serial interface is that \acrshort{i2c} is
-asymmetric: one device acts as the bus controllers, and passive devices can
+asymmetric: one device acts as the bus controller, and passive devices can
 only receive or transmit data when addressed by it.
@@
 
+\subsection{Flight Software}
+
+%TODO: formulation
+The main role of the flight software is to link the subsystems discussed
+previously and handle the data flowing through the system, from each payload's
+raw data to the resulting radio frames. It is also responsible for keeping track
+of the platform geographic co-ordinates and its altitude.
+
+To avoid the added workload of designing and programming a real-time firmware
+from nothing, the \acrfull{fcore}, \acrshort{ahabus}'s flight software, was
+based upon FreeRTOS and the SDK provided by Espressif for the ESP8266
+micro-controllers, which was used as the \acrshort{obc} for the prototype.
+
+FreeRTOS is a real-time kernel that provides a prioritised task model: each task
+is given a function (usually, an event loop) to run and a priority level. At
+regular intervals, the kernel establishes a list of all the tasks that must run,
+and gives processor time to the one with the highest priority. Such a
+deterministic system is desirable for applications like \acrshort{ahabus}, which
+must run on embedded devices with limited computing power, and where certain
+actions must imperatively run at regular intervals. \acrshort{fcore} was
+designed around three main tasks, along with the standalone radio subsystem
+
+\subsubsection{Payload Bus Task}
+Each payload is assigned its own task, with a running
+interval depending on how the payload was configured before flight. Each task
+fetches data from its payload, packetises it and adds it to the radio queue.
+
+\subsubsection{System Health Task}
+
+The system task runs at fixed intervals to check the status of each subsystem
+and create a system health message, sent to the ground station through the
+radio subsystem. It was assigned the highest priority level, so that even if a
+subsystem were to fail and not explicitly release control, FreeRTOS would give
+the System Task the execution context, thus allowing health packets to be sent
+to the ground station.
+
+The health of each subsystem is stored globally, and each system can indicate
+an issue by modifying its entry. There are four health states that subsystems
+can have, represented each by a single ASCII character: booting (B), up (U),
+recovering (R) and down (D). Each system is flagged as booting on startup.
+Payloads are flagged as recovering if the last attempt to poll their data
+failed, and down if more than five attempts to communicate failed in a row.
+the \acrshort{gnss} subsystem is flagged as recovering if no valid position
+fix was received from the receiver, and down if the received messages were not
+parsed correctly.
+
+Health messages were designed to be human-readable: this improves the chances
+that ground teams can understand the status of the platform, even if the
+message's packet and frame were to be corrupted beyond the Frame Layer's
+\acrshort{fec} limits.
+
+The message is sent as a list of comma-separated entries. Each entry starts
+with a letter indicating its type (G for the \acrshort{gnss} subsystem, P for
+payloads). In the case of payloads, the type is followed by the payload's
+identifier as a two-character hexadecimal number. The identifier is then
+followed by a forward slash (``/'') and the status character as discussed above.
+Listing \ref{lst:health-message} shows an example health message.
+
+\begin{lstlisting}[caption=FCORE - Example Health Message, label=lst:health-message]
+FCORE//SYS_HEALTH G/U,P0A/R,P14/U
+\end{lstlisting}
+
+\subsubsection{GNSS Task}
+
+The \acrshort{gnss} task was designed to run at fixed intervals to parse the
+\acrshort{gnss} receiver's output, and update the platform's cached
+co-ordinates and altitude. This cached location is the one included in radio
+packets.
+
+\subsubsection{Radio Subsystem}
+
+\subsection{Mission Design and Configuration Files}
+\label{sssec:conf-file}
+
 \subsection{Hardware Prototype}
 \label{sec:flight-hardware}
-
-\subsection{Flight Software}
-
-\subsubsection{Payload Priority System}
-
-\subsubsection{Mission Design and Configuration Files}
-\label{sssec:conf-file}
 
@@
 \subsection{Extracting Raw Binary Stream}
+\label{sec:binary-extraction}
 
-\subsection{Packet}
+\subsection{Packet Parser}
 

### 2017-04-09 chapters/04-results.tex
Start of discussion
Begin FCORE

@@
 
-The following section presents the results of the tests done as part of the
-evaluation process of the project, as discussed in section \ref{sec:testing}.
+The following section presents the observations and numerical results obtained
+during the tests done as part of the evaluation process of the project, as
+discussed in section \ref{sec:testing}.
 
-% \section{Component Testing}
-%
-% \section{Integrated Testing}
+\section{Observed Issues}
+
+\subsection{Radio Communications}
+
+The initial tests of the radio encoder and transmission stack ported to the
+ESP8266 revealed issues with the micro-controller's interrupt system: events for
+both \acrshort{uart} share the same interrupt, and \texttt{UART0} events (used
+when communicating with the \acrshort{gnss} receiver) were detected at the same
+time as \texttt{UART1} events. Additionally, the design of the ESP8266 meant
+that the interrupts would fire not just when the transmission buffer crossed
+its low threshold, but whenever it was filled below the threshold, locking the
+controller into an interrupt loop and eventually triggering the watchdog timer.
+
+Once those issues were fixed, as discussed in section \ref{sec:flight-software},
+the radio protocol performed without issues on the transmission side. Reception
+proved mostly stable apart from one main issue caused by the frame format: since
+the end of a frame is detected by the number of byte received, byte losses
+while receiving a frame would cause the start of the subsequent frame to be
+counted as the end of the current one. As a result, dropped bytes in a frame
+would most often cause the next frame to not be detected even if it was formed
+properly.
+
+\subsection{Payload Bus}
+
+Though the payload bus was found to be reliable overall, two issues were found
+during platform testing. First, the default \acrshort{i2c} library provided
+with the ESP8266 SDK had a very short clock stretch timeout, which meant that
+payloads had a very short time window to respond to any request from the bus
+controller, leading to a high rate of failed communications until the stretch
+value was raised. Second, there were some cases where the last command from the
+bus controller (clear the connection flag register) would fail. As a result,
+the payload would be prevented to write any new data until the next
+communication from the bus controller.
 
@@
 \SI{110}{bps} (out of a maximum bandwidth of \SI{200}{bps}). Over the full
+simulation's duration, an average of \SI{0.63}{Bps} were lost due to
+transmission errors.
+
+Figure \ref{fig:ditl-bandwidth} shows that the average data reception rate was
+stable over the whole simulation, with an average of \SI{11}{Bps}, or
+\SI{121}{bps} (out of a maximum bandwidth of \SI{200}{bps}). Over the full
 simulation's duration, an average of \SI{0.63}{Bps} were lost due to

### 2017-04-09 chapters/03-methodology.tex
Add GNSS task

@@
 actions must imperatively run at regular intervals. \acrshort{fcore} was
-designed around three main tasks, along with the standalone radio subsystem
+designed around three main tasks, along with the standalone radio subsystem.
+Figure \ref{fig:fcore-arch} shows an overview of the the \acrshort{fcore}
+architecture, and the component presented in the following sections.
+
+\begin{figure}[H]
+\includegraphics[width=0.6\textwidth]{fcore-arch.pdf}
+\centering
+\caption{\acrshort{fcore} architecture}
+\label{fig:fcore-arch}
+\end{figure}
 
 \subsubsection{Payload Bus Task}
-Each payload is assigned its own task, with a running
-interval depending on how the payload was configured before flight. Each task
-fetches data from its payload, packetises it and adds it to the radio queue.
+
+The payload bus subsystem was designed so that each payload connected to the
+platform is given a unique identifier, which is also its address on the
+\acrshort{i2c} bus. To avoid errors emanating from one payload to propagate
+and take the whole payload bus offline, each payload is assigned a Payload
+Bus Task.
+
+A Bus Tasks runs at a fixed interval, configured at compile time depending on
+the payload's priority level, as discussed in more details in section
+\ref{sssec:conf-file}. The task establishes a connection with the payload using
+the \acrshort{ahabus} payload bus protocol. If data is available, it is
+collected into a temporary buffer and a packet with the platform's cached
+geographical coordinates is created and sent to the radio submodule.
+
+While the ESP8266 micro-controller does not have any hardware \acrshort{i2c}
+support, the FreeRTOS SDK provides an implementation based on hardware timers
+that was used for bus operations. As discussed in the results (section
+\ref{ssec:results-bus}), the library had to be modified to extend the
+\acrshort{i2c} clock stretch timeout to \SI{100000}{\micro\second}, so that
+payloads with relatively slow processors had enough time to reply to commands.
+
+Because all the payload bus tasks run at the same priority level, there was a
+possibility that a payload's task could take control of the kernel while another
+was still communicating with its own payload. To avoid this, the portion of the
+code that handles communications over the bus and packetisation was marked
+as critical, preventing FreeRTOS from triggering a context switch until all data
+is collected.
 
@@
 
-The \acrshort{gnss} task was designed to run at fixed intervals to parse the
-\acrshort{gnss} receiver's output, and update the platform's cached
-co-ordinates and altitude. This cached location is the one included in radio
-packets.
+The \acrshort{gnss} task was designed to run at fixed intervals of
+\SI{20}{\second} to parse the \acrshort{gnss} receiver's output, and update the
+platform's cached co-ordinates and altitude. This cached location is the one
+included in radio packets.
+
+\acrshort{fcore} was designed to use \acrshort{gnss} receivers that comply with
+the NMEA 0183 standard (\cite{Bekte2001}). In order to avoid the added risk of
+writing a custom parser for an existing standard, the minmea library
+(\cite{Moczek2016}) was chosen for its simplicity and reasonable size.
+
+The NMEA sentences used to keep track of a receiver positions are called GGA.
+Since receivers transmit multiple types of sentences every second, the
+\acrshort{gnss} task was designed to attempt to parse up to five sentences
+every time it runs, until a GGA sentence is found. If no GGA sentence is 
+received, the task sets the \acrshort{gnss} health status to recovering to
+indicate that transmitted positions might be out of date.
 

### 2017-04-10 chapters/03-methodology.tex
Add bus frequency

@@
 The following chapter describes the design processes used to develop and
-integrate those components into a functioning prototype of the Abertay
-High-Altitude Bus (\textit{AHABus}), and the methods used to test the viability
-and performance of the design.
+integrate those components into a functioning prototype of the
+\acrfull{ahabus}, and the methods used to test the viability and performance of
+the design.
 
@@
 
-\section{Development Approach}
+The proposed solution was designed to run on low-cost embedded platforms, and
+the prototype was built around the Espressif ESP8266 micro-controller
+(\acrlong{obc}). The radio link used a Radiometrix NTX2B transmitter
+(\cite{radiometrix2012}), and an unnamed Software-Defined Radio receiver and
+antenna.
+
+The software packages used during the project were:
+
+\begin{description}
+
+\item[ESP OpenRTOS SDK] A FreeRTOS-based real-time kernel for the ESP8266.
+\item[Clang/LLVM] C compiler for macOS (version 802.0.38).
+\item[GQRX] macOS Software-Defined Radio application.
+\item[dl-fldigi] digital radio decoding software.
+\item[Textmate] Code editor.
+\item[C (C99 standard)] Language used for all parts of \acrshort{ahabus}.
+\item[Python 2.7] Language used for log file analysis
+
+\end{description}
 
@@
 
-\subsubsection{Initial Testing}
+The decoding functions of the library were designed with threaded applications
+in mind: the decoding function is given a callback called whenever a packet
+is decoded, and never returns.
+
+Decoding is done in a sequential way. Once the start of a frame has been
+detected, the next 256 bytes are collected. Those 256 bytes are then checked
+at increasing levels of details. Once a frame has been received, the first
+action that the decoder runs is \acrlong{fec}. If all errors can be corrected,
+the packet header is then extracted from the frame, as well as the beginning of
+the packet's data. Subsequent frames are decoded the same way, except all of
+their content is added to the packet's data, until the enough bytes have been
+decoded to account for the packet's indicated length.
+
+The user-provided callback function is invoked as soon as a full packet is
+decoded. It is given a flag indicating whether any erroneous frames were
+detected, so that potentially corrupted packets can be marked.
 
@@
 
+The radio subsystem was designed around the ESP8266's \texttt{UART1}. While it
+provides a hardware transmission buffer, tis size is only 127 bytes. Given the
+radio channel's low bandwidth (\SI{200}{bps}), this was considered insufficient.
+To allow whole packets to be sent to the radio without forcing the calling code
+to wait until the buffer is empty, the hardware buffer was supplemented with
+a much larger buffer in software. Whenever data is sent to the radio system,
+it is placed in this buffer. An Interrupt Service Routine so that whenever the
+hardware buffer is empty, 127 bytes are copied from the software buffer to the
+hardware one.
+
+To avoid overhead, the radio subsystem was not given any threading protections.
+Instead, calling code was designed so that no two places can be interacting with
+the radio subsystem at the system: any function that creates packets and places
+them in the radio queue is marked as critical, preventing FreeRTOS to switch
+contexts.
+
 \subsection{Mission Design and Configuration Files}
@@
 
+Because \acrshort{fcore} was designed to run on embedded platforms with 
+potentially no support for configuration files, the settings for a given mission
+(payload addresses, priority levels...) had to be written as part of the
+firmware source code. To avoid requiring future users having to understand and
+modify the source code themselves, a code generation approach was taken: a
+tool parses a configuration file and generates C source files containing the
+mission configuration data, after which the firmware can be compiled and flashed
+without manual editing of its source. An example configuration file is shown
+in listing \ref{lst:conf-file}
+
+\begin{lstlisting}[
+    caption=FCORE - Example Configuration File,
+    label=lst:conf-file
+]
+{
+    "name": "FCORE Test 1",
+    "payloads": [
+        {
+            "address":  10,
+            "name":     "zero-mass spectrometer",
+            "priority": 2
+        },
+        {
+            "address":  20,
+            "name":     "thermometer",
+            "priority": 1
+        }
+    ]
+}
+\end{lstlisting}
+
+This tool was also used to optimise the polling frequency for each payload.
+A payload's priority level defines how often it is polled: level two payloads
+are polled half as often as level one, and level three a third as often. Since
+the number of payloads, their level and the available bandwidth are known when
+the file is parsed, a formula could be obtained to calculate the optimal
+polling interval, provided the maximum size of a packet was fixed (during the
+tests, this value was fixed to 420 bytes, which fit in two 256-byte frames,
+resulting in at most 512 bytes generated each time a payload is polled).
+
+If $B$ is the total available bandwidth, $n_1, n_2, n_3$ the number of payloads
+at each level, $F_P$ the average byte rate of a level 1 payload and $F_S$ the
+average byte rate of the system health packets:
+
+\[B = F_S + n_{1}F_{P} + \frac{n_{2}F_{P}}{2} + \frac{n_{3}F_{P}}{3} \]
+
+\[B = F_S + F_{P}(n_{1} + \frac{n_2}{2} + \frac{n_3}{3}) \]
+
+\[F_P = \frac{B-F_S}{n_{1} + \frac{n_2}{2} + \frac{n_3}{3}} \]
+
+Once the average byte rate of a payload, its priority level $P$ and how many
+frames are required to encapsulate the data it can return, the interval can be
+obtained:
+
+\[ \Delta_T = P \times \frac{256 \times N_{Frames}}{F_P} \]
+
+The configuration tool was written in ruby because it allows easy parsing of
+JSON files. An example output source file is shown in figure \ref{fig:conf-out},
+generated from the example configuration file discussed previously.
+
+\begin{lstlisting}[
+    caption=FCORE - Example Configuration Output,
+    label=lst:conf-out
+]
+// FCORE Configuration File - FCORE Test 1
+// Generated on 2017-03-29 by fcore_config.rb
+
+#define FCORE_MISSION_NAME    "FCORE Test 1"
+#define FCORE_CONFIG_GENDATE  "2017-03-29"
+#define FCORE_PAYLOAD_COUNT   (2)
+#define FCORE_BUS_MAXDATA     (420)
+#define FCORE_BUS_FREQ        (10.164040404040405)
+#define FCORE_BUS_INTERVAL    (51)
+
+const FCPayload fcore_payloads[FCORE_PAYLOAD_COUNT] = {
+    {10, "zero-mass spectrometer", 2, STATUS_BOOT},
+    {20, "thermometer", 1, STATUS_BOOT},
+};
+\end{lstlisting}
+
 \subsection{Hardware Prototype}
 \label{sec:flight-hardware}
+
+Because a large part of the project was focused on developing the protocols and
+software that support the \acrshort{ahabus} platform, most of the testing was
+done on development boards and breadboards. Because the NTX2 radio transmitter
+requires voltage differentials to control the frequency, and it was controlled
+using a \acrshort{uart} signal, a voltage divider bridge was added between the
+digital I/O pin and its frequency control pin.
+
+\begin{figure}[H]
+\includegraphics[width=0.4\textwidth]{ahabus-schema.png}
+\centering
+\caption{AHABus - Electronic Circuit}
+\label{fig:ahabus-schema}
+\end{figure}
+
+Some preliminary printed circuit board layouts were designed for a next round
+of testing. The schematics and the PCB layout are shown in figures
+\ref{fig:ahabus-schema} and \ref{fig:ahabus-pcb}.
+
+\begin{figure}[H]
+\includegraphics[width=0.4\textwidth]{ahabus-pcb.png}
+\centering
+\caption{AHABus - Printed Circuit Board Layout}
+\label{fig:ahabus-pcb}
+\end{figure}
 

### 2017-04-10 chapters/03-methodology.tex
Finish methodology

@@
 The configuration tool was written in ruby because it allows easy parsing of
-JSON files. An example output source file is shown in figure \ref{fig:conf-out},
-generated from the example configuration file discussed previously.
+JSON files. An example output source file is shown in listing 
+\ref{lst:conf-out}, generated from the example configuration file discussed
+previously.
 
@@
     caption=FCORE - Example Configuration Output,
-    label=lst:conf-out
+    label=lst:conf-out,
+    language=c
 ]
@@
 
+The ground segment was based on Dl-Fldigi, an open-source binary radio decoder.
+While Dl-Fldigi can decode RTTY signals, it was meant to decode RTTY text and
+not binary data like the \acrshort{ahabus} packet radio protocol requires.
+The data decoded by the software is not written to any file, but instead printed
+in the main window, which means that byte which values are outside the range of
+ASCII printable characters were invisible.
+
+In order to allow a decoding tool to work on the full binary stream, a fork of
+Dl-Fldigi was modified so that the binary stream could be extracted. When the
+modified version starts, it creates a local network socket, to which any
+byte decoded from RTTY streams are forwarded. This design was chosen because
+most programming languages provide support for network communications, which
+means that tools could easily be written to work with AHABus frame streams, such
+as the packet parser discussed in the following section.
+
 \subsection{Packet Parser}
+\label{sec:packet-dispatcher}
+
+For the purpose of debugging, a simple command-line based packet parser and
+monitor was built upon the decoder discussed in section
+\ref{sssec:rtx-decoder}. Because the platform was still at the prototype stage,
+it was kept fairly simple. The location metadata contained in the packets is
+displayed directly, as are system health packets. The data contained in packets
+coming from payloads is written directly to one file for each payload. Listing
+\ref{lst:fcore-parser-out} shows an example output from the packet parser.
+
+\begin{lstlisting}[
+    caption=Ground Segment - Example Packet Parser Output,
+    label=lst:fcore-parser-out
+]
+[1490711516]: frame start detected
+[1490711531]: decoded valid payload packet (0x14)
+[1490711531]: 	>>>> FIX(56.462000deg, -2.970700deg, 5579m)
+[1490711531]: rx stats: 7569 received, 7424/0 frame bytes (0 fixed)
+[1490711545]: frame start detected
+[1490711559]: decoded valid system packet
+[1490711559]: 	>>>> FIX(56.462000deg, -2.970700deg, 5747m)
+[1490711559]: 	>>>> FCORE//SYS_HEALTH :: G/U,P0A/U,P14/U
+[1490711559]: rx stats: 7830 received, 7680/0 frame bytes (0 fixed)
+[1490711569]: frame start detected
+\end{lstlisting}
+
+For testing purposes, the parser was also set to output statistics on the amount
+of data received, fixed errors and number of frames decoded or lost.
 
@@
 
+Evaluation and testing of the project proved difficult because of the nature
+of the prototype: embedded systems provide little, if any way to debug and
+observe code behaviour. Instead, most of the testing had to focus on the data
+received by the ground segment. Nonetheless, testing was conducted through
+two main methods discussed in this section.
+
+For all tests, the experimental setup was kept consistent: an ESP8266
+development board was used as the \acrshort{obc}, while two Arduino Uno boards
+were connected as payloads, generating fake instrument data. A \SI{5}{\volt}
+power supply was connected to the power lines.
+
 \subsection{Day in the life Testing}
@@
 
+The day-in-the-life test was meant to simulate a full \acrshort{hab} mission,
+in order to evaluate both the overall viability of the platform, and test the
+performance of the \acrshort{ahabus} platform's subsystems.
+
+In order to come as close as possible to real-life conditions, the
+\acrshort{gnss} receiver was replaced with an Arduino board programmed to
+simulate the position updates of a receiver attached to a \acrlong{hab}
+throughout ascent and descent.
+
+Because the test was executed locally, the long-range aspect of the radio
+communications could only be simulated by using a short antenna on the reception
+side, and no antenna on the simulated platform.
+
 \subsection{Resilience Testing}
 \label{ssec:method-resilience}
+
+The goal of the resilience tests was to test the performance of the packet radio
+protocol, and specifically the \acrlong{fec}/Frame layer. To simulate a noisy
+radio channel, a modified version of \acrshort{fcore} was built so that it
+would randomly change the value of a byte before transmitting it. Knowing the
+maximum theoretical correction limit of the \acrshort{fec} algorithm
+($\frac{16}{256}$), three average error rates
+($\frac{corrupted}{transmitted}$) were tested:
+
+\begin{itemize}
+    \item well under the limit ($\frac{8}{256}$);
+    \item just under the limit ($\frac{14}{256}$);
+    \item above the limit ($\frac{18}{256}$).
+\end{itemize}

### 2017-04-12 chapters/04-results.tex
Add GNSS task
pre-discussion

@@
 
-\section{Observed Issues}
+\section{General Observations}
+
+Overall, the \acrshort{ahabus} prototype met expectations and performed well.
+While some minor issues were encountered during the initial round of testing,
+both endpoints of the platform – flight platform and ground segment – functioned
+well enough that the next round of testing (range and flight tests) would be
+achievable.
 
 \subsection{Radio Communications}
+\label{ssec:results-rtx}
 
@@
 \subsection{Payload Bus}
+\label{ssec:results-bus}
 

### 2017-04-12 chapters/05-discussion.tex
pre-discussion

@@
 
+While the platform proved reliable and would be well suited for scientific
+missions, where most of the generated telemetry is numerical measurements, it
+might be less adequate for student missions where most users would expect to
+transmit more data, like real-time pictures: the packets size limit
+(\SI{420}{\byte}) would make transmitting any reasonably-sized image impossible.
+This issue could potentially be solved by adding time division algorithms other
+than the one presented in section \ref{ssec:conf-file}.
+
 Due to lack of time, very little work was done on the hardware aspect of

### 2017-04-12 chapters/05-discussion.tex
Add ground segment discussion

@@
 through a low-power radio channel. In this chapter, the \acrshort{ahabus}
-platform prototype is evaluated against these aims, first qualitatively through
-general observations and critical analysis, then in quantitatively, using the
-testing results presented in chapter \ref{ch:results}. This chapter finally
-presents the further work that would be required for the platform to fully meet
-its design goals.
+platform prototype is evaluated against these aims, first by analysing the
+results of tests discussed in chapter \ref{ch:results}, then through more
+general qualitative observation and analysis. This chapter finally presents the
+further work that would be required for the platform to fully meet its design
+goals, leading to potential test flights.
+
+\section{Analysis of Test Results}
+
+This section aims to explain and analyse the quantitative results obtained
+during the testing phase of the project. The results are presented in chapter
+\ref{ch:results}, and the raw data files obtained from each tests are included
+in appendices \ref{} through \ref{}, or on the project's GitHub account at
+https://github.com/ahabus/results
+
+\subsection{Day-in-the-Life Test}
+
+\subsection{Radio System Resilience Tests}
 
@@
 
-\section{Analysis of Test Results}
+The ground segment software was found to be reliable and sufficient for the
+initial round of testing. All requirements were met: the modified version of
+Dl-Fldigi was stable during the tests and successfully broadcasted the decoded
+raw binary stream. The packet decoder/parser was sufficient to gather
+information about the radio communication statistics and the platform's health
+during missions, and received payload data was successfully written to disk
+and marked as either valid or invalid.
 
-\subsection{Day-in-the-Life Test}
-
-\subsection{Radio System Resilience Tests}
+While the current toolchain was sufficient for testing, and would likely be
+usable for further tests, better packet forwarding would be required for a
+finished software suite. While the current solution writes data directly to the
+local drive of the receiving computer, a better solution might be to forward
+decoded data and metadata for each packet to a socket on the local network, or
+a server, so that client applications can be built for each team. This would
+allow customised dashboards more apt to decode payload-specific data, and the
+parser would only act as a dispatcher, and possibly a way to monitor the
+\acrshort{ahabus} platform itself during its flight.
 

### 2017-04-12 chapters/05-discussion.tex
Add byte stuffing to radio

@@
 
-\section{Analysis of Test Results}
+\section{Project Evaluation}
 
-This section aims to explain and analyse the quantitative results obtained
-during the testing phase of the project. The results are presented in chapter
-\ref{ch:results}, and the raw data files obtained from each tests are included
-in appendices \ref{} through \ref{}, or on the project's GitHub account at
-https://github.com/ahabus/results
-
-\subsection{Day-in-the-Life Test}
-
-\subsection{Radio System Resilience Tests}
-
-\section{Project Evaluation}
+Overall, the project could be considered a success. As further section discuss,
+some issues were encountered, and most aspects of the \acrshort{ahabus}
+platform, protocols and toolchain still require refinement and further testing,
+but the results and observation presented in chapter \ref{ch:results} showed
+that developing a modular platform for \acrlong{hab} missions based on the
+concept of satellite buses is feasible.
 
@@
 
+The radio protocol and its software implementation proved equally time consuming
+and successful. For reasons explained in this section, the protocol could not
+be considered final or perfect, but the version developed over the duration of
+the project proved reliable enough to transmit large amounts of data over long
+simulated test missions.
+
+There were however some observed issues, both in the design of the protocol and
+in the frame and packet decoder implementation. First, the current use of a
+16-bit unsigned integer to encode the platform's altitude in packet's header
+could be problematic even though it did not cause problems during testing: first
+because some locations on Earth are located below the mean sea level, leading to
+negative altitudes that cannot be encoded with unsigned integers. Then because
+the use of 16 bits, while sufficient to encode the altitudes at which balloons
+normally fly, leave little to no margin for expansion, for example to
+\acrshort{hab}-launched small rockets, or eventually nano-satellites.
+
+Furthermore, it was observed that the reliance on frame length to detect the
+end of a frame is error-prone. The current implementation of the frame decoder
+counts the 256 bytes that follow a frame start marker as the frame,
+unconditionally. As described in section \ref{ssec:results-rtx}, this causes
+an issue where dropped bytes in the middle of the frame will cause the start
+of the following frame to be consumed as the end of the current one.
+
+One solution to this issue would be the use of an end-of-frame marker instead,
+and to reject frames that are shorter than 256 bytes. However, this approach
+would be challenging to implement in the context of the \acrshort{ahabus} radio
+protocol. Because the end marker value might occur in the frame's content, it
+must be escaped so that the decoder does not interpret a content byte as the
+frame end. This can be achieved through \textit{byte stuffing}: an escape value
+is chosen, and any end marker or escape value that occurs in the content must
+be preceded by the escape value. The use of \acrshort{fec} in the protocol
+further complicates the issue: the \acrlong{ecc} would have to be escaped too,
+making frames impossible to error-correct before the escaped bytes have been
+resolved by the decoder. This means that a single corrupted escape byte in a
+received frame would prevent the frame from being corrected at all, even if the
+number of invalid bytes is inferior to the \acrshort{fec} algorithm limit.
+
 \subsection{Payload Bus}
@@
 
+\section{Analysis of Test Results}
+
+This section aims to explain and analyse the quantitative results obtained
+during the testing phase of the project. The results are presented in chapter
+\ref{ch:results}, and the raw data files obtained from each tests are included
+in appendices \ref{} through \ref{}, or on the project's GitHub account at
+https://github.com/ahabus/results
+
+\subsection{Day-in-the-Life Test}
+
+\subsection{Radio System Resilience Tests}
+
 \section{Further Work} % << mix in with the rest?

### 2017-04-13 chapters/03-methodology.tex
pre-discussion
Change bus status semantics

@@
 recovering (R) and down (D). Each system is flagged as booting on startup.
-Payloads are flagged as recovering if the last attempt to poll their data
-failed, and down if more than five attempts to communicate failed in a row.
-the \acrshort{gnss} subsystem is flagged as recovering if no valid position
-fix was received from the receiver, and down if the received messages were not
-parsed correctly.
+Payloads are flagged as recovering if the payload attempted to transmit more
+data than it is allowed, and down if it failed to respond to polling. the
+\acrshort{gnss} subsystem is flagged as recovering if no valid position fix was
+received from the receiver, and down if the received messages were not parsed
+correctly.
 
@@
 \subsection{Mission Design and Configuration Files}
-\label{sssec:conf-file}
+\label{ssec:conf-file}
 
@@
 \end{itemize}
+
+Each test involved having the prototype platform run for ten minutes in normal
+conditions, an attempting to decode packets using ground segment software.

### 2017-04-13 chapters/05-discussion.tex
Change bus status semantics

@@
 documentation, the \acrshort{fcore} configuration tool discussed in section
-\ref{sssec:conf-file} proved reliable, and the resulting \acrshort{fcore} builds
+\ref{ssec:conf-file} proved reliable, and the resulting \acrshort{fcore} builds
 did not suffer from erroneous bandwidth usage.
@@
 is chosen, and any end marker or escape value that occurs in the content must
-be preceded by the escape value. The use of \acrshort{fec} in the protocol
+be preceded by the escape value. The use of \acrlong{fec} in the protocol
 further complicates the issue: the \acrlong{ecc} would have to be escaped too,

### 2017-04-13 chapters/04-results.tex
Add radio discussion

@@
 \label{ch:results}
+
+\section{Introduction}
 
@@
 
-Figure \ref{fig:dit-alt} shows the altitude of the AHABus platform reported by
+Figure \ref{fig:ditl-alt} shows the altitude of the AHABus platform reported by
 the received headers over the approximately two hours and twenty minutes of the
@@
 \caption{DITL Test - Altitude over time}
-\label{fig:dit-alt}
+\label{fig:ditl-alt}
 \end{figure}
@@
 \end{table}
-
-Figure \ref{fig:ditl-bandwidth} shows that the average data rate used by the
-platform stabilised early into the flight close to \SI{10}{bps}, or
-\SI{110}{bps} (out of a maximum bandwidth of \SI{200}{bps}). Over the full
-simulation's duration, an average of \SI{0.63}{Bps} were lost due to
-transmission errors.
 

### 2017-04-13 chapters/05-discussion.tex
Add radio discussion

@@
 \label{ch:discussion}
+
+\section{Introduction}
 
@@
 
+% Finally, it was observed during all tests that the loss of a frame could be
+% reported twice in the reception statistics: the last decoded frame number is
+% only changed when a valid frame is parsed, and the frame sequence number is
+% discarded if the frame contains too many erroneous bytes and cannot be
+% corrected. While this avoid tracking corrupt frame sequence numbers in the
+% decoder, it means that a corrupt frame will not be registered in the sequence,
+% and the decoder will report a dropped frame when
+
 \subsection{Payload Bus}
+
+The payload bus
 
@@
 
+The main goal of the day-in-the-life test was to evaluate whether the platform
+was viable for full-duration \acrlong{hab} missions. In that respect, the test
+can be considered successful: as shown in figure \ref{fig:ditl-alt}, the
+simulation spanned a whole mission sequence, from launch through burst altitude,
+to descent and landing, over a slightly more than two hours and twenty minutes.
+
+While actions were taken for the test to mirror an actual mission as much as
+possible, like the use of fake \acrshort{gnss} messages, one effect in
+particular that had to be ignored was the weakening of the radio signal: in a
+real mission, the increase in the distance between the ground station and the
+platform over time means that the radio channel qualities (signal-to-noise
+ratio, line of sight interruptions) are not constant. For this test, neither the
+transmission nor the reception antenna were moved. Range resilience would have
+to be tested through dedicated tests before the platform is used on real
+missions.
+
+The test allowed a long duration check of the radio protocol and systems. Figure
+\ref{fig:ditl-bandwidth} shows that the amount of corrupted frames was well
+below the total amount of data that was received. The average rate of
+transmission, however, was only slightly above half of the available bandwidth,
+despite the use of the calculation of the optimal intervals using the tool
+discussed in section \ref{ssec:conf-file}.
+
+This disparity between actual and optimal use of the bandwidth was caused by the payloads: the time division formula used to compute the polling interval solves for the case where each payload always transmit exactly the maximum amount of data that can be contained in packet.
+During the test, one of the payload generated data at a low enough rate that it fit into small packets that could be encapsulated into a single frame, reducing the required bandwidth.
+This issue would be challenging to address with the current architecture since it requires the polling intervals to be fixed.
+A better use of the bandwidth could be approached by selecting appropriate payload priority levels during the mission design phase, or designing payload to use the full capacity of the radio communications layer.
+
+Finally, table \ref{tab:ditl-health} shows that even though the data bus
+performed well overall, both test payloads were marked as unresponsive in
+nearly \SI{10}{\percent} of the system health packets transmitted by the
+platform. The source of the problem could not be identified due to the lack of
+time and the difficulty of debugging this subsystem in particular: the
+Wemos D1 development board that was used as the \acrlong{obc} does not provide
+useful printing devices that could have been used to send log messages.
+
 \subsection{Radio System Resilience Tests}

### 2017-04-13 chapters/05-discussion.tex
Add bus discussion

@@
 
-% Finally, it was observed during all tests that the loss of a frame could be
-% reported twice in the reception statistics: the last decoded frame number is
-% only changed when a valid frame is parsed, and the frame sequence number is
-% discarded if the frame contains too many erroneous bytes and cannot be
-% corrected. While this avoid tracking corrupt frame sequence numbers in the
-% decoder, it means that a corrupt frame will not be registered in the sequence,
-% and the decoder will report a dropped frame when
+\subsection{Payload Bus}
+\label{ssec:discussion-bus}
 
-\subsection{Payload Bus}
+The payload data bus worked well overall, even though payloads failed to
+respond in certain occurrences. Due to the absence of debugging facilities on
+the Wemos D1 development board and lack of time, the root cause of this issue
+could not be successfully identified. One possibility is the difference in
+logical levels between the \acrlong{obc} (\SI{3.3}{\volt}) and the test
+payloads (\SI{5.0}{\volt}). While in practice both side were able to detect
+logical highs and lows, it is possible that performance could be improved by
+using a voltage regulator on the data and clock lines. Additionally, the test
+payloads were written using the Arduino libraries, which provide an easy-to-use
+API to interact with AVR-based boards, but include a lot of overhead. It is
+possible that some length interrupt service routines or I/O actions were
+delaying the payloads' responses to \acrshort{i2c} requests, causing the
+\acrshort{obc} to flag them as unresponsive.
 
-The payload bus
+Of all the systems designed for this project, the payload bus was the most
+simple – it relies heavily on \acrshort{i2c}, an existing protocol – but also
+the one that was the most complex to test thoroughly, and the one that would
+require the most additional testing to be approved before a real test mission.
 
@@
 
+It is important to note that the amount of erroneous data received was lower
+than that reported by the decoder/parser, due to software design. 256 bytes are
+added to the erroneous data count when a frame has too many errors to be
+error-corrected, or when frames are skipped – the current frame sequence number
+is higher than expected. When a corrupt frame is detected, the last decoded
+sequence number is not changed to avoid overwriting it with a potentially
+corrupted value. Because of this, the next valid frame to be detected will
+also cause a dropped frame to be reported, even though the corrupt's frame
+256 bytes have already been recorded.
+
 Finally, table \ref{tab:ditl-health} shows that even though the data bus
 performed well overall, both test payloads were marked as unresponsive in
-nearly \SI{10}{\percent} of the system health packets transmitted by the
-platform. The source of the problem could not be identified due to the lack of
-time and the difficulty of debugging this subsystem in particular: the
-Wemos D1 development board that was used as the \acrlong{obc} does not provide
-useful printing devices that could have been used to send log messages.
+nearly \SI{10}{\percent}, which can be considered high. As discussed in section
+\ref{ssec:discussion-bus}, a single root cause could not be identified for this
+particular issue, and more testing would be required.
 

### 2017-04-13 chapters/05-discussion.tex
Add stress tests discussion

@@
 
+The resilience tests were aimed at testing the reliability of the \acrlong{fec}
+algorithm used in the radio protocol's frame layer. Those tests were considered
+critical because the frame layer must provide reliability to the
+\acrshort{ahabus} radio protocol. Since the chosen \acrshort{fec} algorithm,
+Reed-Solomon codes $RS(255,223)$, can fix a maximum of 16 bytes per 256-byte
+code block, the tests were designed to simulate average error rates below (8
+bytes out of 256), near (14 bytes out of 256) and above (18 bytes out of 256
+that theoretical limit, through the software method discussed in section
+\ref{ssec:method-resilience}.
+
+Overall, figures \ref{fig:stress-8.256}, \ref{fig:stress-14.256} and
+\ref{fig:stress-18.256} show the expected behaviour: the higher the average
+error rate, the more erroneous frames. It is interesting to note that even at
+the two lower rates ($\frac{8}{256}$ and $\frac{16}{256}$), which are below
+the maximum rate that the \acrlong{ecc} should be able to fix, some frames are
+dropped. The main cause is the use of start sequences. While \acrshort{ecc} are
+able to correct up to 16 erroneous bytes, the detection of the frame itself is a
+pre-requisite. If the frame start marker itself is corrupted, the decoder cannot
+identify the frame, and it will is counted as lost, even if the total number
+of erroneous bytes in the frame is less than 16.
+
+Additionally, it is important to note that the target error rates were averages.
+For those three tests, the code used was set to corrupt the current byte if a
+pseudo-random number between 0.0 and 1.0, obtained through the C standard
+library's \texttt{rand()} function, was above the error rate. \texttt{rand()},
+while allowing a good estimate, is not statistically accurate, and some frames
+had a higher or lower than average erroneous bytes. This is believed to be the
+source of the discrepancy between the theoretical average error rates and the
+actual correction rates shown for the first two tests in table
+\ref{tab:stress-fec}.
+
+Finally, the third tests demonstrates the inaptitude of the \acrshort{ecc}
+to cope when the average error rate is above the the algorithm's maximum. Most
+of the frames were lost, the only few decoded being just under the threshold
+due to the fluctuations discussed above. Table \ref{tab:stress-fec} shows that
+the correction rate is significantly lower than the $\frac{14}{256}$ test, which
+was expected: Reed-Solomon codes cannot fix even some errors if the number of
+erroneous bytes in a code block is above the chosen algorithm's limit.
+
+Overall, these three tests showed that the \acrshort{fec} algorithm chosen for
+the frame layer worked as expected, so long as the error rate remains under
+16 bytes per code block (frame). Ultimately, this limit is a trade-off between
+reliability and bandwidth: raising it would require more bytes in each frame
+to be allocated to \acrlong{ecc}.
+
 \section{Further Work} % << mix in with the rest?

### 2017-04-17 chapters/02-lit-review.tex
re-structure paper
Add radio discussion
Revamp Lit. Revie

@@
 
-Most \acrshort{hab} missions led by universities, schools or amateur are built 
-around commercially available small computers or micro-controller boards:
-Redland Green School students used a Raspberry Pi (\cite{rpi2014}) running
-python scripts as an \acrfull{obc} for their HAB mission
-(\cite{Hinschelwood2015}), while other projects have been ran on AVR
-micro-controller based boards like the Arduino Mega
-(\cite{AtmelCorporation2015}).
+\section{Introduction}
 
-In the United Kingdom, radio licenses do not cover airborne transmission, making
-packet radio protocols like AX.25 unusable for HAB missions 
-(\cite{ukhasradio2016}). However, the Office of Communications allows the use
-of certain \textit{license-exempt} frequencies like the 70cm wavelength band,
-provided that the transmission power does not exceed 10mW (\cite{Ofcom2014}).
-For this reason, most missions launched from the United Kingdom use the
-Radiometrix NTX2 transmitter (\cite{radiometrix2012}). Because of the
-constraints, these missions usually transmit their telemetry as text using the
-\acrfull{rtty} protocol.
+In this chapter, existing \acrlong{hab} architectures are first presented and analysed in order to identify current shortcomings.
+A comparison with equivalent systems in nano-satellites is then made to identify concepts that can be translated to \acrshort{hab} applications, before a more in-depth analysis of applicable telemetry and data transfer technologies.
 
-While most \acrshort{hab} projects make use of commercial-off-the-shelf
-components, the finished payloads are usually custom, single-purpose designs
-that are little, if at all modular. In the design document for the Titan-1 HAB
-mission, Bombasaro describes a hardware bus based on tightly coupled sensors
-which communicate over two different data buses, SPI and \acrshort{i2c}, and a
-custom-made, single-purpose flight software (\cite{Bombasaro2015}).
+\section{Existing \acrlong{hab} Architectures}
 
-Because of the risks and cost involved, nano-satellites rely more on modular
-designs, allowing each team to work on their module independently. In his paper,
-Volstad describes the design of the data bus of the NTNU CubeSat: while the OBC
-is a based on a custom made circuit board, it is designed to provide a standard
-power interface, as well as access to an \acrshort{i2c} standard bus
-(\cite{NXPSemiconductors2014}) to each payload (\cite{Volstad2011}). The
-\acrshort{i2c} protocol was chosen because of its low power consumption, and
-because if only requires two lines (clock and data). Thomas Clausen describes
-in his 2001 paper how a simple packet protocol built on top of \acrshort{i2c}
-itself is used for data transfer and error detection in Aalborg University's
-CubeSats (\cite{Clausen2001}).
+For this paper, small-budget \acrlong{hab} missions led by universities, schools or amateurs were investigated.
+More complex architectures exist, like the ones used by NASA for their long-duration, high altitude experiments (\cite{Eggers2016}).
+Such architectures were not considered because of the differences in scope, budget and complexity.
 
-Satellite missions last a lot longer than High-Altitude Balloons' and the
-requirements for radio transmissions differ: while satellites can only
-communicate when their orbit passes above a ground station and require precise
-speed and data volume planning, as described by Sandy Anthunes in his book
-(\cite{Antunes2015}), non-floating HAB missions allow for line-of-sight
-communications from liftoff to late into the descent of the payload.
+While small-budget missions are built around commercial off-the-shelf small computers, or micro-controller-based boards for their \acrlong{obc} and instruments, the resulting designs are often non-modular, combine different communications standard in order to communicate with an heterogenous set of instruments, and make use of customised, text-based telemetry formats in the United Kingdom.
 
-There are however some technologies that can be adapted from satellites to be
-used in \acrshort{hab} applications. The University of Arizona uses a custom
-packet format over 434MHz radio, containing raw binary values for each
-instrument's measurements (\cite{Eatchel2002}). The BRITE-Austria CubeSat
-mission uses the AX.25 packet radio protocol (used by amateur radio users),
-which allows the use of off-the-shelf transmission and reception hardware
-rather than custom-made circuits (\cite{Traussnig2007}).
+Redland Green School students based their design on a Raspberry Pi (ARM-based) computer (\cite{rpi2014}).
+Each software module of the project was developed on individual machines using the Python language (\cite{VanRossum2011}), before ``Each development Pi and its corresponding breadboard of components [were] combined to a single 256MB Pi'' (\cite{Hinschelwood2015}).
+While the approach was modularised, the use of commercial sensors with hard-coded communication protocols meant that moving final modules to the flight computer involved ``some code editing as different components need to use the same GPIO pins'', which introduce the potential for a single point human error.
 
-Some nano-satellites use custom telemetry data format, which allows the team
-to minimise the volume of data to be sent over radio. The Planetary Society's
-Lightsail mission has an uplink connection that allows the ground station
-to request specific logs or data. By default, The spacecraft only communicates
-a custom beacon containing a summary of the its housekeeping data
-(\cite{planetary2016}). Uplink being impractical with the transmission power
-limits imposed in the United Kingdom, such a selective telemetry system cannot
-be relied upon for \acrshort{hab} missions.
+On other projects like Emmanuel Bonbasaro's (\cite{Bombasaro2015}) Titan-1 mission, a micro-controller boards like an ``Arduino Mega 2560 is [...] used as flight computer'', alongside multiple commercial sensors connected to the \acrshort{obc} through multiple different data buses (in this case \acrfull{i2c} and \acrfull{spi}), because of the disparate communication standards in use in affordable commercial sensors.
+The flight software discussed by the author is a custom-made solution developed specifically for Titan 1's architecture, which would render its re-use in a different mission complicated.
 
-To improve collaboration and allow the deployment of large networks of
-satellites, probes and other space vehicles, the Consultative Committee for
-Space Data Systems has defined multiple communication standards that for
-the different types of networks encountered in spacecrafts.
+Transmitting data to the ground as it is captured is desirable because of the likelihood that the payload cannot be recovered.
+Outside the United Kingdom, missions often make use of amateur radio derived systems.
+Jason Winningham (\cite{Winningham2006}) provides guidelines on the use of the \acrfull{aprs} for \acrshort{hab} missions, which allows the use of an existing infrastructure that can repeat packets received by any stations and ensure that data is received even if the ballon is not in range of the tracking station (\cite{TheAPRSWorkingGroup2000}).
 
-SpaceWire defines how instruments' data can be accessed by having the OBC
-remotely poll their memory (\cite{parkes2005}). A similar system is used to
-fetch data from ROM chips over \acrshort{i2c} and could be used to build a
-simple data bus.
+Amateur radio licenses, to which the use of APRS is linked, are restricted in the United Kingdom by the Office of Communications and licensees ``shall not establish or use the Radio Equipment in any Aircraft or other Airborne Vehicle.
+[...] ``Aircraft'' and ``Airborne Vehicles'' includes full size and models and also includes balloons whether tethered or free'' (\cite{Ofcom2017}).
+However, the use of certain \textit{license-exempt} frequencies like the 70cm wavelength band, provided that the transmission power does not exceed 10mW (\cite{Ofcom2014}).
+For this reason, most missions launched from the United Kingdom make use of the Radiometrix NTX2 transmitter (\cite{radiometrix2012}), coupled with a de-facto telemetry standard using plain-text sentences, sent over low-speed  \acrfull{rtty} links to minimise losses due to low signal-to-noise ratio (\cite{UKHAS2015}).
 
-The \acrshort{ccsds} Space Packet protocol defines a packet format used to
-encapsulate application data (from instruments for example) that can be sent
-over a ``Space Link'', an analogue of the OSI model's link layer
-(\cite{Stallings1987}) using \acrshort{ccsds} frames to encapsulate packets
-originating from multiple instruments, ground stations and spacecrafts
-(\cite{ccsds2003}).
+\section{Satellites Bus Architectures}
 
-While \acrshort{hab} applications do not require the level of complexity of
-\acrshort{ccsds} protocols — addressing is not needed since the network only
-contains two endpoints, communicating in a single direction — some details of
-the Space Packet protocol are worth reusing (Forward Error Correction,
-different Application IDs for each payload).
+When space conditions are required, or when the targeted scientific observations require longer or more global exposure times, nano-satellites have become popular over the last decade ( \cite{McDowell2017a}, \cite{McDowell2017}) as a relatively affordable – though much more costly than \acrshort{hab} missions – access to space.
+The most popular standard is CubeSat (\cite{Heidt2000}), based on \SI{10}{\centi\metre} cubes small and light enough to be launched as secondary payloads, and thus only require a small portion of the launch costs to be covered by nano-satellites teams.
 
-\section{Analysis of the Problem}
+Because of the risks and cost involved, nano-satellites rely more on modular designs, allowing each team to work on their module independently.
+In his paper, Marius Volstad describes the design of the data bus of the NTNU CubeSat.
+While the \acrlong{obc} is a based on a custom-made circuit board, it is designed to provide a standard power interface to each payload, as well as access to an \acrshort{i2c} standard bus (\cite{NXPSemiconductors2014}) to each payload (\cite{Volstad2011}).
+The \acrshort{i2c} protocol was chosen because of its low power consumption, and because if only requires two lines (clock and data).
+
+Thomas Clausen (\cite{Clausen2001}) describes in his 2001 paper the design of an internal bus for CubeSats based on the \acrshort{i2c} protocol as well.
+In this case too, \acrshort{i2c} was chosen for its low power consumption and I/O line usage, but also for its speed and reliance to noise, and because the addressing of payloads on the bus can be done entirely in software.
+T.
+Clausen proposes the use of a limited packet system on top of \acrshort{i2c} in order to provide stable data transfers between payloads and the \acrlong{obc}.
+
+These modular designs are taken further in full-scale satellites and exploration probes.
+Lockheed Martin's 1999 patent (\cite{Benedetti1999}) describes a system that provides a modular power, propulsion and and data management platform on which payloads are connected to simplify satellite and probe development.
+In order improve collaboration between divisions and improve the reusability of software and hardware modules, the \acrfull{ccsds} and \acrfull{esa} have designed SpaceWire, a series of standards for satellites' internal computer and network systems.
+The SpaceWire Remote Access Protocol (\cite{parkes2005}) proposes a simple model in which instruments' data is accessed by the \acrlong{obc} by addressing each instrument like random-access memory cells.
+A similar system is used to fetch data from ROM chips over \acrshort{i2c} and could be used to build a simple data bus.
+
+\section{Satellite Radio Technologies}
+
+Satellite missions last a lot longer than High-Altitude Balloons' and the requirements for radio transmissions differ.
+While satellites can only communicate when their orbit passes above a ground station and require precise speed and data volume planning, as described by Sandy Anthunes in his book (\cite{Antunes2015}), non-floating HAB missions normally allow for line-of-sight communications from liftoff to late into the descent of the payload.
+
+Furthermore, numerous nano-satellites use custom telemetry data format, which allows the team to minimise the volume of data to be sent over radio.
+The Planetary Society's Lightsail mission (\cite{planetary2016}) uses an uplink connection that allows the ground station to request specific logs or data.
+By default, The spacecraft only communicates a custom beacon containing a summary of the its housekeeping data.
+Uplink being impractical with the transmission power limits imposed in the United Kingdom, such a selective telemetry system cannot be relied upon for HAB missions.
+
+There are however some technologies that can be adapted from satellites to be used in HAB applications.
+In a 2002 paper (\cite{Eatchel2002}), the telemetry standard for the CubeSat programmes of the University of Arizona are presented as a custom binary packet format.
+Binary data obtained from each instrument is encapsulated in packets of variable length (between 0 and 256 bytes), preceded by a header ``[containing] a six byte call sign and a satellite ID code'' and followed by a CRC16 checksum for error detection.
+Conversely, the BRITE-Austria CubeSat mission described by Wolfgang Traussnig (\cite{Traussnig2007}) uses the existing AX.25 packet radio standard as its link layer, and proposes the use of the \acrshort{ccsds} frame and packet protocols on top of it.
+
+The \acrshort{ccsds} Space Packet protocol defines a packet format used to encapsulate application data (from instruments for example) that can be sent over a ``Space Link'', an analogue of the OSI model's link layer (\cite{Stallings1987}) using \acrshort{ccsds} frames to encapsulate packets originating from multiple instruments, ground stations and spacecrafts (\cite{ccsds2003}).
+While \acrlong{hab} applications do not require the level of complexity of \acrshort{ccsds} protocols — addressing is not needed since the network only contains two endpoints, communicating in a single direction — some details of the Space Packet protocol are worth reusing (\acrfull{fec}, different Application IDs for each payload).
+
+% bit about FEC? Real-Time OS
+
+\section{Summary}
+
+As discussed in this chapter, the current state of the art of low-budget \acrlong{hab} missions relies on single-purpose, custom hardware and software designs that cannot be easily reused across missions and payload configurations.
+In the following chapter, a solution based on the technologies and architectures used in related but less affordable satellite systems is presented and analysed.

### 2017-04-17 chapters/05-discussion.tex
Finish discussion
Revamp Lit. Revie

@@
 results of tests discussed in chapter \ref{ch:results}, then through more
-general qualitative observation and analysis. This chapter finally presents the
-further work that would be required for the platform to fully meet its design
-goals, leading to potential test flights.
+general qualitative observation and analysis.%  This chapter finally presents the
+% further work that would be required for the platform to fully meet its design
+% goals, leading to potential test flights.
 
@@
 to be allocated to \acrlong{ecc}.
-
-\section{Further Work} % << mix in with the rest?

### 2017-04-17 paper.tex
Better table styles
Better data parsing
Start of discussion
Add ground segment discussion
Change bus status semantics
Add radio discussion
Finish discussion
Revamp Lit. Revie

@@
 \usepackage{pgfplots}
+\usepackage{pgfplotstable}
 \usepackage{csvsimple}
+\usepackage{hyperref}
+
+\usepackage{mathptmx}
+
 \usepackage[detect-weight=true, binary-units=true]{siunitx}
@@
 \usepackage[T1]{fontenc}
-\usepackage[a4paper, margin=2.54cm]{geometry}
+\usepackage[a4paper, margin=3cm]{geometry}
 \usepackage[nottoc]{tocbibind}
+\usepackage{array}
 
@@
 \usepackage[
-style=authoryear,
-citestyle=authoryear,
+style=authoryear-comp,
+citestyle=authoryear-comp,
 backend=biber,
@@
 \author{Amy Parent}
-\title{AHABus \& FCORE:\\Modular Bus for High-Altitude Balloon Missions}
+\title{AHABus \& FCORE: A Modular Bus System for High-Altitude Balloon Missions}
 \date{April 2017}
@@
 \setlength{\baselineskip}{1.5em}
+\setcounter{tocdepth}{4}
+\setcounter{secnumdepth}{4}
 
@@
 \cleardoublepage
-    \pagenumbering{roman}}
+\pagenumbering{roman}
+}
 
@@
 \cleardoublepage
-    \pagenumbering{arabic}}
+\pagenumbering{arabic}
+}
 

### 2017-04-18 appendices.tex
First draft, minus photos

\begin{appendices}
%\addtocontents{toc}{\protect\setcounter{tocdepth}{0}}

\chapter{Source Code Repositories}
\label{apx:source-repos}

All source code and documentation for the \acrfull{ahabus} platform and, its flight software \acrfull{fcore} and the ground segment software are made available and grouped under the AHABus GitHub organisation, at https://github.com/ahabus. Below is a list of the different repositories:

\begin{description}
\item[\acrshort{fcore}] main flight software, based on FreeRTOS.\\https://github.com/ahabus/fcore

\item[Dl-Fldigi] Fork of Dl-Fldigi able to export raw RTTY binary streams.\\https://github.com/ahabus/dl-fldigi

\item[PacketRadio] AHABus Radio Protocol encoder/decoder reference implementation.\\https://github.com/AHABus/packet-radio
\end{description}

\chapter{Test Data Repositories}
\label{apx:test-data-repos}

The data collected during the testing phase is made available under the AHABus GitHub organisation. Below is a list of the different repositories:

\begin{description}
\item[Raw Data] https://github.com/ahabus/test-rawdata

\item[Data Processing Scripts] https://github.com/ahabus/test-processing
\end{description}

\end{appendices}

### 2017-04-18 chapters/01-introduction.tex
First draft, minus photos

\chapter{Introduction}
\label{ch:introduction}

Nano-satellites have lowered the cost of access to orbit by allowing light payloads to be launched as secondary passengers on heavy-lift rocket flights. However, the cost of the hardware required and the launch itself — when it is not subsidised by the launch provider or space agencies — still limit their use  to well-funded universities and companies.

For applications that require low atmospheric density or medium altitudes, but do not depend on the payload being in a \acrfull{leo}, \acrfull{hab} present an affordable alternative. Such balloons, inflated with low-density gases (usually helium or hydrogen), can reach the stratosphere (altitudes of 30 to 45km). The payload is encased in a thermally insulated container attached under the balloon itself. Depending on the mission, balloons either float at a constant altitude and drift for long durations or climb to the target altitude, then burst under the pressure of the expanding gas. Once the balloon is destroyed, the payload is brought back to the ground under a parachute.

The main component of the payload is the \acrfull{obc}, which controls the different instruments through the different phases of flight (ascent, observation, free-fall and parachute descent). Because high-altitude winds can reach speeds upwards of 150km/h and carry the payload over long distances, most balloons carry a \acrfull{gnss} receiver that the OBC can use to keep track of the balloon's position and altitude. To avoid loosing flight data if the payload is lost or lands in an unaccessible area, the OBC also controls a radio transmission system used to forward payload data and metadata (\acrshort{gnss} coordinates, altitude, systems health) to a ground station as it is acquired.

Depending on the mission, various instruments are added to the payload. The \acrshort{obc} is in charge of gathering, storing and transmitting their data to the ground. Most balloon payloads are designed as custom hardware and software to fit the mission, which means common systems like \acrshort{gnss}, on-board data transfer and radio communications are re-implemented every time.

The satellite industry solves this problem by making use of buses for most missions. Instead of designing a custom, monolithic platform for each new satellite, manufacturers sell buses that provide utility (power, data bus, attitude control and space-to-ground communications) to which the customer or science team connect mission-specific payloads.

\begin{figure}[H]
\includegraphics[width=0.35\textwidth]{mcsb.png}
\centering
\caption{LADEE Spacecraft, based on the Modular Common Spacecraft Bus (NASA
Ames)}
\end{figure}

Such modular designs allow mission specialists and scientists to focus on their experiments and payloads, and help reduce cost by spreading common development costs over many missions. Bringing their use to \acrshort{hab} missions would make high-altitude research accessible to a wider range of users, by taking away the computer science and electrical engineering required to design and build the flight computer and radio telemetry system.
\vskip 0.5cm
What are the advantages, issues and obstacles in designing a modular intra-payload and long-range communications platform for \acrlong{hab} scientific missions? The aim of the project presented in this paper was to design, document, implement the protocols, ground and flight software of a low-cost, modular platform to provide data transfer and long-range radio telemetry to payloads carried on \acrlong{hab} missions, and evaluate the prototype's performance. The objectives of the project were:

\begin{itemize}
\item Evaluate existing platforms and protocols used for communications between
payloads and over radio in high-altitude balloons and nano-satellites.

\item Evaluate the feasibility of a modular \acrlong{hab} bus based on low-cost, open-source, off-the-shelf hardware.

\item Design the required protocols and build a prototype for an open-source high-altitude balloon \acrlong{obc} and communication platform.

\item Evaluate the performance of the prototype's payload data bus and radio telemetry system.

\item Evaluate the robustness, maintainability and ease of use of the bus' software compared to that of single-purpose payload firmwares.
\end{itemize}

The project was focused on the software design and implementation of the platform first, and the hardware prototype design second, depending on available resources and time. To limit the scope, the electrical engineering aspect of the proposed platform was not considered past what was required to operate the development boards and the prototype platform.

In the following chapters, the current state of \acrlong{hab} and nano-satellite mission architectures is discussed. The proposed platform is then presented, followed by the results of the evaluation, and finally a critical discussion of those results and the overall performance of the platform.

### 2017-04-18 chapters/03-methodology.tex
Add radio discussion
Add bus discussion
First draft, minus photos

@@
 \label{ch:methodology}
+
+\section{Introduction}
+
+In this chapter, the \acrfull{ahabus} is presented as a prototype solution to the issues exposed in chapter \ref{ch:lit-review}.
 
@@
 only generate data and make it available to the platform.
+A global overview of the platform and the environment used for its development is presented, followed by a detailed discussion of the development process of each of the four main components of \acrshort{ahabus} (Radio Communications, Payload Bus, Flight Software and Ground Segment).
+Finally, the methods used to evaluate the viability of the prototype platform are presented, leading to the results presented in chapter \ref{ch:results}.
+
+The source code and protocol documentation resulting from the development process discussed in this chapter are available in the project's repositories listed in appendix \ref{apx:source-repos}.
 
@@
 
-The proposed solution was designed to run on low-cost embedded platforms, and
-the prototype was built around the Espressif ESP8266 micro-controller
-(\acrlong{obc}). The radio link used a Radiometrix NTX2B transmitter
-(\cite{radiometrix2012}), and an unnamed Software-Defined Radio receiver and
-antenna.
+The proposed solution was designed to run on low-cost embedded platforms, and the prototype was built around the Espressif ESP8266 (\cite{EspressifInc.2017}) micro-controller (\acrlong{obc}).
+The radio link used a Radiometrix NTX2B transmitter (\cite{radiometrix2012}), and an unnamed Software-Defined Radio receiver and antenna.
 
@@
 for multiplexing data from different payloads.
-
-\subsection{Hardware Setup}
+%
+% \subsection{Hardware Setup}
+%
+% The radio tra
 
@@
 \item The payload sends its data to the \acrshort{obc}.
-    \item The \acrshort{obc} release the payload's $\overline{CE}$ line to high.
+\item The \acrshort{obc} releases the payload's $\overline{CE}$ line to high.
 \item The payload releases its $\overline{CR}$ line to high.
@@
 the payload's priority level, as discussed in more details in section
-\ref{sssec:conf-file}. The task establishes a connection with the payload using
+\ref{ssec:conf-file}. The task establishes a connection with the payload using
 the \acrshort{ahabus} payload bus protocol. If data is available, it is
@@
 
-\[B = F_S + n_{1}F_{P} + \frac{n_{2}F_{P}}{2} + \frac{n_{3}F_{P}}{3} \]
+\[B = F_S + n_{1}F_{P} + n_{2}\frac{F_{P}}{2} + n_{3}\frac{F_{P}}{3} \]
 

### 2017-04-18 chapters/04-results.tex
Add stress tests discussion
First draft, minus photos

@@
 during the tests done as part of the evaluation process of the project, as
-discussed in section \ref{sec:testing}.
+discussed in section \ref{sec:testing}. The raw data files from which the
+results presented in this chapter are derived, as well as the Python scripts
+used to process them, are available in the project's repositories listed in
+appendix \ref{apx:test-data-repos}.
 
@@
 \caption {DITL Test - Received packets}
-\label{tab:ditl-health}
+\label{tab:ditl-packet-count}
 \end{table}
@@
 some frames can be corrected, while some where the error rate is higher than
-the average are lost. The fluctuations in actual error rate are due to the use
-of the C language's \texttt{rand()} function. %remote or explain
+the average are lost.
 

### 2017-04-18 chapters/05-discussion.tex
First draft, minus photos

@@
 through a low-power radio channel. In this chapter, the \acrshort{ahabus}
-platform prototype is evaluated against these aims, first by analysing the
-results of tests discussed in chapter \ref{ch:results}, then through more
-general qualitative observation and analysis.%  This chapter finally presents the
-% further work that would be required for the platform to fully meet its design
-% goals, leading to potential test flights.
+platform prototype is evaluated against these aims, firs through general 
+qualitative observations and analysis of the prototype's behaviour, then by
+analysing the results of tests discussed in chapter \ref{ch:results}.
 
@@
 during the testing phase of the project. The results are presented in chapter
-\ref{ch:results}, and the raw data files obtained from each tests are included
-in appendices \ref{} through \ref{}, or on the project's GitHub account at
-https://github.com/ahabus/results
+\ref{ch:results}, and the raw data files obtained from each tests are made available in the repositories specified in appendix \ref{apx:test-data-repos}.
 

### 2017-04-18 chapters/06-conclusion.tex
First draft, minus photos

\chapter{Conclusion}
\label{ch:conclusion}

In summary, the project discussed in this paper was successful. The main goal of the project, investigating the feasibility of a modular bus system for \acrlong{hab} missions, was reached, and a working prototype was successfully designed, implemented and tested.

Overall, the modular bus architecture found in most modern satellites and deep-space probes was found to be portable to \acrlong{hab} applications: no major issues were found  in the \acrshort{ahabus} prototype, and the platform performed well enough in its first round of testing that the concept could be taken further. While the associated tooling would require more work to make it more accessible and easier to use, an \acrshort{ahabus}-based mission could be designed and prepared using the platform in its current state, without requiring payload team to edit the flight software's source code.

As discussed in chapter \ref{ch:discussion}, the \acrshort{ahabus} platform in its current state, while well suited for scientific missions where data is collected at low, steady rates, would not perform as well in simpler missions focused on collecting high-volume data at long intervals such as pictures.

Further work will be required to qualify the \acrshort{ahabus} platform for flight missions. More investigation will be required to identify the root cause, and mitigate the payload bus protocol issues discussed in section \ref{ssec:discussion-bus}. Advanced testing of the radio system must be undertaken, specifically range testing, as well as line-of-sight and interference testing to qualify the behaviour of the radio protocol and software low signal-to-noise ration environments. Power consumption testing would be beneficial so that the current draw of the multi-payload, modular platform can be compared to more common custom-designed, single purpose mission architectures. Finally, test missions will have to be flown, possibly using an \acrshort{ahabus} platform as a passive payload flying alongside a simple, well tested radio and \acrshort{gnss} beacon for backup.

### 2017-04-18 paper.tex
First draft, minus photos

@@
 \documentclass[11pt]{report}
+\usepackage[scaled]{beramono} % BitStream Vera font (code)
+\usepackage[T1]{fontenc}
+\usepackage[a4paper, margin=3cm]{geometry}
 \usepackage[utf8]{inputenc}
@@
 \usepackage[justification=centering,labelfont={small,bf},font={small,bf}]{caption}
-\usepackage[scaled]{beramono} % BitStream Vera font (code)
-\usepackage[T1]{fontenc}
-\usepackage[a4paper, margin=3cm]{geometry}
+\usepackage[titletoc,toc,page]{appendix}
 \usepackage[nottoc]{tocbibind}
@@
 \begin{abstract}
+
+The aim of this project was to investigate the design options for a modular payload bus for \acrfull{hab} missions, and to evaluate one such design by implementing in in hardware and software.
+
+\acrlong{hab} provide affordable access to near-space conditions, even for low-budget entities (universities, schools and amateurs).
+While current low-budget missions rely on commercial off-the-shelf computers and sensors, their architecture and software are often custom-made and targeted for a single use.
+This prevents easy reuse of hardware and software components, and raises the barrier to entry by requiring non-trivial network and computing skills.
+
+This paper presents the development of \acrfull{ahabus}, solution designed by applying the modular bus model used by most modern satellites to \acrlong{hab} architectures.
+A common, generic \acrfull{obc} running a common flight software (\acrshort{fcore}) provides facilities for position tracking and reliable radio telemetry to a ground station, and polls data from payloads, custom for each flight.
+
+The resulting protocols and platform prototype were successful and passed both day-in-the-life testing and radio channel resilience tests.
+Further testing of the and minor modifications to the Payload Bus and Radio Communications protocols will be required in order to qualify the platform for in-flight use.
+Once these steps are taken, \acrshort{ahabus} could render scientific \acrlong{hab} missions more accessible and affordable by allowing students and researchers to focus on payload design and development.
+
 \end{abstract}
@@
 
+\input{appendices.tex}
+
 \end{document}

### 2017-04-18 chapters/03-methodology.tex
First draft

@@
 
-In this chapter, the \acrfull{ahabus} is presented as a prototype solution to the issues exposed in chapter \ref{ch:lit-review}.
+In this chapter, the \acrfull{ahabus} is presented as a prototype solution to the issues exposed in chapter \ref{ch:literature-review}.
+A global overview of the platform and the environment used for its development is presented, followed by a detailed discussion of the development process of each of the four main components of \acrshort{ahabus} (Radio Communications, Payload Bus, Flight Software and Ground Segment).
+Finally, the methods used to evaluate the viability of the prototype platform are presented, leading to the results presented in chapter \ref{ch:results}.
+
+The source code and protocol documentation resulting from the development process discussed in this chapter are available in the project's repositories listed in appendix \ref{apx:source-repos}.
 
@@
 and hardware platform provides common facilities required on all High-Altitude
-flights – position tracking, internal data transfers and radio telemetry to the
-ground station – allowing payloads (scientific instruments for example) to
+flights – position tracking, internal data transfers and radio telemetry to
+the ground station – allowing payloads (scientific instruments for example) to
 only generate data and make it available to the platform.
-A global overview of the platform and the environment used for its development is presented, followed by a detailed discussion of the development process of each of the four main components of \acrshort{ahabus} (Radio Communications, Payload Bus, Flight Software and Ground Segment).
-Finally, the methods used to evaluate the viability of the prototype platform are presented, leading to the results presented in chapter \ref{ch:results}.
-
-The source code and protocol documentation resulting from the development process discussed in this chapter are available in the project's repositories listed in appendix \ref{apx:source-repos}.
 
@@
 This section describes the design process that led to the radio communications
-protocol used by the AHABus platform to send payload data and flight metadata
+protocol used by the \acrshort{ahabus} platform to send payload data and flight metadata
 to the ground station. Its role is to ensure that arbitrary binary data – each
@@
 
-\subsection{AHABus Radio Protocol}
-
-The AHABus radio protocol was designed to match the requirements for the radio
+\subsection{\acrshort{ahabus} Radio Protocol}
+
+The \acrshort{ahabus} radio protocol was designed to match the requirements for the radio
 system as closely as possible. While the specification presented in this section
-is the last version, which was implemented in the AHABus prototype, some parts
+is the last version, which was implemented in the \acrshort{ahabus} prototype, some parts
 of the protocol evolved or had to be changed during the prototyping process.
@@
 like DominoEx or Olivia, sometimes with better transmission rates. RTTY was
-preferred over them because the ESP8266 board used as OBC in the AHABus
+preferred over them because the ESP8266 board used as OBC in the \acrshort{ahabus}
 prototype provides hardware support for UART, making it easier to implement and
@@
 be expected, there is no model to which actual received data can be compared.
-For this reason, the AHABus radio protocol was designed as a framed protocol:
+For this reason, the \acrshort{ahabus} radio protocol was designed as a framed protocol:
 the stream of data is divided in chunks of a set number of bytes, encapsulated
@@
 
-The AHABus frame design was inspired by other frame-based protocols like the
+The \acrshort{ahabus} frame design was inspired by other frame-based protocols like the
 CCSDS Space Packet/Frame protocol. Each frame is composed of two section: a
@@
     \centering
-    \caption{AHABus Frame Format – Version 1}
+    \caption{\acrshort{ahabus} Frame Format – Version 1}
     \label{fig:frame-fmt-orig}
@@
     \centering
-    \caption{AHABus Frame Format – Version 2}
+    \caption{\acrshort{ahabus} Frame Format – Version 2}
     \label{fig:frame-fmt-2}
@@
     \centering
-    \caption{AHABus Frame Synchronisation State Machine}
+    \caption{\acrshort{ahabus} Frame Synchronisation State Machine}
     \label{fig:frame-fsm}
@@
 Figure \ref{fig:frame-fsm} presents the finite state machine that
-AHABus frame decoders must implement to detect frames in a binary
+\acrshort{ahabus} frame decoders must implement to detect frames in a binary
 stream, and figure \ref{fig:frame-fmt-3} the final version of the
-AHABus frame format.
+\acrshort{ahabus} frame format.
 
@@
 percentage \(\frac{n}{n+m}\) of the stream contains actual data. For reasons
-explained further, the AHABus frames were chosen to be 256-byte
+explained further, the \acrshort{ahabus} frames were chosen to be 256-byte
 long. As shown previously, each frame also includes.
@@
 
-To achieve those goals, the AHABus protocol uses packets similar to frames. Each
+To achieve those goals, the \acrshort{ahabus} protocol uses packets similar to frames. Each
 packet encapsulates a chunk of data retrieved from a single payload. To that
@@
 information necessary to attribute the packet's data to a payload, indicate
-the length of the packet, and carry the AHABus platform's location and altitude.
+the length of the packet, and carry the \acrshort{ahabus} platform's location and altitude.
 
@@
     \centering
-    \caption{AHABus Packet Format - Version 1}
+    \caption{\acrshort{ahabus} Packet Format - Version 1}
     \label{fig:packet-fmt-1}
@@
 
-Figure \ref{fig:packet-fmt-1} shows the first and current version of the AHABus
+Figure \ref{fig:packet-fmt-1} shows the first and current version of the \acrshort{ahabus}
 packet format. Like frame discussed in section \ref{sssec:frame-layer}, each
@@
 finding the start of the next packet would have been complicated, or required
-a synchronisation mechanism similar to the one designed for AHABus frames.
+a synchronisation mechanism similar to the one designed for \acrshort{ahabus} frames.
 Aligning packets instead makes such a system unnecessary, reducing the packet
@@
 
-To evaluate the protocol's viability, a full demonstration of an AHABus radio
+To evaluate the protocol's viability, a full demonstration of an \acrshort{ahabus} radio
 stack was written. Its main goals were to provide a reference implementation of
@@
 implementation because of its widespread use, and its availability on most
-micro-controller and embedded platforms, which would be used as AHABus flight
+micro-controller and embedded platforms, which would be used as \acrshort{ahabus} flight
 platforms. To render the code as portable and reusable as possible, care was
@@
 
-\subsection{AHABus Payload Bus Protocol}
+\subsection{\acrshort{ahabus} Payload Bus Protocol}
 
@@
 
-The AHABus Payload Bus protocol was designed to work with a virtual register
+The \acrshort{ahabus} Payload Bus protocol was designed to work with a virtual register
 system: each payload can be thought of like a memory array, and the bus
@@
 \end{center}
-\caption {AHABus Payload Bus Protocol - Virtual Registers}
+\caption {\acrshort{ahabus} Payload Bus Protocol - Virtual Registers}
 \label{tab:bus-protocol-registers}
@@
 \centering
-\caption{AHABus - Electronic Circuit}
+\caption{\acrshort{ahabus} - Electronic Circuit}
 \label{fig:ahabus-schema}
@@
 \centering
-\caption{AHABus - Printed Circuit Board Layout}
+\caption{\acrshort{ahabus} - Printed Circuit Board Layout}
 \label{fig:ahabus-pcb}
@@
 most programming languages provide support for network communications, which
-means that tools could easily be written to work with AHABus frame streams, such
+means that tools could easily be written to work with \acrshort{ahabus} frame streams, such
 as the packet parser discussed in the following section.
@@
 
+\begin{figure}[H]
+\includegraphics[width=0.40\textwidth]{dl-fldigi.png}
+\centering
+\caption{\acrshort{ahabus} - Ground Segment Software}
+\end{figure}
+
 \section{Evaluation}
@@
 
-For all tests, the experimental setup was kept consistent: an ESP8266
-development board was used as the \acrshort{obc}, while two Arduino Uno boards
-were connected as payloads, generating fake instrument data. A \SI{5}{\volt}
-power supply was connected to the power lines.
+For all tests, the experimental setup, as shown in figure \ref{fig:exp-setup} was kept consistent: an ESP8266 development board was used as the \acrshort{obc}, while two Arduino Uno boards were connected as payloads, generating fake instrument data.
+A \SI{5}{\volt} power supply was connected to the power lines.
+
+\begin{figure}[H]
+\includegraphics[width=0.40\textwidth]{test-rig.jpg}
+\centering
+\caption{\acrshort{ahabus} - Test Setup}
+\label{fig:exp-setup}
+\end{figure}
 

### 2017-04-23 appendices.tex
Add specs in appendices, mention of Arduino Bus Library

@@
 \begin{appendices}
-%\addtocontents{toc}{\protect\setcounter{tocdepth}{0}}
+\addtocontents{toc}{\protect\setcounter{tocdepth}{0}}
 
@@
 
+\item[radio-parser] A simple packet parser that decodes incoming \acrshort{ahabus} frame streams and displays packet contents.\\https://github.com/AHABus/radio-parser
+
 \item[PacketRadio] AHABus Radio Protocol encoder/decoder reference implementation.\\https://github.com/AHABus/packet-radio
+
+\item[Arduino-Payload] Example implementation of a \acrshort{ahabus} payload for Arduino-compatible micro-controllers.\\https://github.com/AHABus/arduino-payload
 \end{description}
@@
 
+\chapter{AHABus Packet Radio Protocol Specification}\label{apx:protocol-radio}
+
+\begin{verbatim}
+author:     Amy Parent <amy@amyparrent.com>
+version:    3
+date:       2017-02-15
+\end{verbatim}
+
+This document describes the protocol used by the AHABus platform (or any
+other compliant bus) to transfer telemetry data from an airborne
+platform (typically, a High-Altitude Balloon payload) to a ground
+station over a radio link.
+
+\section{Protocol Overview}\label{protocol-overview}
+
+The protocol is designed with low-bandwidth, low-power radio links in
+mind, to stay within the legal limits for airborne transmission in the
+UK {[}1{]}: 70cm wavelength band transmissions at a power no higher than
+10~mW.
+
+The model used by low-cost High Altitude Balloon missions (one way,
+two-point non multiplexed RTTY link), there is no real network to speak
+of. For this reason, the OSI layers model {[}2{]} is hardly applicable
+to the AHABus protocol.
+
+None the less, the protocol follows a layer model, described here in
+increasing levels of abstraction: the application-level packets are
+encapsulated in frames, which provide order and error correction. Frames
+themselves are sent over the RTTY radio link, which uses Frequency-Shift
+Keying to convert a digital stream of bits to an analogue radio signal.
+
+\section{Requirements}\label{requirements}
+
+The AHABus Packet Protocol's main goal is to support telemetry
+transmission from an AHABus-based High Altitude Balloon mission to a
+ground station. The following requirements all come from this main goal:
+
+\begin{itemize}
+\item
+  provide a system to transmit multiple payload's data on a same link
+\item
+  provide stable transmissions on a low-power link
+\item
+  provide some level of transmission error correction
+\item
+  provide means to track the payload's GNSS location over time
+\end{itemize}
+
+\section{Layers Specification}\label{layers-specification}
+
+The three main layers of the AHABus Radio Protocol are meant to be as
+independent from each other as possible. The rationale behind that is
+that one layer could be swapped for a different implementation without
+impacting the design of the other layers.
+
+On every layer above the data-link layer, any multi-octet field must be
+transmitted in network byte order (little-endian). Integers must be
+transmitted in two's-complement binary format, floating point values in
+IEEE754 single precision. Values sent as fixed-point have their
+fractional and integer bit width specified when required.
+
+\subsection{Radio Link}\label{radio-link}
+
+The lowest-level layer of the AHABus Radio Protocol is the radio link.
+Its only role is to provide a raw binary stream between the payload and
+the ground station, without consideration for the transmitted data's
+structure.
+
+Because of the restrictions imposed on airborne radio transmission
+equipment in the United Kingdom {[}1{]}, the chosen physical link is:
+
+\begin{verbatim}
+frequency:      434MHz band
+power:          < 10mW
+Modulation:     FSK/AFSK
+
+Encoding:       RTTY (8bit-bytes, No parity bit, 2 stop bits)
+Rate:           50-300bauds
+\end{verbatim}
+
+These are guidelines only, and can be swapped for other types of
+modulation and encoding where it is legal to do so, and as long as the
+chosen encoding can provide a binary stream.
+
+\subsection{Frame Layer}\label{fec-layer}
+
+The FEC layer provides transfer reliability to the {[}over{]} layers.
+Any data generated by upper layers is split into frames at the FEC
+layer. Each frame is composed of a frame sync marker, a header, the
+binary data and a Forward Error Correction code.
+
+The header contains the frame's AHABus protocol version and a sequence
+number which allows a partial frame stream to be re-constructed on the
+ground in the case a frame is lost in transmission.
+
+The last 32 octets of each frame make up the frame's Reed-Solomon
+(255,223) {[}3{]} FEC code, used to correct data corruptions on
+reception of each frame. The code is computed using the complete frame,
+sync marker and header included.
+
+A frame should start to be decoded when the incoming byte in the streams
+goes from \texttt{0xAA} to \texttt{0x5A}. There can be any number of
+sync bytes (\texttt{0xAA}). The frame marker (\texttt{0x5A}) is part of
+the frame's 256 bytes, but is not counted when computing the
+Reed-Solomon error-correction code for the frame.
+
+\textbf{AHABus Frame Structure}
+
+\begin{verbatim}
+struct radio_frame {
+    00: u8          start_marker = 0x5A
+    01: u8          protocol_version
+    02: u16         sequence_number
+    04: b8[220]     data
+    e0: b8[32]      fec_code
+}
+\end{verbatim}
+
+\subsection{Packet Layer}\label{application-layer}
+
+The application layer provides the support for the high-level AHABus
+operations: packets encapsulate both each instruments' data samples, but
+also ancillary data that is required to support a High-Altitude Balloon
+mission. Packets also allow multiple instruments' data to be sent over a
+shared connection.
+
+Each packet is composed of a primary and second header, and a body of
+variable length.
+
+\begin{itemize}
+\item
+  The primary header contains data required for the packet decoder to
+  parse and dispatch the packet's data correctly:
+
+  \begin{itemize}
+  \item
+    AHABus protocol version,
+  \item
+    Payload ID of the instruments who's data is carried in the packet,
+  \item
+    Packet's length in bytes, including the header.
+  \end{itemize}
+\item
+  The secondary header contains ancillary data required for the support
+  of the mission:
+
+  \begin{itemize}
+  \item
+    AHABus platform's latitude in decimal format at the time the packet
+    was encoded, in (8.24) fixed point format,
+  \item
+    AHABus platform's longitude in decimal format at the time the packet
+    was encoded, in (8.24) fixed point format,
+  \item
+    AHABus platform's altitude in metres at the time the packet was
+    encoded, in 16-bit unsigned integer format.
+  \end{itemize}
+\end{itemize}
+
+Packets transmitted over AHABus frame streams should be aligned on frame
+boundaries. Each packet should start on a new frame, and the end of a
+packet should be the last data in a frame.
+
+\textbf{AHABus Packet Structure}
+
+\begin{verbatim}
+struct radio_packet {
+    00: u8          protocol_version
+    01: u8          instrument_id
+    04: u16         length
+    06: f32         latitude
+    08: f32         longitude
+    0a: u16         altitude
+    0c: b8[length]  data
+}
+\end{verbatim}
+
+\section{Related Documents}\label{related-documents}
+
+\begin{itemize}
+\item
+  \href{https://github.com/AHABus/src/software/payload-bus.md}{AHABus
+  Payload Bus Protocol Specification}
+\end{itemize}
+
+\section{References}\label{references}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Ofcom (2014). IR 2030 -
+  \href{https://www.ofcom.org.uk/__data/assets/pdf_file/0028/84970/ir_2030-june2014.pdf}{UK
+  Interface Requirements 2030 Licence Exempt Short Range Devices}.
+  London
+\item
+  Stallings, W. (1987). Handbook of computer- communications standards.
+  Macmillan. isbn: 002948071X.
+\item
+  UKHAS (2010). \href{http://ukhas.org.uk/code:rs8encode}{Reed-Solomon
+  Encoder (255,223)}.
+\end{enumerate}
+
+\chapter{AHABus Payload Bus Protocol}\label{apx:protocol-bus}
+
+\begin{verbatim}
+author:     Amy Parent <amy@amyparrent.com>
+version:    3
+date:       2017-02-15
+\end{verbatim}
+
+This document describes the protocol used by the AHABus platform (or any
+other compliant bus) to communicate and fetch data from payload
+instruments hosted on an FCORE data bus.
+
+\section{Protocol Overview}\label{protocol-overview}
+
+The AHABus protocol (AHAP) is designed to provide an asymmetric
+communication platform between a High-Altitude Balloon mission's
+instruments and the main flight computer (FCORE).
+
+AHAP relies on the two-wire (Phillips I2C) protocol to establish a
+connection between the main computer and each payload. Communications
+are controlled solely by the main computer: a payload can only send or
+receive data when it has been addressed.
+
+Because the AHABus platform relies on low-bandwidth data, data rate caps
+can be enforced on each payload. These caps are defined on a per-mission
+basis. If a payload returns more data than it is allowed to the computer
+will stop addressing it.
+
+FCORE (the software component of the AHABus flight computer) will
+address each payload in turn over a certain period of time. By default,
+each payload has the same data rate restrictions. If priorities are
+given to payloads, higher priority payloads will be allowed higher data
+caps than their low-priority counterparts.
+
+AHAP is a virtual remote memory access protocol. After opening a
+communication line with a payload, the computer requests the data at
+certain virtual memory addresses (defined in further sections) to obtain
+payload data and metadata relating to it. There are no requirement that
+those addresses map to physical addresses on each instrument's
+controller -- just that the instrument return the expected pieces of
+data or metadata.
+
+\section{Virtual Registers}\label{virtual-registers}
+
+Payloads that implement AHAP should provided specific information in
+certain virtual registers, mapped at the following addresses:
+
+\begin{longtable}[]{@{}lll@{}}
+Address & Name &\tabularnewline
+\midrule
+\endhead
+\texttt{\$00{[}1{]}} & Tx Flag & \texttt{0x01} when the bus controller
+is addressing\tabularnewline
+\texttt{\$01{[}2{]}} & Data Length & Number of bytes of available
+data\tabularnewline
+\texttt{\$10{[}-{]}} & Data & Start of data made available to the
+bus\tabularnewline
+\end{longtable}
+
+\section{Communications format}\label{communications-format}
+
+AHABus uses the I2C protocol to denote whether an address is being read,
+or written to. Communication is always initiated by the Bus Controller.
+
+To accommodate low-budget, low-power devices which may present small I2C
+transmission buffers, the AHABus protocol does not allow transmission of
+more than 32 bytes at once. If a payload must send more than 32 bytes of
+data, the controller will read its data sequentially, in chunks of 32
+bytes starting at address \texttt{\$10}.
+
+\section{Synopsis}\label{synopsis}
+
+In typical operations, a bus controller/payload interaction follows
+these steps:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  The bus controller addresses the payload's address and writes
+  \texttt{0xff} in the Tx Flag register. At this point, the payload must
+  not change the contents of any of its public registers.
+\item
+  The bus controller requests the contents of the of Data Length
+  register by writing its address on the bus, then reading two bytes. If
+  no data is available (Data Length = \texttt{0x0000}), the controller
+  goes to step 4.
+\item
+  The bus controller requests the data by writing the address of the
+  data area, and reads \texttt{Data\ Length} bytes.
+\item
+  The bus controller writes \texttt{0x0} in the payload's Tx Flag
+  register. The connection is considered close after this point.
+\end{enumerate}
+
+In the event where any of the connection's communication would fail, the
+bus controller skips any further steps and attempts to close the
+connection.
+
+\section{Related Documents}\label{related-documents}
+
+\begin{itemize}
+\item
+  \href{https://github.com/AHABus/src/software/packet-radio.md}{AHABus
+  Packet Radio Protocol Specification}
+\end{itemize}
+
 \end{appendices}

### 2017-04-23 chapters/03-methodology.tex
Add specs in appendices, mention of Arduino Bus Library

@@
 
-Encapsulating the transmitted data adds overhead. If \(n\) is the size of a
-chunk in bytes, and each frame adds \(m\) bytes of metadata, then only a
-percentage \(\frac{n}{n+m}\) of the stream contains actual data. For reasons
-explained further, the \acrshort{ahabus} frames were chosen to be 256-byte
-long. As shown previously, each frame also includes.
+% Encapsulating the transmitted data adds overhead. If \(n\) is the size of a
+% chunk in bytes, and each frame adds \(m\) bytes of metadata, then only a
+% percentage \(\frac{n}{n+m}\) of the stream contains actual data. For reasons
+% explained further, the \acrshort{ahabus} frames were chosen to be 256-byte
+% long. As shown previously, each frame also includes.
 
@@
 more than 32 bytes of data available, the controller will send multiple read
-requests with increasing offsets from the base data address.
-
-Once the data has been successfully collected, the controller closes the
-connection by writing \texttt{0x00} in the connection flag.
+requests with increasing offsets from the base data address. Once the data has 
+been successfully collected, the controller closes the connection by writing
+\texttt{0x00} in the connection flag.
+
+To make development of payloads more accessible, an simple library was written
+for Arduino-compatible micro-controllers to abstract the bus protocol's
+implementation details behind simple function calls. The library makes use 
+of the Arduino library's implementation of \acrshort{i2c}, \texttt{Wire}.
 

### 2017-04-24 chapters/03-methodology.tex
Add reference to NMEA code library

@@
 \begin{figure}[H]
-\includegraphics[width=0.40\textwidth]{dl-fldigi.png}
+\includegraphics[width=0.80\textwidth]{dl-fldigi.png}
 \centering
@@
 \begin{figure}[H]
-\includegraphics[width=0.40\textwidth]{test-rig.jpg}
+\includegraphics[width=0.80\textwidth]{test-rig.jpg}
 \centering
@@
 simulate the position updates of a receiver attached to a \acrlong{hab}
-throughout ascent and descent.
+throughout ascent and descent, using a UKHAS library to emulate NMEA output
+(\cite{UKHAS2010}).
 

### 2017-04-24 chapters/05-discussion.tex
Fix typos in 5.1

@@
 through a low-power radio channel. In this chapter, the \acrshort{ahabus}
-platform prototype is evaluated against these aims, firs through general 
+platform prototype is evaluated against these aims, first through general 
 qualitative observations and analysis of the prototype's behaviour, then by
-analysing the results of tests discussed in chapter \ref{ch:results}.
+analysing the results of the tests discussed in chapter \ref{ch:results}.
 

### 2017-04-24 appendices.tex
Fix encoding format in appendix C

@@
     AHABus platform's latitude in decimal format at the time the packet
-    was encoded, in (8.24) fixed point format,
+    was encoded as a 32-bit signed integer, multiplied by 10000,
   \item
     AHABus platform's longitude in decimal format at the time the packet
-    was encoded, in (8.24) fixed point format,
+    was encoded as a 32-bit signed integer, multiplied by 10000,
   \item

### 2017-04-24 appendices.tex
Fix wrong datatype in Apx.C

@@
     04: u16         length
-    06: f32         latitude
-    08: f32         longitude
+    06: i32         latitude
+    08: i32         longitude
     0a: u16         altitude

### 2017-04-25 paper.tex
Add specs in appendices, mention of Arduino Bus Library
Add reference to NMEA code library
Proof read abstract done

--- 2017-04-17 paper.tex
@@
 \usepackage{hyperref}
+\usepackage{booktabs}
 
@@
 \usepackage[justification=centering,labelfont={small,bf},font={small,bf}]{caption}
-\usepackage[scaled]{beramono} % BitStream Vera font (code)
-\usepackage[T1]{fontenc}
-\usepackage[a4paper, margin=3cm]{geometry}
+\usepackage[titletoc,toc,page]{appendix}
 \usepackage[nottoc]{tocbibind}
 \usepackage{array}
+\usepackage{etoolbox}
+
+% reinstate the correct level for list of tables and figures
+\appto\listoffigures{\addtocontents{lof}{\protect\setcounter{tocdepth}{1}}}
+\appto\listoftables{\addtocontents{lot}{\protect\setcounter{tocdepth}{1}}}
 
@@
 \begin{abstract}
+
+The aim of this project was to investigate the design options for a modular payload bus for \acrfull{hab} missions, and to evaluate one such design by implementing in in hardware and software.
+
+High-Altitude Ballons (HAB) provide affordable access to near-space conditions, even for low-budget entities (universities, schools and amateurs).
+While current low-budget missions rely on commercial off-the-shelf computers and sensors, their architecture and software are often custom-made and targeted for a single use.
+This prevents easy reuse of hardware and software components, and raises the barrier to entry by requiring non-trivial networking and computing skills.
+
+This paper presents the development of the \acrfull{ahabus}, solution designed by applying the modular bus model used by most modern satellites to \acrlong{hab} architectures.
+An \acrfull{obc} running a common flight software (\acrshort{fcore}) provides facilities for position tracking and reliable radio telemetry to a ground station, and polls data from payloads, custom for each flight.
+
+The resulting protocols and platform prototype were successful and passed both day-in-the-life testing and radio channel resilience tests.
+Further testing of the \acrshort{ahabus} platorm and minor modifications to the Payload Bus and Radio Communications protocols will be required in order to qualify the platform for in-flight use.
+Once these steps are taken, \acrshort{ahabus} could render scientific \acrlong{hab} missions more accessible and affordable by allowing students and researchers to focus on payload design and development.
+
 \end{abstract}
@@
 \tableofcontents
-\setlength{\parskip}{1em}
 
@@
 \printglossary[type=\acronymtype]
+
+\setlength{\parskip}{1em}
 

### 2017-04-25 paper.tex
Proof read acknowledgements

@@
 
-I would like to thank my supervisor, Dr. Ian Ferguson, for the support and
-insight he gave throughout this research project, and as Dr. Natalie Coull for
-her advice during the proposal phase of the project.
+I would like to thank my supervisor, Dr.
+Ian Ferguson, for the support and insight he gave throughout this research project, as well as Dr.
+Natalie Coull for her advice during the proposal phase, Caitlin Smith and Chloë Smyth for their support over the duration of the project.
 

### 2017-04-25 chapters/01-introduction.tex
Proof read introduction

@@
 
-For applications that require low atmospheric density or medium altitudes, but do not depend on the payload being in a \acrfull{leo}, \acrfull{hab} present an affordable alternative.
-Such balloons, inflated with low-density gases (usually helium or hydrogen), can reach the stratosphere (altitudes of 30 to 45km).
+For applications that require low atmospheric density or medium altitudes, but do not depend on the payload being in a \acrfull{leo}, High-Altitude Balloons (\acrshort{hab}) present an affordable alternative.
+Such balloons, inflated with low-density gases (usually helium or hydrogen), can reach the stratosphere (altitudes of \SI{30}{\kilo\metre} to \SI{45}{\kilo\metre}).
 The payload is encased in a thermally insulated container attached under the balloon itself.
@@
 The main component of the payload is the \acrfull{obc}, which controls the different instruments through the different phases of flight (ascent, observation, free-fall and parachute descent).
-Because high-altitude winds can reach speeds upwards of 150km/h and carry the payload over long distances, most balloons carry a \acrfull{gnss} receiver that the OBC can use to keep track of the balloon's position and altitude.
-To avoid loosing flight data if the payload is lost or lands in an unaccessible area, the OBC also controls a radio transmission system used to forward payload data and metadata (\acrshort{gnss} coordinates, altitude, systems health) to a ground station as it is acquired.
+Because high-altitude winds can reach speeds upwards of \SI{150}{\kilo\metre\per\hour} and carry the payload over long distances, most balloons carry a \acrfull{gnss} receiver that the OBC can use to keep track of the balloon's position and altitude.
+To avoid loosing flight data if the payload is lost or lands in an unaccessible area, the OBC also controls a radio transmission system used to forward payload and ancillary data (\acrshort{gnss} co-ordinates, altitude, systems health) to a ground station as it is acquired.
 
@@
 The satellite industry solves this problem by making use of buses for most missions.
-Instead of designing a custom, monolithic platform for each new satellite, manufacturers sell buses that provide utility (power, data bus, attitude control and space-to-ground communications) to which the customer or science team connect mission-specific payloads.
+Instead of designing a custom, monolithic platform for each new satellite, manufacturers sell buses that provide common utilities (power, data bus, attitude control and platform-to-ground communications) to which the customer or science team connect mission-specific payloads.
 
@@
 
-In the following chapters, the current state of \acrlong{hab} and nano-satellite mission architectures is discussed.
-The proposed platform is then presented, followed by the results of the evaluation, and finally a critical discussion of those results and the overall performance of the platform.
+The current state of \acrlong{hab} and nano-satellite mission architectures is first discussed in chapter \ref{ch:literature-review}.
+The proposed platform, \acrshort{ahabus}, and the development process that led to its design are then presented in chapter \ref{ch:methodology}.
+Finally, chapters \ref{ch:results} and \ref{ch:discussion} present the results of the prototype's testing and show that the project was overall successful, even though more testing will be required to qualify the \acrshort{ahabus} platform for flight.

### 2017-04-25 chapters/02-lit-review.tex
Proof read lit review

@@
 
-While small-budget missions are built around commercial off-the-shelf small computers, or micro-controller-based boards for their \acrlong{obc} and instruments, the resulting designs are often non-modular, combine different communications standard in order to communicate with an heterogenous set of instruments, and make use of customised, text-based telemetry formats in the United Kingdom.
+While small-budget missions are built around commercial off-the-shelf small computers or micro-controller boards for their \acrlong{obc} and instruments, the resulting designs are often non-modular, combine different communications standard in order to communicate with an heterogenous set of instruments, and make use of customised, text-based telemetry formats in the United Kingdom.
 
@@
 
-On other projects like Emmanuel Bonbasaro's (\cite{Bombasaro2015}) Titan-1 mission, a micro-controller boards like an ``Arduino Mega 2560 is [...] used as flight computer'', alongside multiple commercial sensors connected to the \acrshort{obc} through multiple different data buses (in this case \acrfull{i2c} and \acrfull{spi}), because of the disparate communication standards in use in affordable commercial sensors.
+On other projects like Emmanuel Bonbasaro's (\cite{Bombasaro2015}) Titan-1 mission, a micro-controller board like an ``Arduino Mega 2560 is [...] used as flight computer'', alongside multiple commercial sensors connected to the \acrshort{obc} through multiple different data buses (in this case \acrfull{i2c} and \acrfull{spi}), because of the disparate communication standards in use in affordable commercial sensors.
 The flight software discussed by the author is a custom-made solution developed specifically for Titan 1's architecture, which would render its re-use in a different mission complicated.
@@
 
-When space conditions are required, or when the targeted scientific observations require longer or more global exposure times, nano-satellites have become popular over the last decade ( \cite{McDowell2017a}, \cite{McDowell2017}) as a relatively affordable – though much more costly than \acrshort{hab} missions – access to space.
+When space conditions are required, or when the targeted scientific observations require longer or more global exposure times than are achievable using High-Altitude Balloons, nano-satellites have become popular over the last decade ( \cite{McDowell2017a}, \cite{McDowell2017}) as a relatively affordable – though much more costly than \acrshort{hab} missions – access to space.
 The most popular standard is CubeSat (\cite{Heidt2000}), based on \SI{10}{\centi\metre} cubes small and light enough to be launched as secondary payloads, and thus only require a small portion of the launch costs to be covered by nano-satellites teams.
@@
 Thomas Clausen (\cite{Clausen2001}) describes in his 2001 paper the design of an internal bus for CubeSats based on the \acrshort{i2c} protocol as well.
-In this case too, \acrshort{i2c} was chosen for its low power consumption and I/O line usage, but also for its speed and reliance to noise, and because the addressing of payloads on the bus can be done entirely in software.
+In this case too, \acrshort{i2c} was chosen for its low power consumption and I/O line usage, but also for its speed and resilience to noise, and because the addressing of payloads on the bus can be done entirely in software.
 T.
